---
title: "Data Quality and Deduplication"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Quality and Deduplication}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dpi = 150
)
library(biblioIntegrator)
```

# Why This Vignette Exists {#sec-intro}

Data quality is the foundation of valid bibliometric analysis. No amount of sophisticated statistical modeling, network visualization, or comparative inference can compensate for a corpus contaminated by duplicate records, missing metadata, or inconsistent identifiers. This vignette provides a comprehensive guide to diagnosing, auditing, and cleaning bibliographic data using the `biblioIntegrator` package.

## The Problem of Data Quality

Bibliographic databases—including Web of Science, Scopus, PubMed, and OpenAlex—each have their own data structures, naming conventions, and quality standards. When researchers export data from multiple sources or over extended time periods, several problems emerge:

1. **Duplicate records**: The same publication may appear multiple times due to database updates, corrections, or cross-database indexing
2. **Missing metadata**: Critical fields such as titles, publication years, DOIs, or author names may be incomplete
3. **Inconsistent identifiers**: DOIs may be formatted differently (with or without URL prefixes), author names may vary across records
4. **Orphan records**: Author-keyword-work linkages may reference non-existent records after filtering

These issues directly impact every downstream analysis:

- **Descriptive statistics**: Duplicate records inflate publication counts, distort annual trends, and misrepresent author productivity
- **Citation metrics**: Duplicate entries split citation counts, leading to underestimation of true impact
- **Network analysis**: Duplicate nodes create artificial connections, fragment communities, and distort centrality measures
- **Comparative inference**: Unbalanced duplicates across groups introduce systematic bias into hypothesis tests
- **Temporal analysis**: Duplicates with different years create phantom trends

## When to Check Quality

Quality checks should precede **any** analytical step. The recommended workflow is:

```
Import → Harmonize → **Quality Check** → Deduplicate → Analyze
```

This vignette assumes you have already imported and harmonized your data using the functions described in `vignette("v01-import-harmonize")`. If you are working with a raw data frame, convert it first:

```{r eval=FALSE}
library(biblioIntegrator)
x <- as_biblio_project(my_data, source = "my_export")
```

## What You Will Not Find Here

This vignette does not cover:

- Data import from specific databases (see `vignette("v01-import-harmonize")`)
- Network construction or community detection (see `vignette("v05-networks")`)
- Comparative statistical inference (see `vignette("v04-comparative-inference")`)
- Scalable backends for large corpora (see `vignette("v08-scalable-backends")`)

These topics build on the clean, deduplicated corpus you will prepare here.


# Learning Objectives {#sec-objectives}

After working through this vignette, you will be able to:

1. **Diagnose corpus health** using `biblio_health()` and interpret each diagnostic check
2. **Identify missing metadata** including titles, years, DOIs, and author information
3. **Detect duplicate records** using DOI matching, title+year combinations, and fuzzy string matching
4. **Understand deduplication strategies** and their tradeoffs between precision and recall
5. **Apply deduplication** systematically using `deduplicate_biblio()` with different methods
6. **Audit data integrity** with `audit_biblio()` to verify provenance tracking
7. **Identify orphan records** and referential integrity issues in harmonized data
8. **Evaluate deduplication quality** by comparing results across methods
9. **Document data transformations** through provenance logs for reproducibility
10. **Avoid common pitfalls** in bibliometric data cleaning workflows

## Prerequisites

This vignette assumes basic familiarity with:

- R data frames and basic data manipulation
- The `biblio_project` object class (see `vignette("v00-overview")`)
- The concept of bibliometric indicators (see `vignette("v03-descriptive-impact")`)

No external packages beyond `biblioIntegrator` are required for the core examples. Optional fuzzy matching demonstrations require the `stringdist` package, which is not a dependency of `biblioIntegrator` but can be installed separately.

```{r prerequisites}
# Core package
library(biblioIntegrator)

# Optional: for fuzzy matching examples
has_stringdist <- requireNamespace("stringdist", quietly = TRUE)
```


# The Health Diagnostic {#sec-health}

The `biblio_health()` function provides a systematic overview of data quality issues in a `biblio_project`. It performs six standardized checks and returns a data frame summarizing the number of records affected by each issue.

## Running biblio_health()

Let us start with the example corpus included in the package:

```{r health-basic}
# Load example data
raw <- example_biblio()

# Create a biblio_project
x <- as_biblio_project(raw, source = "example corpus")

# Run health diagnostics
health_report <- biblio_health(x)
print(health_report)
```

The output is a data frame with two columns:

- **check**: The name of the quality check
- **n**: The number of records flagged by that check

## Understanding the Output Structure

The health report includes six standardized checks. Let us examine what each one detects:

```{r health-structure}
# Display the health report in a readable format
cat("Quality Health Report\n")
cat("====================\n\n")
for (i in seq_len(nrow(health_report))) {
  cat(sprintf("%-25s %d records\n", 
              health_report$check[i], 
              health_report$n[i]))
}
```

The six checks are:

| Check | Description |
|-------|-------------|
| `missing_title` | Records with empty or whitespace-only titles |
| `missing_year` | Records with `NA` publication year |
| `missing_doi` | Records with empty DOI field |
| `duplicate_doi` | Records sharing the same DOI (excluding empty DOIs) |
| `duplicate_title_year` | Records with identical title+year combinations |
| `negative_citations` | Records with negative citation counts |

## Interpreting Each Check

### Missing Titles

Missing titles are problematic because titles are used as identifiers in deduplication and are essential for human verification of results:

```{r health-missing-title}
# Check for missing titles
missing_title_count <- health_report$n[health_report$check == "missing_title"]
cat("Records with missing titles:", missing_title_count, "\n")

# If non-zero, identify the affected records
if (missing_title_count > 0) {
  problem_records <- x$works[!nzchar(trimws(x$works$title)), ]
  print(problem_records[, c("work_id", "year", "doi")])
}
```

**Impact**: Records without titles cannot be reliably deduplicated by title matching and should be flagged for manual review.

### Missing Years

Missing publication years affect temporal analysis, citation velocity calculations, and period-based comparisons:

```{r health-missing-year}
# Check for missing years
missing_year_count <- health_report$n[health_report$check == "missing_year"]
cat("Records with missing years:", missing_year_count, "\n")

# If non-zero, show the problematic records
if (missing_year_count > 0) {
  problem_records <- x$works[is.na(x$works$year), ]
  print(problem_records[, c("work_id", "title", "doi")])
}
```

**Impact**: Functions like `describe_biblio()`, `trend_topics()`, and `temporal_growth()` will exclude or misclassify records with missing years.

### Missing DOIs

DOIs (Digital Object Identifiers) are the most reliable unique identifier for scholarly publications:

```{r health-missing-doi}
# Check for missing DOIs
missing_doi_count <- health_report$n[health_report$check == "missing_doi"]
cat("Records with missing DOIs:", missing_doi_count, "\n")

# Calculate percentage
cat(sprintf("%.1f%% of records lack DOIs\n", 
            100 * missing_doi_count / nrow(x$works)))
```

**Impact**: Without DOIs, deduplication must rely on less reliable title+year matching, increasing the risk of false positives (merging distinct records) or false negatives (missing actual duplicates).

### Duplicate DOIs

Duplicate DOIs are a strong indicator of true duplicates, since DOIs are designed to be unique identifiers:

```{r health-dup-doi}
# Check for duplicate DOIs
dup_doi_count <- health_report$n[health_report$check == "duplicate_doi"]
cat("Records with duplicate DOIs:", dup_doi_count, "\n")

# If non-zero, identify the duplicated DOIs
if (dup_doi_count > 0) {
  doi_table <- table(x$works$doi[nzchar(x$works$doi)])
  repeated_dois <- names(doi_table[doi_table > 1])
  cat("\nDuplicated DOIs:\n")
  for (doi in head(repeated_dois, 5)) {
    cat(sprintf("  %s: %d copies\n", doi, sum(x$works$doi == doi)))
  }
}
```

**Deduplication strategy**: When DOIs match, we can be highly confident these are true duplicates. The DOI-based matching has near-zero false positive rate.

### Duplicate Title+Year

Title+year matching catches duplicates that may have different or missing DOIs:

```{r health-dup-title-year}
# Check for duplicate title+year combinations
dup_ty_count <- health_report$n[health_report$check == "duplicate_title_year"]
cat("Records with duplicate title+year:", dup_ty_count, "\n")

# Show examples if any exist
if (dup_ty_count > 0) {
  title_year <- paste(tolower(x$works$title), x$works$year)
  ty_table <- table(title_year)
  repeated_ty <- names(ty_table[ty_table > 1])
  cat("\nDuplicated title+year combinations:\n")
  for (ty in head(repeated_ty, 3)) {
    cat(sprintf("  '%s'\n", ty))
  }
}
```

**Precision concern**: Some legitimate publications share the same title and year (e.g., preprints and final versions, or corrected/retracted articles). Manual review is recommended for title+year matches without DOI confirmation.

### Negative Citations

Negative citation counts are data errors that should be corrected or excluded:

```{r health-negative-citations}
# Check for negative citations
neg_cite_count <- health_report$n[health_report$check == "negative_citations"]
cat("Records with negative citations:", neg_cite_count, "\n")
```

**Impact**: Negative citations will distort `biblio_metrics()`, `normalized_citations()`, and `citation_velocity()` calculations.

## Filtering the Health Report

For large corpora, you may want to focus on actual problems rather than checks that passed:

```{r health-filter}
# Show only checks that detected issues
problems <- subset(health_report, n > 0)
cat("Active quality issues:\n")
print(problems)

# If no issues found
if (nrow(problems) == 0) {
  cat("No quality issues detected. Corpus is clean.\n")
}
```

## Visualizing Health Status

For a quick overview, you can visualize the health report:

```{r health-plot, eval=FALSE}
# Basic barplot of issues (only if issues exist)
problems <- subset(health_report, n > 0)
if (nrow(problems) > 0) {
  barplot(problems$n, 
          names.arg = problems$check,
          main = "Data Quality Issues",
          xlab = "Check Type",
          ylab = "Number of Records",
          col = ifelse(problems$n > 0, "coral", "lightgreen"),
          las = 2)
}
```


# Introducing Quality Issues {#sec-introduce-issues}

To demonstrate the deduplication workflow, we will systematically introduce quality issues into the example corpus. This approach allows us to verify that our cleaning procedures work correctly.

## Creating a Contaminated Corpus

```{r introduce-issues}
# Start with clean example data
clean_data <- example_biblio()

# Create duplicates by repeating records
# 1. Exact DOI duplicate
dup1 <- clean_data[1, ]  # First record

# 2. Same title+year but different DOI (simulates cross-database duplicate)
dup2 <- clean_data[3, ]
dup2$doi <- "10.9999/alternate.doi.3"

# 3. Similar but not identical title (will test fuzzy matching)
dup3 <- clean_data[5, ]
dup3$title <- "Cover Crop Effects on Soil Aggregation"  # Slightly different
dup3$doi <- ""  # No DOI

# Combine into contaminated corpus
contaminated_data <- rbind(clean_data, dup1, dup2, dup3)

# Also introduce some missing data
contaminated_data$title[2] <- ""  # Missing title
contaminated_data$year[4] <- NA   # Missing year
contaminated_data$doi[6] <- ""    # Missing DOI
contaminated_data$citations[8] <- -5  # Negative citation

cat("Original records:", nrow(clean_data), "\n")
cat("Contaminated records:", nrow(contaminated_data), "\n")
cat("Added records:", nrow(contaminated_data) - nrow(clean_data), "\n")
```

## Converting to biblio_project

```{r convert-contaminated}
# Convert contaminated data to biblio_project
x_dirty <- as_biblio_project(contaminated_data, source = "contaminated example")

# Run health check on contaminated data
health_dirty <- biblio_health(x_dirty)
cat("\nHealth report for contaminated corpus:\n")
print(health_dirty)
```

You should see elevated counts for several checks, reflecting the issues we introduced.

## Examining the Contaminated Data

```{r examine-contaminated}
# Show the problematic records
cat("Records with issues:\n\n")

# Missing title
if (any(!nzchar(trimws(x_dirty$works$title)))) {
  cat("Missing titles:\n")
  print(x_dirty$works[!nzchar(trimws(x_dirty$works$title)), 
                      c("work_id", "title", "year", "doi")])
}

# Missing year
if (any(is.na(x_dirty$works$year))) {
  cat("\nMissing years:\n")
  print(x_dirty$works[is.na(x_dirty$works$year), 
                      c("work_id", "title", "year", "doi")])
}

# Negative citations
if (any(x_dirty$works$cited_by_count < 0, na.rm = TRUE)) {
  cat("\nNegative citations:\n")
  print(x_dirty$works[x_dirty$works$cited_by_count < 0, 
                      c("work_id", "title", "year", "cited_by_count")])
}
```


# Deduplication Strategies {#sec-strategies}

Deduplication is the process of identifying and removing duplicate records from a dataset. In bibliometric analysis, this is critical because duplicates inflate counts and distort analyses. The `biblioIntegrator` package implements a hierarchical deduplication strategy that balances precision (correctly identifying true duplicates) and recall (finding all actual duplicates).

## The Matching Hierarchy

The `deduplicate_biblio()` function implements a two-stage matching hierarchy:

```
Stage 1: DOI Matching (High Precision)
    ↓ No DOI match
Stage 2: Title+Year Matching (Moderate Precision)
    ↓ Results combined
Final: Deduplicated Corpus
```

### Stage 1: DOI Matching

DOI matching is the most reliable deduplication method because DOIs are designed to be globally unique identifiers for scholarly publications.

**Advantages:**
- Near-zero false positive rate
- Fast computation
- Handles abbreviations and URL prefixes automatically

**Limitations:**
- Only works when DOIs are present and correctly formatted
- Some older publications lack DOIs
- Preprints and final versions may have different DOIs

```{r strategy-doi}
# Demonstrate DOI normalization
example_dois <- c(
  "10.1000/agri.1",
  "https://doi.org/10.1000/agri.2",
  "doi:10.1000/ag.3",
  "10.1000/agri.4",
  NA,
  ""
)

# Show how biblioIntegrator normalizes DOIs
cat("DOI normalization examples:\n")
cat("Raw → Normalized\n")
cat("--- → ----------\n")
for (doi in example_dois) {
  # Simple DOI normalization (lowercase, strip prefix)
  normalized <- tolower(trimws(as.character(doi)))
  normalized <- sub("^https?://(dx\\.)?doi\\.org/", "", normalized)
  normalized <- sub("^doi:\\s*", "", normalized)
  cat(sprintf("%s → %s\n", 
              ifelse(is.na(doi), "NA", paste0("'", doi, "'")),
              paste0("'", normalized, "'")))
}
```

### Stage 2: Title+Year Matching

When DOIs are missing or do not match, the algorithm falls back to title+year matching.

**Process:**
1. Normalize titles: convert to lowercase, remove non-alphanumeric characters
2. Combine with year: create key as `normalized_title:year`
3. Match on these combined keys

**Advantages:**
- Works for publications without DOIs
- Catches duplicates across databases with different DOI assignments

**Limitations:**
- Higher false positive rate than DOI matching
- Sensitive to title variations (subtitles, abbreviations, typos)
- Legitimate duplicate titles exist (e.g., corrected papers, errata)

```{r strategy-title-year}
# Demonstrate title normalization
example_titles <- c(
  "Silicon and salinity tolerance in rice",
  "Silicon and Salinity Tolerance in Rice",
  "Silicon  &  salinity  tolerance  in  rice!",
  "Silicon and salinity tolerance in rice: A review"
)

# Show normalization process
cat("Title normalization examples:\n")
for (title in example_titles) {
  # Replicate normalization from deduplicate_biblio()
  normalized <- gsub("[^[:alnum:]]", "", tolower(title))
  cat(sprintf("'%s'\n  → '%s'\n", title, normalized))
}
```

## Precision vs. Recall Tradeoff

The choice of deduplication strategy involves balancing two competing objectives:

| Strategy | Precision | Recall | Best For |
|----------|-----------|--------|----------|
| DOI only | Very High | Low | Clean data with good DOI coverage |
| Title+Year only | Moderate | High | Data with poor DOI coverage |
| DOI then Title+Year | High | High | General use (default) |
| Fuzzy matching | Lower | Very High | Aggressive cleaning, manual review |

**Precision** = (True Duplicates Found) / (All Pairs Flagged as Duplicates)

**Recall** = (True Duplicates Found) / (All Actual Duplicates)

The default `doi_title_year` method prioritizes precision while maintaining reasonable recall by using DOI matching first, then falling back to title+year only for records without DOI matches.

## Alternative: Fuzzy String Matching

For more aggressive duplicate detection, fuzzy string matching can identify records with similar (but not identical) titles. This is **not** built into `deduplicate_biblio()` because:

1. It requires the optional `stringdist` package
2. It has a higher false positive rate
3. It requires careful threshold tuning
4. Results typically need manual verification

However, we demonstrate how to implement it for advanced users:

```{r strategy-fuzzy, eval=has_stringdist}
if (has_stringdist) {
  # Example fuzzy matching (not part of deduplicate_biblio)
  library(stringdist)
  
  # Create a small example
  titles <- c(
    "Cover crops and soil aggregation",
    "Cover Crop Effects on Soil Aggregation",
    "Machine learning for crop yield",
    "Remote sensing of soybean nitrogen"
  )
  
  # Calculate pairwise string distances
  dist_matrix <- stringdistmatrix(titles, titles, method = "jw")
  
  # Find potential duplicates (distance < 0.15)
  threshold <- 0.15
  cat("Potential fuzzy duplicates (Jaro-Winkler < ", threshold, "):\n")
  for (i in 1:(length(titles) - 1)) {
    for (j in (i + 1):length(titles)) {
      if (dist_matrix[i, j] < threshold) {
        cat(sprintf("  [%.3f] '%s' ≈ '%s'\n", 
                    dist_matrix[i, j], titles[i], titles[j]))
      }
    }
  }
}
```

**Recommendation**: Use the default `doi_title_year` method for most analyses. Only consider fuzzy matching if you have evidence of widespread title variations and are prepared to manually verify results.


# Step-by-Step Deduplication Workflow {#sec-workflow}

Now that we understand the strategies, let us apply them systematically to clean our contaminated corpus.

## Step 1: Initial Health Assessment

Always start with a health check to understand the scope of quality issues:

```{r workflow-step1}
# Load contaminated data (from Section 4)
raw <- example_biblio()

# Create duplicates
dup1 <- raw[1, ]  # Exact DOI duplicate
dup2 <- raw[3, ]  # Same title+year, different DOI
dup2$doi <- "10.9999/alternate.3"
dup3 <- raw[5, ]  # Similar title, no DOI
dup3$title <- "Cover Crop Effects on Soil Aggregation"
dup3$doi <- ""

contaminated <- rbind(raw, dup1, dup2, dup3)

# Introduce missing data
contaminated$title[2] <- ""
contaminated$year[4] <- NA
contaminated$doi[6] <- ""
contaminated$citations[8] <- -5

# Create biblio_project
x_dirty <- as_biblio_project(contaminated, source = "contaminated")

# Step 1: Assess quality
cat("=== STEP 1: Initial Health Assessment ===\n\n")
health_before <- biblio_health(x_dirty)
print(health_before)

cat("\nSummary:\n")
cat(sprintf("  Total records: %d\n", nrow(x_dirty$works)))
cat(sprintf("  Records with issues: %d\n", 
            sum(health_before$n[health_before$check != "negative_citations"])))
```

## Step 2: Apply Deduplication

```{r workflow-step2}
# Step 2: Deduplicate using default method
cat("\n=== STEP 2: Deduplication ===\n\n")

x_clean <- deduplicate_biblio(x_dirty)

cat("Deduplication results:\n")
cat(sprintf("  Records before: %d\n", nrow(x_dirty$works)))
cat(sprintf("  Records after:  %d\n", nrow(x_clean$works)))
cat(sprintf("  Records removed: %d\n", nrow(x_dirty$works) - nrow(x_clean$works)))
```

## Step 3: Examine the Deduplication Log

The deduplication log records which records were removed and why:

```{r workflow-step3}
# Step 3: Review deduplication log
cat("\n=== STEP 3: Deduplication Log ===\n\n")

dedup_log <- attr(x_clean, "dedup_log")
cat("Removed records:\n")
print(dedup_log)

cat("\nInterpretation:\n")
cat("- Records are removed when they match an earlier record by DOI or title+year\n")
cat("- The first occurrence is always kept\n")
cat("- The log preserves removed records for verification\n")
```

## Step 4: Verify Post-Deduplication Health

```{r workflow-step4}
# Step 4: Check health after deduplication
cat("\n=== STEP 4: Post-Deduplication Health ===\n\n")

health_after <- biblio_health(x_clean)
print(health_after)

cat("\nComparison:\n")
cat(sprintf("  Duplicate DOIs removed: %d\n", 
            health_before$n[health_before$check == "duplicate_doi"] - 
            health_after$n[health_after$check == "duplicate_doi"]))
cat(sprintf("  Duplicate title+year removed: %d\n",
            health_before$n[health_before$check == "duplicate_title_year"] - 
            health_after$n[health_after$check == "duplicate_title_year"]))
```

## Step 5: Audit the Provenance

The provenance log records all transformations applied to the corpus:

```{r workflow-step5}
# Step 5: Audit provenance
cat("\n=== STEP 5: Provenance Audit ===\n\n")

audit_log <- audit_biblio(x_clean)
cat("Provenance log:\n")
print(audit_log)

cat("\nProvenance entries:\n")
cat(sprintf("  as_biblio_project: %d entries\n", 
            sum(audit_log$operation == "as_biblio_project")))
cat(sprintf("  deduplicate_biblio: %d entries\n",
            sum(audit_log$operation == "deduplicate_biblio")))
```

## Step 6: Compare Records Before and After

```{r workflow-step6}
# Step 6: Verify record counts
cat("\n=== STEP 6: Record Comparison ===\n\n")

cat("Works table:\n")
cat(sprintf("  Before: %d records\n", nrow(x_dirty$works)))
cat(sprintf("  After:  %d records\n", nrow(x_clean$works)))
cat(sprintf("  Change: %d records removed\n", 
            nrow(x_dirty$works) - nrow(x_clean$works)))

cat("\nAuthorships table:\n")
cat(sprintf("  Before: %d links\n", nrow(x_dirty$authorships)))
cat(sprintf("  After:  %d links\n", nrow(x_clean$authorships)))

cat("\nKeywords table:\n")
cat(sprintf("  Before: %d links\n", nrow(x_dirty$keywords)))
cat(sprintf("  After:  %d links\n", nrow(x_clean$keywords)))
```

## Step 7: Test Different Methods

Although the package currently implements only the `doi_title_year` method, understanding the matching logic helps interpret results:

```{r workflow-step7}
# Step 7: Demonstrate matching logic
cat("\n=== STEP 7: Matching Logic ===\n\n")

# Show how keys are constructed
works <- x_dirty$works
keys <- ifelse(
  nzchar(works$doi),
  paste0("doi:", works$doi),
  paste0("ty:", tolower(gsub("[^[:alnum:]]", "", works$title)), ":", works$year)
)

cat("Matching keys for first 6 records:\n")
for (i in 1:min(6, nrow(works))) {
  cat(sprintf("  [%d] %s\n", i, substr(keys[i], 1, 60)))
}

# Identify duplicates by key
dup_keys <- keys[duplicated(keys)]
if (length(dup_keys) > 0) {
  cat("\nDuplicate keys found:\n")
  for (k in unique(dup_keys)) {
    count <- sum(keys == k)
    cat(sprintf("  '%s' appears %d times\n", substr(k, 1, 50), count))
  }
}
```


# Handling Specific Quality Issues {#sec-specific-issues}

Beyond deduplication, several quality issues require attention. This section addresses each systematically.

## Missing Titles

Missing titles prevent reliable deduplication and make records difficult to verify:

```{r missing-titles}
# Identify records with missing titles
x <- as_biblio_project(example_biblio())
x$works$title[3] <- ""  # Simulate missing title

missing_title_idx <- which(!nzchar(trimws(x$works$title)))

if (length(missing_title_idx) > 0) {
  cat("Records with missing titles:\n")
  print(x$works[missing_title_idx, c("work_id", "year", "doi", "source")])
  
  cat("\nOptions:\n")
  cat("1. Look up titles using DOI (if available)\n")
  cat("2. Look up titles using metadata APIs (OpenAlex, CrossRef)\n")
  cat("3. Exclude records with missing titles\n")
  cat("4. Flag for manual review\n")
}
```

## Missing Years

Missing years affect temporal analyses:

```{r missing-years}
# Identify records with missing years
x$works$year[5] <- NA  # Simulate missing year

missing_year_idx <- which(is.na(x$works$year))

if (length(missing_year_idx) > 0) {
  cat("Records with missing years:\n")
  print(x$works[missing_year_idx, c("work_id", "title", "doi", "source")])
  
  cat("\nOptions:\n")
  cat("1. Look up year using DOI (if available)\n")
  cat("2. Infer year from citation patterns\n")
  cat("3. Assign to 'unknown' period and exclude from temporal analysis\n")
  cat("4. Manual correction\n")
}
```

## Missing DOIs

Records without DOIs require alternative identification strategies:

```{r missing-dois}
# Calculate DOI coverage
doi_coverage <- sum(nzchar(x$works$doi)) / nrow(x$works)
cat(sprintf("DOI coverage: %.1f%%\n", 100 * doi_coverage))

if (doi_coverage < 0.9) {
  cat("\nWarning: Low DOI coverage may affect deduplication quality.\n")
  cat("Consider:\n")
  cat("1. Re-exporting with DOI fields included\n")
  cat("2. Using title+year matching (default)\n")
  cat("3. Manual DOI lookup for high-impact records\n")
}
```

## Inconsistent DOIs

DOI format inconsistencies can prevent matching:

```{r inconsistent-dois}
# Demonstrate DOI normalization
example_dois <- c(
  "10.1000/AGRI.1",
  "10.1000/agri.1",
  "https://doi.org/10.1000/agri.1",
  "http://dx.doi.org/10.1000/agri.1",
  "doi:10.1000/agri.1",
  "DOI:10.1000/agri.1"
)

# Show normalization
cat("DOI normalization examples:\n")
for (doi in example_dois) {
  normalized <- tolower(trimws(as.character(doi)))
  normalized <- sub("^https?://(dx\\.)?doi\\.org/", "", normalized)
  normalized <- sub("^doi:\\s*", "", normalized)
  cat(sprintf("  %-45s → %s\n", doi, normalized))
}
```

The `biblioIntegrator` package automatically normalizes DOIs during import, but if you are merging data from multiple sources, verify that normalization was applied consistently.

## Negative Citation Counts

Negative citations are data errors that should be corrected:

```{r negative-citations}
# Identify negative citations
x$works$cited_by_count[2] <- -10  # Simulate error

negative_idx <- which(x$works$cited_by_count < 0)

if (length(negative_idx) > 0) {
  cat("Records with negative citations:\n")
  print(x$works[negative_idx, c("work_id", "title", "year", "cited_by_count")])
  
  # Options for handling
  cat("\nOptions:\n")
  cat("1. Set to 0 (conservative)\n")
  cat("2. Set to NA and exclude from citation analysis\n")
  cat("3. Look up correct value and correct\n")
  
  # Example correction
  x$works$cited_by_count[negative_idx] <- 0
  cat("\nCorrected by setting to 0.\n")
}
```


# The Audit Function {#sec-audit}

The `audit_biblio()` function provides access to the provenance log, which records all transformations applied to a `biblio_project`. This is essential for reproducibility and transparency in bibliometric analysis.

## Understanding Provenance

Every `biblio_project` maintains a provenance log that records:

1. **Timestamp**: When the operation was performed
2. **Operation**: The name of the function called
3. **Details**: Key parameters and outcomes

```{r audit-basic}
# Create a project and apply transformations
x <- as_biblio_project(example_biblio(), source = "audit demo")

# Apply deduplication
x <- deduplicate_biblio(x)

# Access provenance
audit_log <- audit_biblio(x)
cat("Provenance log:\n")
print(audit_log)
```

## Interpreting Provenance Entries

Each row in the provenance log represents one transformation:

```{r audit-interpret}
# Interpret each entry
cat("Provenance interpretation:\n\n")

for (i in seq_len(nrow(audit_log))) {
  entry <- audit_log[i, ]
  cat(sprintf("Entry %d [%s]\n", i, entry$timestamp))
  cat(sprintf("  Operation: %s\n", entry$operation))
  cat(sprintf("  Details: %s\n", entry$details))
  cat("\n")
}
```

## Checking Referential Integrity

After deduplication, verify that all linkages are consistent:

```{r audit-referential}
# Check referential integrity
cat("Referential integrity check:\n\n")

# All work_ids in authorships should exist in works
orphan_authorships <- !x$authorships$work_id %in% x$works$work_id
cat(sprintf("Orphan authorship links: %d\n", sum(orphan_authorships)))

# All work_ids in keywords should exist in works
orphan_keywords <- !x$keywords$work_id %in% x$works$work_id
cat(sprintf("Orphan keyword links: %d\n", sum(orphan_keywords)))

# All author_ids in authorships should exist in authors
orphan_authors <- !x$authorships$author_id %in% x$authors$author_id
cat(sprintf("Orphan author references: %d\n", sum(orphan_authors)))
```

**Note**: The `deduplicate_biblio()` function automatically removes associated authorship and keyword links when a work is removed. Orphan records should be zero after proper deduplication.

## Identifying Orphan Records

Orphan records are linkages that reference non-existent entities:

```{r audit-orphans}
# Demonstrate orphan detection
cat("\nOrphan detection:\n\n")

# Create a scenario with orphans for illustration
x_demo <- as_biblio_project(example_biblio())

# Manually create orphan (not recommended in practice)
orphan_link <- data.frame(
  work_id = "nonexistent_work_123",
  author_id = "A00000001",
  stringsAsFactors = FALSE
)
x_demo$authorships <- rbind(x_demo$authorships, orphan_link)

# Detect orphans
orphan_check <- !x_demo$authorships$work_id %in% x_demo$works$work_id
if (any(orphan_check)) {
  cat("Found orphan authorship links:\n")
  print(x_demo$authorships[orphan_check, ])
  
  # Clean up
  x_demo$authorships <- x_demo$authorships[!orphan_check, ]
  cat("\nOrphans removed.\n")
}
```

## Provenance for Reproducibility

The audit trail enables reproducible analysis:

```{r audit-reproducibility}
# Example: Complete audit trail for a cleaning workflow
x <- as_biblio_project(example_biblio(), source = "reproducibility demo")
x <- deduplicate_biblio(x)

# Export the audit trail
audit_df <- audit_biblio(x)

# Save for documentation (example)
temp_file <- tempfile(fileext = ".csv")
write.csv(audit_df, temp_file, row.names = FALSE)

cat("Audit trail saved to:", temp_file, "\n")
cat("\nAudit trail contents:\n")
print(audit_df)

# Verify all steps are recorded
cat("\nVerification:\n")
cat(sprintf("Total operations recorded: %d\n", nrow(audit_df)))
cat(sprintf("Operations: %s\n", paste(audit_df$operation, collapse = " → ")))
```


# Comparing Deduplication Results {#sec-comparing}

Understanding how different deduplication approaches affect results is essential for methodological transparency.

## Setting Up Comparisons

```{r comparison-setup}
# Create a dataset with known duplicates
raw <- example_biblio()

# Add exact duplicates
dup_exact <- raw[c(1, 3, 5), ]

# Add title+year duplicates (different DOI)
dup_ty <- raw[c(2, 4), ]
dup_ty$doi <- c("10.9999/alt.2", "10.9999/alt.4")

# Add similar titles (fuzzy matches)
dup_fuzzy <- raw[c(6, 8), ]
dup_fuzzy$title <- c(
  "Machine Learning Applications for Crop Yield Prediction",
  "Soil Microbiome Dynamics Under Crop Rotation Systems"
)

# Combine all
full_data <- rbind(raw, dup_exact, dup_ty, dup_fuzzy)

cat("Original records:", nrow(raw), "\n")
cat("Total records:", nrow(full_data), "\n")
cat("Expected duplicates:", nrow(full_data) - nrow(raw), "\n")
```

## Applying Default Deduplication

```{r comparison-default}
# Apply default deduplication
x_original <- as_biblio_project(full_data, source = "comparison")
x_dedup <- deduplicate_biblio(x_original)

# Analyze results
cat("\nDefault deduplication (doi_title_year):\n")
cat(sprintf("  Input records: %d\n", nrow(x_original$works)))
cat(sprintf("  Output records: %d\n", nrow(x_dedup$works)))
cat(sprintf("  Removed: %d (%.1f%%)\n", 
            nrow(x_original$works) - nrow(x_dedup$works),
            100 * (nrow(x_original$works) - nrow(x_dedup$works)) / nrow(x_original$works)))

# Show what was removed
dedup_log <- attr(x_dedup, "dedup_log")
cat("\nRemoved records:\n")
print(dedup_log)
```

## Analyzing Matching Behavior

```{r comparison-analysis}
# Understand why records were removed
cat("\nMatching analysis:\n\n")

# Reconstruct matching keys
works <- x_original$works
keys <- ifelse(
  nzchar(works$doi),
  paste0("doi:", works$doi),
  paste0("ty:", tolower(gsub("[^[:alnum:]]", "", works$title)), ":", works$year)
)

# Count duplicates by key
key_table <- table(keys)
dup_keys <- key_table[key_table > 1]

cat("Duplicate key groups:\n")
for (i in seq_along(dup_keys)) {
  key_name <- names(dup_keys)[i]
  count <- dup_keys[i]
  cat(sprintf("  %s: %d records\n", 
              substr(key_name, 1, 60), count))
}

# Categorize by matching strategy
doi_matches <- sum(grepl("^doi:", keys[duplicated(keys)]))
ty_matches <- sum(grepl("^ty:", keys[duplicated(keys)]))

cat(sprintf("\nDOI matches: %d\n", doi_matches))
cat(sprintf("Title+year matches: %d\n", ty_matches))
```

## Effect on Downstream Analysis

Different deduplication approaches can significantly impact results:

```{r comparison-impact}
# Compare corpus statistics before and after
cat("Impact comparison:\n\n")

# Original vs deduplicated
metrics_orig <- nrow(x_original$works)
metrics_dedup <- nrow(x_dedup$works)

cat("Record counts:\n")
cat(sprintf("  Original:            %d works\n", metrics_orig))
cat(sprintf("  After deduplication: %d works\n", metrics_dedup))
cat(sprintf("  Reduction:           %.1f%%\n", 
            100 * (metrics_orig - metrics_dedup) / metrics_orig))

# Author counts
authors_orig <- nrow(x_original$authors)
authors_dedup <- nrow(x_dedup$authors)
cat(sprintf("\nAuthors:\n"))
cat(sprintf("  Original:            %d unique authors\n", authors_orig))
cat(sprintf("  After deduplication: %d unique authors\n", authors_dedup))

# Citation totals
citations_orig <- sum(x_original$works$cited_by_count, na.rm = TRUE)
citations_dedup <- sum(x_dedup$works$cited_by_count, na.rm = TRUE)
cat(sprintf("\nTotal citations:\n"))
cat(sprintf("  Original:            %d\n", citations_orig))
cat(sprintf("  After deduplication: %d\n", citations_dedup))
```


# Common Mistakes {#sec-mistakes}

This section documents frequent errors in bibliometric data cleaning and how to avoid them.

## Mistake 1: Deduplicating Before Harmonization

**Problem**: Applying deduplication to raw, unharmonized data.

**Why it's wrong**: Different databases use different field names, formats, and conventions. Deduplication before harmonization will miss duplicates that look different in raw form.

**Solution**: Always harmonize first using `as_biblio_project()`.

```{r mistake-harmonize}
# WRONG: Deduplicating raw data
cat("WRONG approach:\n")
cat("  raw_data <- read.csv('scopus_export.csv')\n")
cat("  clean_data <- deduplicate_biblio(raw_data)  # Error!\n\n")

# CORRECT: Harmonize first, then deduplicate
cat("CORRECT approach:\n")
cat("  raw_data <- read.csv('scopus_export.csv')\n")
cat("  x <- as_biblio_project(raw_data, source = 'scopus')\n")
cat("  x_clean <- deduplicate_biblio(x)\n")

# Demonstrate
cat("\nPractical example:\n")
raw <- example_biblio()
x <- as_biblio_project(raw, source = "example")
x_clean <- deduplicate_biblio(x)
cat(sprintf("  Harmonized: %d records\n", nrow(x$works)))
cat(sprintf("  Deduplicated: %d records\n", nrow(x_clean$works)))
```

## Mistake 2: Using Only One Deduplication Method

**Problem**: Relying solely on DOI matching or only on title matching.

**Why it's wrong**: Each method has different strengths:

- DOI matching misses duplicates without DOIs
- Title matching misses duplicates with title variations
- Neither catches all duplicates alone

**Solution**: Use the hierarchical `doi_title_year` method (default).

```{r mistake-methods}
# Show why multiple methods are needed
cat("Method comparison:\n\n")

# Create test data with different duplicate types
test_data <- example_biblio()
dup_doi <- test_data[1, ]  # Will match by DOI
dup_title <- test_data[2, ]  # Will match by title+year
dup_title$doi <- ""  # No DOI, must use title+year

combined <- rbind(test_data, dup_doi, dup_title)
x_test <- as_biblio_project(combined, source = "method test")

# Show matching keys
works <- x_test$works
keys <- ifelse(
  nzchar(works$doi),
  paste0("doi:", works$doi),
  paste0("ty:", tolower(gsub("[^[:alnum:]]", "", works$title)), ":", works$year)
)

cat("Matching keys:\n")
for (i in 1:min(5, nrow(works))) {
  cat(sprintf("  Record %d: %s\n", i, substr(keys[i], 1, 50)))
}

cat("\nThe doi_title_year method handles both cases automatically.\n")
```

## Mistake 3: Ignoring False Positives in Fuzzy Matching

**Problem**: Applying aggressive fuzzy matching without manual verification.

**Why it's wrong**: Legitimate publications can have similar titles:

- Review articles on the same topic
- Sequel papers ("Part I", "Part II")
- Different papers with common keywords

**Solution**: If using fuzzy matching (outside the package), always verify a sample.

```{r mistake-fuzzy}
# Example of legitimate similar titles
similar_titles <- c(
  "Silicon and salinity tolerance in rice",
  "Silicon and salinity tolerance in rice: A review",
  "Silicon and salinity tolerance in rice: Response mechanisms"
)

cat("Legitimate similar titles:\n")
for (t in similar_titles) {
  cat(sprintf("  '%s'\n", t))
}

cat("\nThese are DIFFERENT publications despite similar titles.\n")
cat("Fuzzy matching could incorrectly merge them.\n\n")
cat("Best practice:\n")
cat("  1. Use default doi_title_year method for automatic cleaning\n")
cat("  2. If adding fuzzy matching, require manual review\n")
cat("  3. Check a random sample of 5-10% of fuzzy matches\n")
```

## Mistake 4: Not Recording Deduplication in Provenance

**Problem**: Applying deduplication without documenting what was changed.

**Why it's wrong**: Without provenance, analyses cannot be reproduced or audited.

**Solution**: Always check `audit_biblio()` after deduplication.

```{r mistake-provenance}
# Demonstrate proper provenance tracking
x <- as_biblio_project(example_biblio(), source = "provenance demo")
x <- deduplicate_biblio(x)

# Check provenance
audit_log <- audit_biblio(x)

cat("Provenance after deduplication:\n")
print(audit_log)

cat("\nThis log records:\n")
cat("  - When deduplication was performed\n")
cat("  - How many records were removed\n")
cat("  - The method used\n")

# Verify provenance has deduplication recorded
has_dedup <- any(audit_log$operation == "deduplicate_biblio")
cat(sprintf("\nDeduplication recorded: %s\n", has_dedup))
```

## Mistake 5: Deduplicating Without Checking First

**Problem**: Applying deduplication without understanding what will be removed.

**Why it's wrong**: You might remove legitimate records or miss important duplicates.

**Solution**: Always run `biblio_health()` first and examine the issues.

```{r mistake-check-first}
# Correct workflow
cat("Correct deduplication workflow:\n\n")
cat("1. Run biblio_health() to identify issues\n")
cat("2. Examine specific problems (missing data, duplicates)\n")
cat("3. Decide on strategy (fix vs. exclude)\n")
cat("4. Apply deduplication\n")
cat("5. Verify results with biblio_health() again\n")
cat("6. Check audit_biblio() for documentation\n")

# Demonstrate
x <- as_biblio_project(example_biblio(), source = "workflow demo")

cat("\n=== Step 1: Health check ===\n")
health <- biblio_health(x)
print(health)

cat("\n=== Step 2: Examine issues ===\n")
problems <- subset(health, n > 0)
if (nrow(problems) > 0) {
  cat("Issues found:\n")
  print(problems)
} else {
  cat("No issues found.\n")
}

cat("\n=== Step 4: Deduplicate ===\n")
x_clean <- deduplicate_biblio(x)

cat("\n=== Step 5: Verify ===\n")
health_after <- biblio_health(x_clean)
print(health_after)

cat("\n=== Step 6: Audit ===\n")
audit <- audit_biblio(x_clean)
cat(sprintf("Operations recorded: %d\n", nrow(audit)))
```


# Advanced Topics {#sec-advanced}

This section covers advanced techniques for users with specific needs.

## Custom Deduplication Keys

While `deduplicate_biblio()` uses a fixed key strategy, you can implement custom approaches:

```{r advanced-custom-keys}
# Create custom deduplication keys
x <- as_biblio_project(example_biblio(), source = "custom keys demo")
works <- x$works

# Strategy 1: Author + Year + First Words of Title
custom_key_1 <- paste(
  substr(tolower(works$title), 1, 30),  # First 30 chars of title
  works$year,
  sep = "|"
)

# Strategy 2: DOI (normalized) or Title Hash + Year
custom_key_2 <- ifelse(
  nzchar(works$doi),
  works$doi,
  paste(
    substr(gsub("[^a-z]", "", tolower(works$title)), 1, 20),
    works$year,
    sep = "|"
  )
)

cat("Custom key examples:\n")
for (i in 1:min(3, nrow(works))) {
  cat(sprintf("Record %d:\n", i))
  cat(sprintf("  Strategy 1: %s\n", substr(custom_key_1[i], 1, 50)))
  cat(sprintf("  Strategy 2: %s\n", substr(custom_key_2[i], 1, 50)))
}
```

## Handling Large Corpora

For very large corpora, deduplication can be computationally expensive:

```{r advanced-large, eval=FALSE}
# Strategies for large corpora (pseudocode)
cat("Large corpus strategies:\n\n")
cat("1. Index on DOI first (O(n) lookup)\n")
cat("2. Use hash tables for title matching\n")
cat("3. Parallelize fuzzy matching (if needed)\n")
cat("4. Process in batches by year range\n")
cat("5. Use the Arrow backend for memory efficiency\n")

# Demonstrate batch processing concept
cat("\nBatch processing example (conceptual):\n")
cat("  years <- sort(unique(works$year))\n")
cat("  results <- lapply(years, function(y) {\n")
cat("    batch <- works[works$year == y, ]\n")
cat("    deduplicate_batch(batch)\n")
cat("  })\n")
cat("  final <- do.call(rbind, results)\n")
```

## Integrating with External Tools

The deduplication log can be exported for external verification:

```{r advanced-export}
# Export deduplication log for external review
x <- as_biblio_project(example_biblio(), source = "export demo")
x <- deduplicate_biblio(x)

# Get the log
dedup_log <- attr(x, "dedup_log")

if (nrow(dedup_log) > 0) {
  # Export to CSV for manual review
  temp_file <- tempfile(fileext = ".csv")
  write.csv(dedup_log, temp_file, row.names = FALSE)
  
  cat("Deduplication log exported to:", temp_file, "\n")
  cat("\nLog contents:\n")
  print(dedup_log)
  
  cat("\nThis log can be:\n")
  cat("  1. Reviewed in Excel/Calc\n")
  cat("  2. Compared with external deduplication tools\n")
  cat("  3. Included in supplementary materials\n")
  cat("  4. Used for inter-rater reliability checks\n")
} else {
  cat("No duplicates found - log is empty.\n")
}
```

## Quality Metrics

Calculate quality metrics to assess corpus health:

```{r advanced-metrics}
# Calculate quality metrics
x <- as_biblio_project(example_biblio(), source = "metrics demo")

# Metadata completeness
total_records <- nrow(x$works)
metrics <- data.frame(
  metric = c(
    "Title completeness",
    "Year completeness", 
    "DOI completeness",
    "Author completeness",
    "Keyword completeness"
  ),
  count = c(
    sum(nzchar(trimws(x$works$title))),
    sum(!is.na(x$works$year)),
    sum(nzchar(x$works$doi)),
    nrow(x$authorships[x$authorships$work_id %in% x$works$work_id, ]),
    nrow(x$keywords[x$keywords$work_id %in% x$works$work_id, ])
  ),
  stringsAsFactors = FALSE
)

# Calculate percentages
metrics$percent <- 100 * metrics$count / total_records

cat("Metadata completeness:\n")
print(metrics)
cat(sprintf("\nOverall completeness: %.1f%%\n", mean(metrics$percent)))
```


# Worked Example with Realistic Data {#sec-worked}

This section provides a complete, realistic workflow using the example corpus with simulated quality issues.

## Scenario Description

We simulate a common scenario: merging data from two database exports (Web of Science and Scopus) for a bibliometric study of agronomy research.

```{r worked-scenario}
# Simulate merging two database exports
cat("Scenario: Merging WoS and Scopus exports\n\n")

# Load base data
base_data <- example_biblio()

# Simulate WoS export (first 8 records, some duplicates)
wos_data <- base_data[1:8, ]
wos_data$source <- "Web of Science"

# Simulate Scopus export (records 5-12, with duplicates)
scopus_data <- base_data[5:12, ]
scopus_data$source <- "Scopus"

# Add realistic quality issues
# 1. Slightly different DOI format in Scopus
scopus_data$doi <- paste0("https://doi.org/", scopus_data$doi)

# 2. Missing DOIs for older records in Scopus
scopus_data$doi[1:2] <- ""

# 3. Title variation in one record
scopus_data$title[3] <- toupper(scopus_data$title[3])

# 4. Missing year in one record
scopus_data$year[4] <- NA

# 5. Duplicate entry in Scopus
scopus_data <- rbind(scopus_data, scopus_data[1, ])

# Merge exports
merged_data <- rbind(wos_data, scopus_data)
cat("Records after merge:\n")
cat(sprintf("  WoS: %d records\n", nrow(wos_data)))
cat(sprintf("  Scopus: %d records\n", nrow(scopus_data)))
cat(sprintf("  Total: %d records\n", nrow(merged_data)))
cat(sprintf("  Expected unique: ~12 records\n"))
```

## Step 1: Harmonize the Merged Data

```{r worked-harmonize}
# Step 1: Harmonize
cat("\n=== Step 1: Harmonization ===\n\n")

x_merged <- as_biblio_project(merged_data, source = "WoS+Scopus merge")
cat("Harmonization results:\n")
cat(sprintf("  Works: %d records\n", nrow(x_merged$works)))
cat(sprintf("  Authors: %d unique\n", nrow(x_merged$authors)))
cat(sprintf("  Keywords: %d links\n", nrow(x_merged$keywords)))
```

## Step 2: Initial Health Assessment

```{r worked-health1}
# Step 2: Health assessment
cat("\n=== Step 2: Health Assessment ===\n\n")

health_initial <- biblio_health(x_merged)
print(health_initial)

# Identify critical issues
critical <- subset(health_initial, n > 0 & check %in% 
                     c("duplicate_doi", "duplicate_title_year"))
if (nrow(critical) > 0) {
  cat("\nCritical issues requiring action:\n")
  print(critical)
}
```

## Step 3: Deduplication

```{r worked-dedup}
# Step 3: Deduplication
cat("\n=== Step 3: Deduplication ===\n\n")

x_clean <- deduplicate_biblio(x_merged)

cat("Deduplication results:\n")
cat(sprintf("  Input records: %d\n", nrow(x_merged$works)))
cat(sprintf("  Output records: %d\n", nrow(x_clean$works)))
cat(sprintf("  Records removed: %d\n", 
            nrow(x_merged$works) - nrow(x_clean$works)))
cat(sprintf("  Reduction: %.1f%%\n", 
            100 * (nrow(x_merged$works) - nrow(x_clean$works)) / nrow(x_merged$works)))

# Show what was removed
dedup_log <- attr(x_clean, "dedup_log")
if (nrow(dedup_log) > 0) {
  cat("\nRemoved records:\n")
  print(dedup_log)
}
```

## Step 4: Post-Deduplication Verification

```{r worked-verify}
# Step 4: Verification
cat("\n=== Step 4: Verification ===\n\n")

# Health check after deduplication
health_final <- biblio_health(x_clean)
cat("Post-deduplication health:\n")
print(health_final)

# Verify no duplicates remain
remaining_dups <- health_final$n[health_final$check == "duplicate_doi"] +
                  health_final$n[health_final$check == "duplicate_title_year"]
cat(sprintf("\nRemaining duplicates: %d\n", remaining_dups))
cat(sprintf("Status: %s\n", ifelse(remaining_dups == 0, "CLEAN", "NEEDS ATTENTION")))

# Check referential integrity
orphan_auth <- !x_clean$authorships$work_id %in% x_clean$works$work_id
orphan_kw <- !x_clean$keywords$work_id %in% x_clean$works$work_id
cat(sprintf("\nReferential integrity:\n"))
cat(sprintf("  Orphan authorships: %d\n", sum(orphan_auth)))
cat(sprintf("  Orphan keywords: %d\n", sum(orphan_kw)))
```

## Step 5: Provenance Documentation

```{r worked-provenance}
# Step 5: Documentation
cat("\n=== Step 5: Provenance Documentation ===\n\n")

audit_log <- audit_biblio(x_clean)
cat("Complete audit trail:\n")
print(audit_log)

cat("\nSummary of transformations:\n")
for (op in unique(audit_log$operation)) {
  count <- sum(audit_log$operation == op)
  cat(sprintf("  %s: %d operation(s)\n", op, count))
}
```

## Step 6: Prepare for Analysis

```{r worked-prepare}
# Step 6: Ready for analysis
cat("\n=== Step 6: Corpus Ready for Analysis ===\n\n")

cat("Final corpus statistics:\n")
cat(sprintf("  Works: %d\n", nrow(x_clean$works)))
cat(sprintf("  Year range: %d-%d\n", 
            min(x_clean$works$year, na.rm = TRUE),
            max(x_clean$works$year, na.rm = TRUE)))
cat(sprintf("  Sources: %d journals\n", length(unique(x_clean$works$source))))
cat(sprintf("  Total citations: %d\n", 
            sum(x_clean$works$cited_by_count, na.rm = TRUE)))

cat("\nThe corpus is now ready for:\n")
cat("  - Descriptive analysis (describe_biblio)\n")
cat("  - Impact metrics (biblio_metrics)\n")
cat("  - Network analysis (bibliographic_network)\n")
cat("  - Comparative inference (compare_groups)\n")
cat("  - Temporal analysis (trend_topics)\n")
```

# References {#sec-references}

## Key Literature

The following references provide background on bibliometric data quality and deduplication:

1. **Aria, M., & Cuccurullo, C. (2017)**. bibliometrix: An R-tool for comprehensive science mapping analysis. *Journal of Informetrics*, 11(4), 959-975. doi: [10.1016/j.joi.2017.08.007](https://doi.org/10.1016/j.joi.2017.08.007)

   - Introduces the bibliometrix package and discusses data quality in science mapping
   - Provides the foundation for many bibliometric workflows in R

2. **Mongeon, P., & Paul-Hus, A. (2016)**. The journal coverage of Web of Science and Scopus: A comparative analysis. *Scientometrics*, 106(1), 213-228. doi: [10.1007/s11192-015-1765-5](https://doi.org/10.1007/s11192-015-1765-5)

   - Compares database coverage and discusses implications for bibliometric validity

3. **Visser, M., Van Eck, N. J., & Waltman, L. (2021)**. Large-scale comparison of bibliographic data sources: Scopus, Web of Science, Dimensions, Crossref, and Microsoft Academic. *Quantitative Science Studies*, 2(1), 20-41. doi: [10.1162/qss_a_00112](https://doi.org/10.1162/qss_a_00112)

   - Latest comprehensive comparison of major bibliographic databases

## Package Documentation

- **Package overview**: `vignette("v00-overview")`
- **Import and harmonization**: `vignette("v01-import-harmonize")`
- **Descriptive and impact analysis**: `vignette("v03-descriptive-impact")`
- **Comparative inference**: `vignette("v04-comparative-inference")`
- **Network analysis**: `vignette("v05-networks")`

## Function References

- `as_biblio_project()`: `?as_biblio_project`
- `biblio_health()`: `?biblio_health`
- `deduplicate_biblio()`: `?deduplicate_biblio`
- `audit_biblio()`: `?audit_biblio`
- `biblio_import()`: `?biblio_import`

## External Resources

- **DOI Foundation**: https://www.doi.org/
- **CrossRef**: https://www.crossref.org/
- **OpenAlex**: https://openalex.org/

## Reproducibility

To reproduce the examples in this vignette:

```{r reproducibility, eval=FALSE}
# Install biblioIntegrator
# install.packages("biblioIntegrator")

# Load the package
library(biblioIntegrator)

# Set seed for reproducibility (if using random processes)
set.seed(42)

# Run examples as shown in each section
```

For the most up-to-date version of this vignette, visit the package documentation:
```{r eval=FALSE}
browseVignettes("biblioIntegrator")
```

## Session Info

```{r session-info}
sessionInfo()
```

---

*Vignette generated with `biblioIntegrator` version `r packageVersion("biblioIntegrator")`.*
