---
title: "Descriptive Analysis and Impact Metrics"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive Analysis and Impact Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150
)
library(biblioIntegrator)
```

# 1. Why this vignette exists

## The role of descriptive analysis in bibliometrics

Descriptive analysis serves as the essential first step in any bibliometric study. Before conducting complex network analyses, comparative inference, or predictive modeling, researchers must understand the fundamental characteristics of their corpus. This vignette provides a comprehensive guide to the descriptive and impact analysis capabilities of `biblioIntegrator`.

### Understanding your corpus

Every bibliographic corpus has a story to tell. Descriptive statistics reveal:

- **Publication patterns**: How has research output evolved over time?
- **Thematic focus**: What topics dominate the literature?
- **Production sources**: Which journals, conferences, or venues publish the most?
- **Collaboration structures**: How do authors work together?
- **Citation impact**: How influential are the publications?

### Why impact metrics matter

Impact metrics quantify scholarly influence beyond simple publication counts. They help answer questions like:

- Which authors are most productive and influential?
- How do citations accumulate over time?
- What is the relative impact of publications across different fields or years?
- Are there emerging topics gaining traction?

### Text analysis as a window into content

Bibliographic databases contain rich textual information in titles, abstracts, and keywords. Text analysis techniques—term frequency, TF-IDF, and trend analysis—reveal:

- **Core terminology**: The most common terms in a field
- **Distinctive vocabulary**: Terms that characterize specific subgroups or time periods
- **Emerging themes**: Topics gaining prominence in recent publications

## When to use this vignette

Use this vignette when you need to:

1. Generate a quick overview of a new bibliographic corpus
2. Compute standard bibliometric metrics for publication or grant reports
3. Identify the most cited works, authors, or sources
4. Analyze thematic patterns in titles or abstracts
5. Compare impact across time periods or subgroups (preliminary)
6. Present descriptive findings in a publication or thesis

## What this vignette does NOT cover

This vignette focuses on *descriptive* and *text-based* analysis. For more advanced analyses, see:

- **Comparative inference** (`v04`): Formal statistical tests between groups
- **Network analysis** (`v05`): Co-authorship, citation, and keyword networks
- **Temporal patterns** (`v06`): Growth models and reference publication year spectroscopy

---

# 2. Learning objectives

By the end of this vignette, you will be able to:

1. **Generate descriptive summaries** of a bibliographic corpus using `describe_biblio()`
2. **Compute author-level metrics** including h-index, g-index, and m-index with `biblio_metrics()`
3. **Analyze term frequency** patterns in titles and abstracts using `term_frequency()`
4. **Apply TF-IDF analysis** to identify distinctive terms across groups using `tfidf_terms()`
5. **Identify trending topics** over time using `trend_topics()`
6. **Understand normalized citations** and their interpretation with `normalized_citations()`
7. **Compute disruption indices** to measure paradigm-shifting influence using `disruption_index()`
8. **Interpret metrics in context**, avoiding common pitfalls of bibliometric indicators
9. **Apply text analysis** to reveal thematic structure in bibliographic data
10. **Create visualizations** of descriptive findings using base R graphics

---

# 3. Prerequisites

## Required packages

```{r prerequisites}
# Core package
library(biblioIntegrator)

# Optional: enhanced visualizations
if (requireNamespace("plotly", quietly = TRUE)) {
  library(plotly)
}
```

## Example data

Throughout this vignette, we use the built-in example corpus:

```{r example-data}
# Load the example corpus
raw_data <- example_biblio()

# Inspect the structure
str(raw_data)
head(raw_data[, c("title", "year", "citations")])
```

The example corpus contains 12 agronomy-related publications spanning 2017--2025, with fields:

- `title`: Publication title
- `year`: Publication year
- `doi`: Digital Object Identifier
- `authors`: Semicolon-separated author names
- `keywords`: Semicolon-separated author keywords
- `citations`: Citation count
- `source`: Journal or venue name

## Constructing a project

All analytical functions require a `biblio_project` object:

```{r construct-project}
# Convert raw data to a project
x <- as_biblio_project(raw_data, source = "tutorial corpus")

# Inspect the project structure
x
head(x$works)
head(x$authorships)
head(x$keywords)
```

---

# 4. Descriptive analysis with `describe_biblio()`

## Function overview

`describe_biblio()` generates a comprehensive descriptive summary of a bibliographic corpus. It returns a list with:

| Element | Description |
|---------|-------------|
| `n_documents` | Total number of works |
| `years` | Range of publication years |
| `total_citations` | Sum of all citations |
| `annual` | Annual publication and citation counts |
| `top_sources` | Frequency table of sources |
| `top_keywords` | Frequency table of keywords |

## Basic usage

```{r describe-basic}
# Generate descriptive summary
desc <- describe_biblio(x)

# Total documents
desc$n_documents

# Year range
desc$years

# Total citations
desc$total_citations
```

## Annual production

```{r describe-annual}
# Annual publication and citation counts
desc$annual
```

The `annual` data frame shows:

- `year`: Publication year
- `documents`: Number of publications
- `citations`: Total citations received

### Visualizing annual production

```{r annual-plot, fig.cap="Figure 1. Annual publication output"}
# Plot annual publications
barplot(
  desc$annual$documents,
  names.arg = desc$annual$year,
  main = "Annual Publication Output",
  xlab = "Year",
  ylab = "Number of Publications",
  col = "steelblue",
  border = "white"
)
```

```{r citations-plot, fig.cap="Figure 2. Annual citation accumulation"}
# Plot annual citations
barplot(
  desc$annual$citations,
  names.arg = desc$annual$year,
  main = "Annual Citation Accumulation",
  xlab = "Year",
  ylab = "Total Citations",
  col = "darkorange",
  border = "white"
)
```

## Top sources

```{r describe-sources}
# Most frequent publication sources
desc$top_sources
```

### Interpreting source distribution

A concentrated source distribution may indicate:

- **Field-specific journals**: Specialization in a narrow subfield
- **Database bias**: Overrepresentation of certain venues in the source database
- **Regional patterns**: Certain journals dominate in particular regions

```{r sources-plot, fig.cap="Figure 3. Top publication sources"}
# Visualize top sources (top 8)
top_n <- min(8, length(desc$top_sources))
barplot(
  head(desc$top_sources, top_n),
  main = "Top Publication Sources",
  xlab = "Source",
  ylab = "Number of Publications",
  col = "darkgreen",
  border = "white",
  las = 2,
  cex.names = 0.8
)
```

## Top keywords

```{r describe-keywords}
# Most frequent author keywords
desc$top_keywords
```

### Keyword analysis patterns

High-frequency keywords reveal:

- **Core topics**: Central themes in the research domain
- **Methodological approaches**: Common methods or techniques
- **Emerging themes**: New keywords appearing in recent years (see Section 7)

```{r keywords-plot, fig.cap="Figure 4. Top author keywords"}
# Visualize top keywords (top 10)
top_kw <- min(10, length(desc$top_keywords))
barplot(
  head(desc$top_keywords, top_kw),
  main = "Top Author Keywords",
  xlab = "Keyword",
  ylab = "Frequency",
  col = "purple",
  border = "white",
  las = 2,
  cex.names = 0.7
)
```

## Collaboration patterns from descriptions

While `describe_biblio()` does not directly compute collaboration metrics, the underlying data enable basic collaboration analysis:

```{r collaboration-basic}
# Average authors per document
avg_authors <- nrow(x$authorships) / nrow(x$works)
cat("Average authors per document:", round(avg_authors, 2), "\n")

# Author frequency
author_freq <- sort(table(x$authorships$author_id), decreasing = TRUE)
head(author_freq)

# Map author IDs to names
author_names <- x$authors$display_name[
  match(names(author_freq), x$authors$author_id)
]
data.frame(
  author = author_names,
  publications = as.integer(author_freq)
)
```

## Geographic distribution (if available)

If your data includes country or affiliation information, geographic analysis is possible:

```{r geographic-example, eval=FALSE}
# Example: if affiliations are available in your data
# geo_data <- x$works$country
# table(geo_data)
#
# # For OpenAlex data, affiliations are in the works table
# if ("country_code" %in% names(x$works)) {
#   geo_table <- sort(table(x$works$country_code), decreasing = TRUE)
#   head(geo_table, 10)
# }
```

---

# 5. Author-level metrics with `biblio_metrics()`

## Function overview

`biblio_metrics()` computes author-level bibliometric indices. It returns a data frame with:

| Column | Description |
|--------|-------------|
| `author` | Author display name |
| `documents` | Number of publications |
| `citations` | Total citations received |
| `h_index` | Hirsch index |
| `g_index` | Egghe's g-index |
| `m_index` | h-index / career length |

## Basic usage

```{r metrics-basic}
# Compute author-level metrics
metrics <- biblio_metrics(x)

# View all metrics
metrics
```

## Understanding h-index

The **h-index** (Hirsch, 2005) is defined as:

> An author has index *h* if *h* of their papers have at least *h* citations each.

```{r h-index-explanation}
# Example: understanding h-index calculation
# Take an author's citation counts sorted descending
example_citations <- sort(metrics$citations, decreasing = TRUE)[1]
example_h <- metrics$h_index[1]

cat("Example author:", metrics$author[1], "\n")
cat("Documents:", metrics$documents[1], "\n")
cat("Total citations:", metrics$citations[1], "\n")
cat("h-index:", example_h, "\n")
```

### Properties of h-index

- **Robustness**: Insensitive to a single highly-cited paper
- **Cumulative**: Only increases over time (cannot decrease)
- **Field-dependent**: Varying benchmark values across disciplines
- **Career-stage dependent**: Favors senior researchers

### Limitations of h-index

```{r h-index-limitations, eval=FALSE}
# The h-index cannot capture:
# 1. A researcher with 1 paper with 1000 citations
#    h-index = 1, despite high impact
#
# 2. A researcher with 100 papers each with 10 citations
#    h-index = 10, same as someone with 10 papers with 10 citations
#
# 3. Different career lengths
#    h-index naturally increases with time active
```

## Understanding g-index

The **g-index** (Egghe, 2006) addresses some h-index limitations:

> An author has index *g* if the top *g* papers account for at least *g*² citations cumulatively.

```{r g-index-explanation}
# Compare h and g indices
metrics[, c("author", "h_index", "g_index")]

# The g-index is always >= h-index
all(metrics$g_index >= metrics$h_index)
```

### Properties of g-index

- **Rewards top papers**: Gives more weight to highly-cited works
- **Sensitive to outliers**: A breakthrough paper increases g more than h
- **Also cumulative**: Only increases over time

## Understanding m-index

The **m-index** (h-index divided by career length) normalizes for career stage:

```{r m-index-explanation}
# m-index: h-index per active year
metrics[, c("author", "h_index", "m_index")]

# Interpretation: higher m-index suggests faster impact accumulation
# But requires careful interpretation for very short careers
```

### Interpreting m-index

```{r m-index-interpretation}
# Career length calculation
# (internal to biblio_metrics: max_year - min_year + 1)
# m-index = h_index / career_years

# Example interpretation
cat("Author with h=3 in 5 years: m =", round(3/5, 2), "\n")
cat("Author with h=3 in 2 years: m =", round(3/2, 2), "\n")
cat("The second author accumulates impact faster\n")
```

## Filtering and ranking authors

```{r metrics-filtering}
# Authors with at least 2 documents
prolific <- subset(metrics, documents >= 2)
prolific

# Top authors by citations
metrics[order(-metrics$citations), ]

# Top authors by h-index
metrics[order(-metrics$h_index), ]
```

## Visualizing author metrics

```{r metrics-plot, fig.cap="Figure 5. Author metrics: documents vs citations", fig.height=6}
# Scatter plot: documents vs citations
plot(
  metrics$documents,
  metrics$citations,
  main = "Author Productivity vs. Impact",
  xlab = "Number of Documents",
  ylab = "Total Citations",
  pch = 19,
  col = "steelblue",
  cex = 1.5
)
text(
  metrics$documents,
  metrics$citations,
  labels = metrics$author,
  pos = 4,
  cex = 0.7
)
```

```{r h-index-plot, fig.cap="Figure 6. Author h-index comparison"}
# Barplot of h-indices
barplot(
  metrics$h_index,
  names.arg = metrics$author,
  main = "Author h-indices",
  xlab = "Author",
  ylab = "h-index",
  col = "darkred",
  border = "white",
  las = 2,
  cex.names = 0.7
)
```

---

# 6. Citation analysis

## Raw citation counts

Citation counts are the simplest measure of impact, but require careful interpretation:

```{r raw-citations}
# Distribution of citations
summary(x$works$cited_by_count)

# Highly cited works
head(x$works[order(-x$works$cited_by_count), c("title", "year", "cited_by_count")], 5)

# Citation distribution
hist(
  x$works$cited_by_count,
  main = "Distribution of Citation Counts",
  xlab = "Citations",
  ylab = "Number of Works",
  col = "lightblue",
  border = "white",
  breaks = 10
)
```

## Why raw counts are problematic

Raw citation counts are influenced by:

1. **Publication year**: Older papers accumulate more citations
2. **Field of study**: Citation rates vary dramatically across disciplines
3. **Document type**: Reviews receive more citations than original research
4. **Database coverage**: Some databases count more citing sources
5. **Language bias**: English-language papers receive more citations globally

```{r citation-problems}
# The problem: older papers appear more "impactful"
# due to time, not necessarily quality
x$works[, c("title", "year", "cited_by_count")]
```

---

# 7. Normalized citations with `normalized_citations()`

## Why normalization matters

Normalized citations address the confounding effect of time and field. The function computes:

$$\text{Normalized}_i = \frac{\text{Citations}_i}{\text{Expected}_i}$$

where Expected is the mean citation count within the chosen stratum (default: publication year).

## Basic usage

```{r normalized-basic}
# Compute normalized citations (by year)
norm_cits <- normalized_citations(x)
head(norm_cits)
```

## Interpreting normalized values

```{r normalized-interpretation}
# Values > 1: above average for the stratum
# Values = 1: average
# Values < 1: below average

summary(norm_cits$normalized)

# Works above average
above_avg <- subset(norm_cits, normalized > 1 & !is.na(normalized))
nrow(above_avg)
above_avg

# Works below average
below_avg <- subset(norm_cits, normalized < 1 & !is.na(normalized))
nrow(below_avg)
```

### Contextual interpretation

```{r normalized-context}
# Merge with titles for context
norm_with_titles <- merge(
  norm_cits,
  x$works[, c("work_id", "title", "year", "cited_by_count")],
  by = "work_id"
)

# View normalized scores with context
norm_with_titles[
  order(-norm_with_titles$normalized),
  c("title", "year", "cited_by_count", "expected", "normalized")
]
```

## Choosing strata

The default stratum is `year`, but you can normalize by other fields:

```{r normalized-strata, eval=FALSE}
# Normalize by source (journal)
# norm_by_source <- normalized_citations(x, strata = "source")
# head(norm_by_source)

# Normalize by multiple strata
# norm_multi <- normalized_citations(x, strata = c("year", "source"))
# head(norm_multi)
```

## Visualizing normalized citations

```{r normalized-plot, fig.cap="Figure 7. Raw vs normalized citations"}
# Compare raw and normalized
par(mfrow = c(1, 2))

# Raw citations
hist(
  norm_cits$citations,
  main = "Raw Citations",
  xlab = "Citations",
  col = "lightcoral",
  border = "white",
  breaks = 8
)

# Normalized citations
hist(
  norm_cits$normalized[!is.na(norm_cits$normalized)],
  main = "Normalized Citations",
  xlab = "Normalized Score",
  col = "lightgreen",
  border = "white",
  breaks = 8
)
abline(v = 1, lty = 2, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

```{r normalized-bar, fig.cap="Figure 8. Normalized citation scores by work"}
# Barplot of normalized scores
norm_sorted <- norm_cits[order(-norm_cits$normalized), ]
norm_sorted <- norm_sorted[!is.na(norm_sorted$normalized), ]

barplot(
  norm_sorted$normalized,
  names.arg = paste0("W", seq_len(nrow(norm_sorted))),
  main = "Normalized Citation Scores",
  xlab = "Works (ranked)",
  ylab = "Normalized Score",
  col = ifelse(norm_sorted$normalized > 1, "darkgreen", "gray"),
  border = "white",
  las = 2,
  cex.names = 0.6
)
abline(h = 1, lty = 2, col = "red", lwd = 2)
text(
  x = 0.5,
  y = 1.05,
  labels = "Average = 1",
  pos = 4,
  col = "red",
  cex = 0.8
)
```

---

# 8. Term frequency with `term_frequency()`

## Function overview

`term_frequency()` extracts and counts terms from titles or abstracts. It provides a simple but effective way to identify the most common vocabulary in a corpus.

## Basic usage

```{r term-freq-basic}
# Term frequency from titles
tf_title <- term_frequency(x)
head(tf_title, 15)
```

## Parameters

```{r term-freq-params}
# Default stopwords are removed
# You can customize them:
tf_custom <- term_frequency(
  x,
  field = "title",
  stopwords = c("and", "the", "for", "with", "under", "of", "in", "to", "a", "an")
)
head(tf_custom, 10)

# Use abstracts (if available)
tf_abstract <- term_frequency(x, field = "abstract")
head(tf_abstract, 10)
```

## Visualizing term frequency

```{r term-freq-plot, fig.cap="Figure 9. Top terms in titles"}
# Plot top 15 terms
top_terms <- head(tf_title, 15)

barplot(
  top_terms$n,
  names.arg = top_terms$term,
  main = "Most Frequent Terms in Titles",
  xlab = "Term",
  ylab = "Frequency",
  col = "darkblue",
  border = "white",
  las = 2,
  cex.names = 0.7
)
```

## Interpreting term frequency

### What frequent terms reveal

- **Core vocabulary**: Terms central to the research field (e.g., "soil", "silicon", "crop")
- **Methodological focus**: Terms indicating research approach (e.g., "analysis", "modeling")
- **Study systems**: Organisms or environments studied (e.g., "rice", "maize", "soybean")

### Limitations of term frequency

```{r term-freq-limitations}
# Common terms are not necessarily distinctive
# "analysis" appears frequently but doesn't characterize any specific topic
# See tfidf_terms() for distinctive terms

# Terms may have multiple meanings
# "model" could mean statistical model, crop model, etc.
# Context from abstracts often needed
```

## Combining with keyword analysis

```{r term-keyword-comparison}
# Compare title terms with author keywords
cat("Top title terms:\n")
head(tf_title, 10)

cat("\nTop author keywords:\n")
head(desc$top_keywords, 10)
```

---

# 9. TF-IDF analysis with `tfidf_terms()`

## The TF-IDF concept

**TF-IDF** (Term Frequency--Inverse Document Frequency) measures how distinctive a term is for a particular group relative to the entire corpus.

The formula is:

$$\text{TF-IDF}_{t,g} = \text{TF}_{t,g} \times \log\left(\frac{N}{\text{DF}_t}\right)$$

where:

- $\text{TF}_{t,g}$ = frequency of term $t$ in group $g$
- $N$ = total number of groups
- $\text{DF}_t$ = number of groups containing term $t$

## Basic usage

```{r tfidf-basic}
# TF-IDF by year (default)
tfidf_year <- tfidf_terms(x)
head(tfidf_year, 15)
```

## Grouping options

```{r tfidf-groups}
# TF-IDF by source
tfidf_source <- tfidf_terms(x, group = "source")
head(tfidf_source, 15)
```

## Interpreting TF-IDF

### High TF-IDF values

High TF-IDF indicates terms that are:

- **Frequent** in a specific group
- **Rare** across other groups
- **Distinctive** markers of that group

```{r tfidf-interpretation}
# Top distinctive terms by year
# These terms characterize specific years
head(tfidf_year[order(-tfidf_year$tfidf), ], 10)

# Terms with zero TF-IDF appear in all groups equally
# They are common and not distinctive
zero_tfidf <- subset(tfidf_year, tfidf == 0)
head(zero_tfidf)
```

### Examining specific years

```{r tfidf-by-year}
# Distinctive terms for each year
years <- sort(unique(tfidf_year$group))

for (yr in head(years, 5)) {
  cat("\nYear", yr, "- Top distinctive terms:\n")
  yr_terms <- subset(tfidf_year, group == yr)
  yr_terms <- yr_terms[order(-yr_terms$tfidf), ]
  print(head(yr_terms[, c("term", "n", "tfidf")], 5))
}
```

## Visualizing TF-IDF

```{r tfidf-plot, fig.cap="Figure 10. Top TF-IDF terms by year"}
# Top TF-IDF terms
top_tfidf <- head(tfidf_year[order(-tfidf_year$tfidf), ], 15)

barplot(
  top_tfidf$tfidf,
  names.arg = paste(top_tfidf$term, top_tfidf$group, sep = "\n"),
  main = "Top Distinctive Terms (TF-IDF)",
  xlab = "Term (Year)",
  ylab = "TF-IDF Score",
  col = "darkviolet",
  border = "white",
  las = 2,
  cex.names = 0.6
)
```

```{r tfidf-source-plot, fig.cap="Figure 11. Top TF-IDF terms by source", fig.height=6}
# TF-IDF by source
top_src_tfidf <- head(
  tfidf_source[order(-tfidf_source$tfidf), ],
  12
)

barplot(
  top_src_tfidf$tfidf,
  names.arg = paste(
    top_src_tfidf$term,
    substr(top_src_tfidf$group, 1, 10),
    sep = "\n"
  ),
  main = "Distinctive Terms by Source (TF-IDF)",
  xlab = "Term (Source)",
  ylab = "TF-IDF Score",
  col = "darkcyan",
  border = "white",
  las = 2,
  cex.names = 0.6
)
```

## TF-IDF vs term frequency

```{r tfidf-vs-tf}
# Compare: most frequent terms vs most distinctive terms
cat("Most frequent title terms:\n")
print(head(tf_title, 5))

cat("\nMost distinctive terms (TF-IDF):\n")
print(head(tfidf_year[order(-tfidf_year$tfidf), ], 5))

# Key difference:
# - Term frequency highlights common, possibly generic terms
# - TF-IDF highlights terms distinctive to specific subgroups
```

---

# 10. Trending topics with `trend_topics()`

## Function overview

`trend_topics()` identifies topics that are gaining or losing prominence over time by tracking keyword frequencies across years.

## Basic usage

```{r trends-basic}
# Topic trajectories
trends <- trend_topics(x)
head(trends, 20)
```

## Filtering by minimum frequency

```{r trends-filtering}
# Only keywords appearing at least 2 times total
trends_filtered <- trend_topics(x, min_total = 2)
head(trends_filtered, 15)
```

## Visualizing topic trends

```{r trends-plot, fig.cap="Figure 12. Topic trajectories over time"}
# Select top keywords for visualization
top_kw <- names(head(sort(
  tapply(trends$n, trends$keyword, sum),
  decreasing = TRUE
), 5))

# Subset data
trends_top <- subset(trends, keyword %in% top_kw)

# Create time series plot
years_range <- sort(unique(trends_top$year))
plot(
  NULL,
  xlim = range(years_range),
  ylim = range(trends_top$n),
  main = "Topic Trajectories",
  xlab = "Year",
  ylab = "Frequency",
  type = "n"
)

colors <- c("red", "blue", "darkgreen", "purple", "orange")
for (i in seq_along(top_kw)) {
  kw_data <- subset(trends_top, keyword == top_kw[i])
  lines(
    kw_data$year,
    kw_data$n,
    type = "b",
    col = colors[i],
    pch = 19,
    lwd = 2
  )
}
legend(
  "topleft",
  legend = top_kw,
  col = colors,
  lwd = 2,
  pch = 19,
  cex = 0.8
)
```

## Interpreting trends

### Rising topics

```{r trends-rising}
# Identify topics with increasing frequency
# Simple approach: compare recent vs earlier years
recent_year <- max(x$works$year, na.rm = TRUE)
earlier_year <- recent_year - 2

recent_trends <- subset(trends, year >= earlier_year)
early_trends <- subset(trends, year < earlier_year)

# Frequency in recent vs earlier periods
recent_freq <- tapply(recent_trends$n, recent_trends$keyword, sum)
early_freq <- tapply(early_trends$n, early_trends$keyword, sum)

# Topics appearing in recent but not earlier
new_topics <- setdiff(names(recent_freq), names(early_freq))
if (length(new_topics) > 0) {
  cat("Emerging topics (recent only):\n")
  print(recent_freq[new_topics])
}
```

### Declining topics

```{r trends-declining}
# Topics in earlier but not recent
declining <- setdiff(names(early_freq), names(recent_freq))
if (length(declining) > 0) {
  cat("Declining topics (earlier only):\n")
  print(early_freq[declining])
}
```

### Stable topics

```{r trends-stable}
# Topics present in both periods
stable <- intersect(names(recent_freq), names(early_freq))
if (length(stable) > 0) {
  cat("Stable topics (present in both periods):\n")
  comparison <- data.frame(
    topic = stable,
    earlier = as.integer(early_freq[stable]),
    recent = as.integer(recent_freq[stable])
  )
  comparison$change <- comparison$recent - comparison$earlier
  print(comparison)
}
```

---

# 11. Disruption index with `disruption_index()`

## The disruption concept

The **disruption index** (CD index, Funk & Owen-Smith, 2017; Wu, Wang, & Evans, 2019) measures whether a work creates a new paradigm (disruptive) or consolidates existing knowledge (consolidating).

### Formula

The disruption index uses the CD-style definition:

$$\text{DI} = \frac{N_i - N_j}{N_i + N_j + N_k}$$

where:

- $N_i$ = works that cite the focal work but NOT its references
- $N_j$ = works that cite both the focal work AND its references
- $N_k$ = works that cite the focal work's references but NOT the focal work

### Interpretation

| DI Value | Interpretation |
|----------|----------------|
| Close to +1 | Highly disruptive: new paradigm |
| Close to 0 | Neutral: balanced influence |
| Close to -1 | Highly consolidating: builds on existing work |

## Basic usage

```{r disruption-basic}
# Create example citation network
# This requires citation edges (who cites whom)
citation_edges <- data.frame(
  citing_id = c("w1", "w2", "w3", "w4", "w5", "w2", "w4"),
  cited_id = c("focal", "focal", "focal", "focal", "focal", "ref1", "ref1")
)

# Focal work's references
focal_refs <- c("ref1")

# Compute disruption index
di <- disruption_index("focal", citation_edges, focal_refs)
di
```

## Interpreting results

```{r disruption-interpretation}
# Interpret the disruption index
cat("N_i (cites focal only):", di$N_i, "\n")
cat("N_j (cites both):", di$N_j, "\n")
cat("N_k (cites references only):", di$N_k, "\n")
cat("Disruption index:", round(di$disruption, 3), "\n\n")

if (!is.na(di$disruption) && di$disruption > 0.5) {
  cat("Interpretation: Work is DISRUPTIVE\n")
} else if (!is.na(di$disruption) && di$disruption < -0.5) {
  cat("Interpretation: Work is CONSOLIDATING\n")
} else {
  cat("Interpretation: Work has MIXED influence\n")
}
```

## More complex examples

```{r disruption-complex}
# Example with multiple references
complex_edges <- data.frame(
  citing_id = c(
    "a", "a", "b", "b", "c", "d", "e",
    "c", "d", "f", "g"
  ),
  cited_id = c(
    "focal", "ref1", "focal", "ref2", "focal",
    "focal", "focal", "ref1", "ref2", "ref1", "ref2"
  )
)

disruption_index("focal", complex_edges, c("ref1", "ref2"))
```

## Combining with other metrics

```{r disruption-combined, eval=FALSE}
# For a comprehensive analysis, combine disruption with:
# 1. Citation counts (normalized_citations)
# 2. Author metrics (biblio_metrics)
# 3. Term analysis (tfidf_terms)
#
# This provides a multi-dimensional view of impact:
# - Citations: how much the work is used
# - Disruption: whether it changes the field
# - Terms: what themes it introduces
```

---

# 12. Step-by-step workflow

## Complete example

Here is a full workflow from raw data to comprehensive descriptive analysis:

```{r workflow-complete, eval=FALSE}
# ============================================
# Step 1: Load and prepare data
# ============================================

library(biblioIntegrator)

# Load example data
raw <- example_biblio()
cat("Raw data:", nrow(raw), "records\n")

# Construct project
x <- as_biblio_project(raw, source = "analysis")
x

# ============================================
# Step 2: Descriptive summary
# ============================================

desc <- describe_biblio(x)

cat("\n=== Descriptive Summary ===\n")
cat("Documents:", desc$n_documents, "\n")
cat("Year range:", desc$years[1], "-", desc$years[2], "\n")
cat("Total citations:", desc$total_citations, "\n")

cat("\nAnnual production:\n")
print(desc$annual)

cat("\nTop sources:\n")
print(head(desc$top_sources, 5))

cat("\nTop keywords:\n")
print(head(desc$top_keywords, 10))

# ============================================
# Step 3: Author metrics
# ============================================

metrics <- biblio_metrics(x)

cat("\n=== Author Metrics ===\n")
print(metrics)

cat("\nProlific authors (≥2 papers):\n")
print(subset(metrics, documents >= 2))

# ============================================
# Step 4: Citation analysis
# ============================================

# Raw citations
cat("\n=== Citation Analysis ===\n")
cat("Mean citations:", mean(x$works$cited_by_count), "\n")
cat("Median citations:", median(x$works$cited_by_count), "\n")

# Normalized citations
norm <- normalized_citations(x)
cat("\nNormalized citations (summary):\n")
summary(norm$normalized[!is.na(norm$normalized)])

# ============================================
# Step 5: Text analysis
# ============================================

# Term frequency
tf <- term_frequency(x)
cat("\n=== Term Frequency ===\n")
print(head(tf, 10))

# TF-IDF
tfidf <- tfidf_terms(x)
cat("\n=== TF-IDF (Top 10) ===\n")
print(head(tfidf[order(-tfidf$tfidf), ], 10))

# ============================================
# Step 6: Trend analysis
# ============================================

trends <- trend_topics(x, min_total = 2)
cat("\n=== Trending Topics ===\n")
print(trends)

# ============================================
# Step 7: Visualizations
# ============================================

par(mfrow = c(2, 2))

# Annual publications
barplot(
  desc$annual$documents,
  names.arg = desc$annual$year,
  main = "Annual Output",
  col = "steelblue"
)

# Author h-indices
barplot(
  metrics$h_index,
  names.arg = metrics$author,
  main = "Author h-indices",
  col = "darkred",
  las = 2,
  cex.names = 0.6
)

# Top terms
barplot(
  head(tf$n, 10),
  names.arg = head(tf$term, 10),
  main = "Top Title Terms",
  col = "darkblue",
  las = 2,
  cex.names = 0.7
)

# Normalized citations
hist(
  norm$normalized[!is.na(norm$normalized)],
  main = "Normalized Citations",
  col = "lightgreen",
  breaks = 8
)
abline(v = 1, lty = 2, col = "red")

par(mfrow = c(1, 1))
```

## Workflow with your own data

```{r workflow-own-data, eval=FALSE}
# ============================================
# Working with your own data
# ============================================

# Option 1: From CSV
# my_data <- read.csv("my_bibliography.csv")
# x <- as_biblio_project(my_data, source = "my study")

# Option 2: From bibliometrix (if installed)
# x <- biblio_import(
#   "my_export.bib",
#   dbsource = "scopus",
#   format = "bibtex"
# )

# Then run the same descriptive workflow
# ...
```

---

# 13. Common mistakes and pitfalls

## Mistake 1: Using raw citation counts without normalization

```{r pitfall-raw}
# WRONG: Comparing raw citations across years
# An older paper with 100 citations might be less impactful
# than a recent paper with 20 citations

# RIGHT: Use normalized citations
norm <- normalized_citations(x)
# A normalized score of 1.5 means 50% above the year average
# regardless of how old the paper is
```

## Mistake 2: Ignoring field-specific citation patterns

```{r pitfall-field}
# WRONG: Comparing h-indices across different fields
# Medicine typically has higher citation rates than Mathematics

# RIGHT: Compare within fields or use field-normalized metrics
# bibliometrix provides field-normalized indicators
# When using biblio_metrics(), interpret relative to field norms
```

## Mistake 3: Confusing term frequency with importance

```{r pitfall-frequency}
# WRONG: Assuming the most frequent terms are the most important
tf <- term_frequency(x)
# "analysis" might be frequent but not distinctive

# RIGHT: Use TF-IDF for distinctive terms
tfidf <- tfidf_terms(x)
# High TF-IDF terms are both frequent in a group AND rare elsewhere
```

## Mistake 4: Not considering time window effects

```{r pitfall-time}
# WRONG: Not accounting for publication year in trend analysis
# A keyword appearing in 2024 but not 2018 might simply be new
# rather than "trending"

# RIGHT: Use trend_topics() which tracks temporal patterns
# Look for topics with *increasing* frequency, not just new ones
```

## Mistake 5: Over-interpreting small corpora

```{r pitfall-size}
# WRONG: Drawing strong conclusions from < 50 documents
# Small samples lead to unstable metrics

# RIGHT: Report confidence and sample size
cat("Corpus size:", nrow(x$works), "\n")
cat("Note: Small corpora yield unstable metrics.\n")
cat("Consider bootstrap confidence intervals for small samples.\n")
```

## Mistake 6: Double-counting authors

```{r pitfall-authors}
# WRONG: Counting the same author twice due to name variants
# "Smith J" and "J. Smith" might be the same person

# RIGHT: Check for duplicates
# biblioIntegrator uses author_id deduplication internally
# But name variants require manual curation
cat("Number of unique authors:", nrow(x$authors), "\n")
cat("Check for potential name duplicates:\n")
head(x$authors, 10)
```

## Mistake 7: Ignoring self-citations

```{r pitfall-self}
# WRONG: Including self-citations in impact analysis

# RIGHT: When possible, identify and exclude self-citations
# This requires author-level citation data
# (Not directly available in simple bibliographic exports)
```

## Mistake 8: Treating all document types equally

```{r pitfall-doc-type}
# WRONG: Comparing citations between reviews and original research

# RIGHT: Separate by document type before analysis
# Reviews systematically receive more citations
# biblio_import preserves document type when available
```

---

# 14. Advanced analysis patterns

## Combining multiple functions

```{r advanced-combined}
# Create a comprehensive descriptive profile
create_corpus_profile <- function(x) {
  # Descriptive summary
  desc <- describe_biblio(x)

  # Author metrics
  metrics <- biblio_metrics(x)

  # Normalized citations
  norm <- normalized_citations(x)

  # Term analysis
  tf <- term_frequency(x)
  tfidf <- tfidf_terms(x)

  # Return comprehensive list
  list(
    summary = desc,
    authors = metrics,
    citations = norm,
    terms = tf,
    distinctive_terms = tfidf
  )
}

profile <- create_corpus_profile(x)
str(profile, max.level = 2)
```

## Exporting results

```{r advanced-export, eval=FALSE}
# Export descriptive results to CSV
write.csv(
  describe_biblio(x)$annual,
  "annual_production.csv",
  row.names = FALSE
)

write.csv(
  biblio_metrics(x),
  "author_metrics.csv",
  row.names = FALSE
)

write.csv(
  normalized_citations(x),
  "normalized_citations.csv",
  row.names = FALSE
)
```

## Combining with network analysis

```{r advanced-network, eval=FALSE}
# For a complete analysis, combine descriptive and network approaches:

# 1. Descriptive overview (this vignette)
# desc <- describe_biblio(x)

# 2. Network structure (see v05-networks.Rmd)
# g <- bibliographic_network(x, "coauthor")
# centrality <- network_centrality(g)

# 3. Compare descriptive and network-based author rankings
# metrics <- biblio_metrics(x)
# Compare metrics$h_index with centrality metrics
```

---

# 15. Interpretation guidelines

## Contextualizing metrics

Every metric should be interpreted in context:

| Metric | What it measures | Context needed |
|--------|------------------|----------------|
| Citation count | Attention received | Field, year, document type |
| h-index | Sustained impact | Career length, field norms |
| g-index | Weighted impact | Alternative to h-index |
| m-index | Impact velocity | Career stage |
| Normalized citations | Relative impact | Stratum definition |
| TF-IDF | Term distinctiveness | Group definition |
| Disruption index | Paradigm influence | Citation network completeness |

## Red flags to watch for

```{r red-flags, eval=FALSE}
# Potential data quality issues:
# 1. Very high citation counts for recent papers
#    -> Possible database error or preprint citation
#
# 2. Very low h-indices despite high publication counts
#    -> Possible predatory journal publications
#
# 3. All TF-IDF values near zero
#    -> Possible homogenized corpus or data error
#
# 4. Missing keywords for most documents
#    -> Database export issue or author oversight
#
# 5. Negative or zero normalized citations
#    -> Calculation error or edge case
```

## Reporting best practices

```{r reporting, eval=FALSE}
# When reporting descriptive bibliometric findings:

# 1. Always report corpus size and time range
# "This study analyzed N = X publications from year1 to year2."

# 2. Report central tendency AND dispersion
# "Mean citations were M = X (SD = Y, range = A-B)."

# 3. Use normalized metrics for comparisons
# "After normalization, work A scored 1.5× the year average."

# 4. Acknowledge limitations
# "Results are limited to records in the [database] and may
#  not reflect the complete literature."

# 5. Visualize key findings
# Include publication trends, top authors, and key term clouds
```

---

# 16. References

## Core bibliometric methods

- Aria, M., & Cuccurullo, C. (2017). bibliometrix: An R-tool for comprehensive science mapping analysis. *Journal of Informetrics*, 11(4), 959--975. doi:10.1016/j.joi.2017.08.007

- Hirsch, J. E. (2005). An index to quantify an individual's scientific research output. *Proceedings of the National Academy of Sciences*, 102(46), 16569--16572. doi:10.1073/pnas.0507655102

- Egghe, L. (2006). Theory and practise of the g-index. *Scientometrics*, 69(1), 131--152. doi:10.1007/s11192-006-0144-7

## Disruption and innovation

- Funk, R. J., & Owen-Smith, J. (2017). A dynamic network measure of technological change. *Management Science*, 63(3), 791--817. doi:10.1287/mnsc.2015.2366

- Wu, L., Wang, D., & Evans, J. A. (2019). Large teams develop and small teams disrupt science and technology. *Nature*, 566(7744), 378--382. doi:10.1038/s1586-019-0941-9

## Text analysis in bibliometrics

- Callon, M., Courtial, J. P., Turner, W. A., & Bauin, S. (1983). From translations to problematic networks: An introduction to co-word analysis. *Social Science Information*, 22(2), 191--235. doi:10.1177/053901883022002003

- Rip, A., & Courtial, J. P. (1984). Co-word maps of biotechnology: An example of cognitive scientometrics. *Scientometrics*, 6(6), 381--400. doi:10.1007/BF02025826

## Normalized citation indicators

- Waltman, L., & van Eck, N. J. (2012). The inconsistency of the h-index. *Journal of the American Society for Information Science and Technology*, 63(2), 406--415. doi:10.1002/asi.21678

- Bornmann, L., & Marx, W. (2015). Methods for the generation of normalized citation impact scores in bibliometrics: Which method best reflects the judgments of experts? *Journal of Informetrics*, 9(2), 408--418. doi:10.1016/j.joi.2015.01.006

## biblioIntegrator package

- Pereira, W. E., & Martinez, M. H. P. (2025). biblioIntegrator: Harmonized, Comparative and Network-Based Bibliometric Analysis. R package version 0.2.0.

---

# 17. Session information

```{r session-info}
sessionInfo()
```

# 18. Quick reference card

## Function summary

| Function | Purpose | Returns |
|----------|---------|---------|
| `describe_biblio(x)` | Descriptive summary | List with n, years, citations, annual, top sources/keywords |
| `biblio_metrics(x)` | Author metrics | Data frame with h-index, g-index, m-index |
| `normalized_citations(x, strata)` | Normalized citations | Data frame with raw, expected, normalized |
| `term_frequency(x, field, stopwords)` | Term counts | Data frame with term, n |
| `tfidf_terms(x, group, field)` | Distinctive terms | Data frame with group, term, n, tfidf |
| `trend_topics(x, min_total)` | Topic trajectories | Data frame with year, keyword, n |
| `disruption_index(focal_id, edges, refs)` | Disruption score | Data frame with N_i, N_j, N_k, disruption |

## Common workflows

```{r quick-ref-workflow, eval=FALSE}
# Standard descriptive analysis
x <- as_biblio_project(my_data)
describe_biblio(x)
biblio_metrics(x)
normalized_citations(x)
term_frequency(x)

# Text analysis
tfidf_terms(x, group = "year")
trend_topics(x, min_total = 2)

# Disruption analysis
disruption_index("work_id", citation_edges, references)
```

## Parameter defaults

| Function | Parameter | Default | Options |
|----------|-----------|---------|---------|
| `term_frequency` | field | "title" | "title", "abstract" |
| `term_frequency` | stopwords | c("and","the","for","with","under","of","in","to") | Any character vector |
| `tfidf_terms` | group | "year" | "year", "source" |
| `tfidf_terms` | field | "title" | "title", "abstract" |
| `trend_topics` | min_total | 1 | Any positive integer |
| `normalized_citations` | strata | "year" | Any column in works |

---

# 19. Glossary

- **Bibliometrics**: The application of mathematical and statistical methods to books and other media of communication (Pritchard, 1969).

- **Citation count**: The number of times a publication has been cited by other works.

- **h-index**: An author has index *h* if *h* of their papers have at least *h* citations each.

- **g-index**: The highest number *g* such that the top *g* papers have together at least *g*² citations.

- **m-index**: h-index divided by the number of active years.

- **Normalized citations**: Citation counts adjusted for field and/or time period effects.

- **TF-IDF**: Term Frequency--Inverse Document Frequency; a weight measuring term distinctiveness.

- **Disruption index**: A measure of whether a work creates a new paradigm (positive) or consolidates existing knowledge (negative).

- **Provenance**: A record of the operations applied to data, ensuring reproducibility.

- **Corpus**: The collection of bibliographic records being analyzed.

---

# 20. Next steps

This vignette covered descriptive analysis and impact metrics. To continue your analysis:

1. **Comparative inference** (`v04-comparative-inference.Rmd`): Test whether groups differ significantly in their bibliometric characteristics.

2. **Network analysis** (`v05-networks.Rmd`): Visualize and analyze co-authorship, citation, and keyword networks.

3. **Temporal and text analysis** (`v06-temporal-text.Rmd`): Deep dive into growth models and reference publication year spectroscopy.

4. **Foundations-to-advanced tutorial** (`v07-foundations-to-advanced-tutorial.Rmd`): Complete workflow integrating all analysis types.

---

```{r cleanup, include=FALSE}
# Clean up
rm(list = ls())
```
