---
title: "Comparative inference"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biblioIntegrator)
```

# Purpose

The native engine handles exclusive or overlapping groups. For overlaps it permutes complete membership rows, preserving the overlap pattern.

# Workflow

```{r}
x <- as_biblio_project(example_biblio())
g <- ifelse(x$works$year < 2022, "early", "recent")
z <- compare_groups(x, g, permutations = 99, bootstrap = 49, seed = 10)
z
head(association_residuals(z))
group_ca(z)
sensitivity_analysis(x, g, thresholds = 1:2, permutations = 49, seed = 10)
```

# Interpretation

Results should be interpreted in relation to database coverage, time window, entity normalization and analytical thresholds. Provenance should be retained whenever data are merged, deduplicated or enriched.
