---
title: "Comparative Inference Between Groups"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparative Inference Between Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5,
  out.width  = "100%",
  warning    = FALSE,
  message    = FALSE,
  eval       = FALSE
)
options(warnPartialMatchArgs = FALSE)
```

<!-- ================================================================== -->
<!-- 1. WHY THIS VIGNETTE EXISTS                                       -->
<!-- ================================================================== -->

## 1. Why This Vignette Exists

### 1.1 The Core Question of Bibliometric Enquiry

Every bibliometric investigation eventually confronts a deceptively
simple question: **"What really distinguishes one subset of publications
from another?"**

Whether we are comparing early-career and established researchers,
open-access and subscription publications, domestic and international
collaborations, or studies funded by different agencies, we need a
principled statistical framework that separates meaningful patterns from
random noise.  Without such a framework, reported "differences" may be
nothing more than sampling fluctuation dressed up in confident language.

### 1.2 Why Common Approaches Fail

Rudimentary bibliometric reports frequently present descriptive
statistics side-by-side---mean citations differ between Groups A and B,
the proportion of international co-authors differs, keyword frequencies
differ---but descriptive differences alone are *never* sufficient.
A difference observed in one particular sample might easily arise by
chance, especially when the number of publications per group is modest
or when multiple comparisons are made.

Traditional parametric tests (Student's *t*, classical chi-squared,
one-way ANOVA) require distributional assumptions that bibliometric data
consistently violate:

| Assumption | Reality in bibliometric data |
|:---|:---|
| Normality | Citation counts follow heavy-tailed, right-skewed distributions (often power-law or negative binomial) |
| Homoscedasticity | Variances grow systematically with means; high-impact journals create heteroscedastic clusters |
| Independence | Collaborators share citations; institutional affiliations create publication clusters |
| Large-cell counts | Cross-tabulations of rare keywords or niche journals produce many zero or near-zero cells |

Permutation (randomization) inference sidesteps *all* these problems by
deriving the null distribution empirically---directly from the data
themselves.

### 1.3 What This Vignette Covers

This vignette presents a complete workflow for comparative inference in
bibliometric research using `biblioIntegrator`.  We cover:

1. Defining exclusive and overlapping groups from bibliometric
   metadata.
2. Constructing contingency tables and computing expected counts.
3. Permutation-based group comparison with the native R engine.
4. Interpreting association residuals and Cramér's *V*.
5. Correspondence analysis (CA) and multiple correspondence analysis
   (MCA) for dimensionality reduction.
6. Cross-validating results against the Biblium Python engine.
7. Assessing sensitivity to classification thresholds.
8. Recognising common methodological pitfalls.

The companion vignettes `v01-data-import`,
`v02-bibliometric-overview`, and `v03-network-analysis` cover data
ingestion, descriptive bibliometrics, and network-based analyses,
respectively.

---

<!-- ================================================================== -->
<!-- 2. LEARNING OBJECTIVES                                             -->
<!-- ================================================================== -->

## 2. Learning Objectives

After completing this vignette you will be able to:

1. **Define exclusive groups** (each publication in exactly one group)
   and **overlapping groups** (each publication may belong to zero or
   more groups) from bibliometric metadata.
2. **Construct contingency tables** and compute expected counts and
   standardised residuals by hand, then verify against `biblioIntegrator`
   output.
3. **Run `compare_groups()`** with the native R engine and interpret
   chi-squared, Cramér's *V*, and permutation *p*-values.
4. **Extract and visualise** association residuals using
   `association_residuals()` and heatmaps.
5. **Perform correspondence analysis** on a group contingency table
   via `group_ca()` and interpret biplots.
6. **Perform multiple correspondence analysis** via `group_mca()` for
   polytomous categorical attributes.
7. **Cross-validate** native R and Biblium Python engines using
   `validate_biblium()` and `biblium_compare_groups()`.
8. **Assess threshold sensitivity** with `sensitivity_analysis()` and
   recognise when findings are artefacts of arbitrary boundary choices.
9. **Recognise common pitfalls** including multiple-comparisons
   inflation, confusion of statistical significance with practical
   importance, and failure to validate against alternative engines.

---

<!-- ================================================================== -->
<!-- 3. OVERVIEW OF THE PERMUTATION FRAMEWORK                           -->
<!-- ================================================================== -->

## 3. The Permutation Inference Framework

### 3.1 Fisher's Logic

Permutation inference follows directly from the logic of design-based
inference (Fisher, 1935):

1. **Null hypothesis**: group labels carry no information about the
   observed publication attributes.  Any assignment of labels to the
   observed data would produce equally extreme statistics.
2. **Test statistic**: any scalar summary of the group--attribute
   relationship (chi-squared, Cramér's *V*, or a custom statistic).
3. **Null distribution**: generated by recomputing the statistic after
   randomly permuting group labels.
4. **Inference**: compare the observed statistic against this null
   distribution.

### 3.2 Validity Assumptions

Permutation tests require exactly one assumption: *exchangeability
under the null hypothesis*---the joint distribution of observations
remains unchanged when labels are permuted.  This is a much weaker
requirement than normality, homoscedasticity, or classical independence,
making it ideal for bibliometric data.

### 3.3 The Role of B (Number of Permutations)

| *B* | Minimum *p* | Precision | Compute time |
|:---:|:---:|:---:|:---:|
| 999 | 0.001 | ±0.0007 | fast |
| 9 999 | 0.0001 | ±0.00007 | moderate |
| 99 999 | 0.00001 | ±0.000007 | slow |

Convention: *B* = 9 999 provides good precision for typical
α = 0.05 testing.  If *p* is close to the significance boundary,
increase *B*.

The permutation *p*-value is computed as

$$
p_{\text{perm}} = \frac{1 + \sum_{b=1}^{B}\,
\mathbf{1}\!\bigl(\chi^2_b \geq \chi^2_{\text{obs}}\bigr)}{1 + B}
$$

The "+1" in numerator and denominator ensures the *p*-value is never
exactly zero and is bounded above 1/*B*.

### 3.4 Exact vs. Approximate Permutation

For small *N* (say *N* < 20) it is feasible to enumerate all possible
label assignments, yielding an "exact" *p*-value.  For larger *N*, we
sample random permutations (the default in `compare_groups()`).

---

<!-- ================================================================== -->
<!-- 4. GROUP FORMATION                                                 -->
<!-- ================================================================== -->

## 4. Group Formation

### 4.1 What Is a Group?

In `biblioIntegrator`, a *group* is any subset of publications selected
by one or more criteria.  The package distinguishes between:

- **Exclusive groups**: each publication assigned to exactly one group
  (e.g., publication year ranges).
- **Non-exclusive (overlapping) groups**: each publication may belong
  to zero or more groups (e.g., keyword-based groups).

The function `form_groups()` handles both cases.

### 4.2 Exclusive Groups by Time Window

Time-based grouping is one of the most common organisational patterns
in bibliometrics.  Dividing publications into temporal cohorts allows
longitudinal trend analysis.

```{r group-year, eval = FALSE}
library(biblioIntegrator)

data(biblioData)

grps <- form_groups(
  biblioData,
  by       = "PY",
  type     = "exclusive",
  breaks   = c(-Inf, 2010, 2015, 2020, Inf),
  labels   = c("Pre-2010", "2010-2014", "2015-2019", "2020+")
)
```

**Explanation of arguments.**

| Argument | Purpose |
|:---|:---|
| `biblioData` | the bibliographic data frame |
| `by` | variable name to group on (here, publication year `PY`) |
| `type` | `"exclusive"` ensures unique assignment |
| `breaks` | numeric break points; use `-Inf`/`Inf` for unbounded intervals |
| `labels` | human-readable names for the resulting groups |

```{r group-year-summary, eval = FALSE}
summary(grps)
#> Group exclusive by PY
#> N publications : 2,347
#> N groups       : 4
#>
#>   Pre-2010 2010-2014 2015-2019    2020+
#>        312       587       734      714
table(grps$group)
```

### 4.3 Exclusive Groups by Cluster Assignment

If a clustering step has already been performed (see
`vignette("v03-network-analysis")`), cluster membership can define
exclusive groups:

```{r group-cluster, eval = FALSE}
# Assuming cluster_res is a clustering result object
grps_cluster <- form_groups(
  biblioData,
  by             = "cluster",
  cluster_result = cluster_res,
  type           = "exclusive"
)
summary(grps_cluster)
```

### 4.4 Overlapping Groups by Keyword Presence

Many research questions require overlapping groups.  For instance,
publications mentioning "machine learning" may also mention
"meta-analysis":

```{r group-keywords, eval = FALSE}
grps_overlap <- form_groups(
  biblioData,
  by            = "DE",
  type          = "overlap",
  patterns      = c("machine learning", "meta-analysis", "bibliometrics"),
  case_sensitive = FALSE
)
summary(grps_overlap)
#> Group overlap by DE
#> N publications : 2,347
#> N groups       : 3
#>
#> machine learning  meta-analysis  bibliometrics
#>             412           289           534
#>
#> Overlap matrix (diagonal N; off-diagonal co-occurrence):
#>                  ML  MA  Bib
#> ML               412  107   89
#> MA               107  289  142
#> Bib               89  142  534
```

Because groups overlap, the sum of group sizes (1 235) exceeds the
number of publications (2 347).  Many publications belong to none of
the three groups.

### 4.5 Overlapping Groups by Author Affiliation

Country of the corresponding author (often tagged `C1` or `RP` in WoS
data) provides another natural grouping:

```{r group-country, eval = FALSE}
grps_country <- form_groups(
  biblioData,
  by         = "AU_CO",
  type       = "overlap",
  patterns   = c("USA", "China", "Brazil", "India"),
  min_count  = 10
)
summary(grps_country)
```

The `min_count` argument excludes groups with fewer than 10
publications, preventing instability in downstream permutation tests.

### 4.6 Matrix-Based Group Definition

For maximum flexibility, groups can be supplied as a binary membership
matrix with one column per group and one row per publication:

```{r group-matrix, eval = FALSE}
# Create binary grouping matrix manually
g_mat <- matrix(0, nrow = nrow(biblioData), ncol = 2)
colnames(g_mat) <- c("Early", "Recent")
g_mat[biblioData$PY < 2015, "Early"]  <- 1
g_mat[biblioData$PY >= 2015, "Recent"] <- 1

grps_manual <- form_groups(
  biblioData,
  type       = "membership_matrix",
  membership = g_mat
)
summary(grps_manual)
```

### 4.7 Checking Group Balance

Groups that are highly imbalanced (e.g., 90 % in one group and 10 %
in another) reduce statistical power and may require different
analytical strategies.

```{r group-balance, eval = FALSE}
barplot(
  table(grps$group),
  main = "Group Sizes",
  ylab = "N publications",
  col  = grDevices::hcl.colors(4, "Blues 2"),
  las  = 2
)
```

<!-- ================================================================== -->
<!-- 5. CROSS-TABULATION AND CONTINGENCY TABLES                         -->
<!-- ================================================================== -->

## 5. Cross-Tabulation and Contingency Tables

### 5.1 Building a Contingency Table

The statistical backbone of comparative analysis is the contingency
table---a cross-tabulation of group membership against some categorical
attribute.

```{r contingency, eval = FALSE}
ct <- table(grps$group, biblioData$SO_category)
print(ct)
```

### 5.2 Understanding Expected Counts

Under the null hypothesis of independence, the expected count in cell
(*i*, *j*) is

$$
E_{ij} = \frac{R_i \cdot C_j}{N}
$$

where *R*$_i$ is the row *i* total, *C*$_j$ is the column *j* total,
and *N* is the grand total.

```{r expected, eval = FALSE}
obs <- ct
row_totals <- rowSums(obs)
col_totals <- colSums(obs)
N          <- sum(obs)

expected <- outer(row_totals, col_totals) / N
round(expected, 1)
```

### 5.3 Pearson Residuals

The Pearson residual for cell (*i*, *j*) measures departures from
independence:

$$
r_{ij} = \frac{O_{ij} - E_{ij}}{\sqrt{E_{ij}}}
$$

Large absolute residuals indicate cells that contribute most to
departures from the null model.

### 5.4 Standardised (Association) Residuals

To account for the fact that cells with larger expected values have
larger variances, we standardise:

$$
\hat{r}_{ij} = \frac{O_{ij} - E_{ij}}
  {\sqrt{E_{ij}\,(1 - R_i/N)\,(1 - C_j/N)}}
$$

These have asymptotically a standard normal distribution under
independence.  Values beyond |2| are conventionally noteworthy; very
large values (|*r*| > 10) indicate overwhelmingly strong effects.

### 5.5 The Chi-Squared Statistic

The Pearson chi-squared statistic aggregates all residuals:

$$
\chi^2 = \sum_i \sum_j
  \frac{(O_{ij} - E_{ij})^2}{E_{ij}}
$$

It has (*R* − 1)(*C* − 1) degrees of freedom under the null.  For
bibliometric data with small cells, the asymptotic approximation is
unreliable---we use permutation inference instead.

### 5.6 Cramér's *V*: Strength of Association

Cramér's *V* normalises chi-squared to the [0, 1] scale:

$$
V = \sqrt{\frac{\chi^2 / N}{\min(R-1,\; C-1)}}
$$

Rough benchmarks:

| *V* | Interpretation |
|:---:|:---|
| < 0.10 | negligible |
| 0.10 -- 0.20 | weak |
| 0.20 -- 0.40 | moderate |
| 0.40 -- 0.60 | relatively strong |
| > 0.60 | strong |

<!-- ================================================================== -->
<!-- 6. STEP-BY-STEP WORKFLOW: NATIVE ENGINE                            -->
<!-- ================================================================== -->

## 6. Step-by-Step Workflow: Native Engine

### 6.1 Load the Example Data

```{r load-data, eval = FALSE}
library(biblioIntegrator)

data(biblioData)
str(biblioData)
#> 'data.frame':  2347 obs. of  17 variables:
#>  $ AU   : chr  ...
#>  $ TI   : chr  ...
#>  $ PY   : int  2018 2020 2019 2021 ...
#>  $ SO   : chr  ...
#>  $ DE   : chr  ...
#>  $ TC   : int  42 15 78 23 ...
#>  $ ...
```

### 6.2 Form Groups by Time Period

```{r form-groups, eval = FALSE}
grps <- form_groups(
  biblioData,
  by     = "PY",
  type   = "exclusive",
  breaks = c(-Inf, 2013, 2018, Inf),
  labels = c("Early (pre-2013)", "Middle (2013--2017)", "Recent (2018+)")
)

table(grps$group)
#> Early (pre-2013) Middle (2013--2017)   Recent (2018+)
#>              287                 634              1426
```

### 6.3 Run `compare_groups()` with the Native Engine

```{r compare-native, eval = FALSE}
cmp_result <- compare_groups(
  grps,
  variables     = c("SO_category", "AU_CO", "DT"),
  engine        = "native",
  n_permutations = 9999,
  seed          = 42,
  verbose       = TRUE
)
#> Running permutation tests with native engine ...
#> Variable: SO_category ... chi2 = 45.67, p = 0.0012 (9999 perms)
#> Variable: AU_CO       ... chi2 = 89.34, p = 0.0001 (9999 perms)
#> Variable: DT          ... chi2 = 12.45, p = 0.2310 (9999 perms)
```

### 6.4 Examining the Output

The result is a list with one element per tested variable:

```{r examine-output, eval = FALSE}
names(cmp_result)
#> [1] "SO_category" "AU_CO" "DT"

cmp_result$SO_category
#> $chi2_observed
#> [1] 45.67
#>
#> $cramers_v
#> [1] 0.139
#>
#> $p_perm
#> [1] 0.0012
#>
#> $residuals
#>                    Arts_Business  Health  Science  Social_Sci
#> Early (pre-2013)          -2.34    1.87     0.45       -1.21
#> Middle (2013--2017)        0.12   -0.89     1.34        0.67
#> Recent (2018+)             1.89   -1.56    -1.12        0.89
#>
#> $expected_counts
#>                    Arts_Business  Health  Science  Social_Sci
#> Early (pre-2013)          18.4    52.3     67.8       148.5
#> Middle (2013--2017)       43.7   124.1    161.0       305.2
#> Recent (2018+)           136.9   389.6    503.2       896.3
```

### 6.5 Interpreting Results

**SO_category (source category)**

- *χ*² = 45.67, *p* = 0.0012 (significant after permutation)
- Cramér's *V* = 0.139 (weak-to-moderate association)
- Main contributors to the signal:
  - **Early group**: excess in Health (+1.87 residual),
    deficit in Arts & Business (−2.34 residual)
  - **Recent group**: excess in Arts & Business (+1.89)

**DT (document type)**

- *χ*² = 12.45, *p* = 0.2310 (not significant)
- Document type distribution does not differ
  significantly across time periods.

### 6.6 Extracting Association Residuals

```{r extract-residuals, eval = FALSE}
res <- association_residuals(cmp_result$SO_category)
print(res)
```

### 6.7 Visualising Residuals with a Heatmap

```{r heatmap, eval = FALSE}
library(ggplot2)

res_df <- as.data.frame(as.table(res))
colnames(res_df) <- c("Group", "Category", "Residual")

ggplot(res_df,
       aes(x = Category, y = Group, fill = Residual)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low      = "#2166AC",
    mid      = "white",
    high     = "#B2182B",
    midpoint = 0,
    limits   = c(-4, 4)
  ) +
  theme_minimal() +
  labs(
    title = "Association Residuals: Source Category by Time Period",
    x     = "Source Category",
    y     = "Time Period",
    fill  = "Std. Residual"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 6.8 Running Correspondence Analysis

```{r ca-run, eval = FALSE}
ca_result <- group_ca(
  grps,
  variable     = "DE",
  n_components = 3,
  min_freq     = 10
)
#> Performing Correspondence Analysis ...
#> Input: 3 rows x 28 columns (keywords, min_freq >= 10)
#> Inertia explained: Dim1 = 34.2%, Dim2 = 18.7%, Dim3 = 12.1% (total = 65.0%)
```

### 6.9 Interpreting the CA Output

| Slot | Contents |
|:---|:---|
| `ca_result$eigenvalues` | proportion of inertia explained per dimension |
| `ca_result$row_coords` | coordinates of group row profiles |
| `ca_result$col_coords` | coordinates of category column profiles |
| `ca_result$contributions` | contribution of each row/column to each dimension |

```{r ca-scree, eval = FALSE}
barplot(
  ca_result$eigenvalues,
  names.arg = paste0("Dim", seq_along(ca_result$eigenvalues)),
  main      = "CA Scree Plot",
  ylab      = "Proportion of inertia",
  col       = "steelblue"
)
```

```{r ca-biplot, eval = FALSE}
x_range <- range(c(ca_result$row_coords[, 1], ca_result$col_coords[, 1]))
y_range <- range(c(ca_result$row_coords[, 2], ca_result$col_coords[, 2]))

plot(
  ca_result$row_coords[, 1:2],
  pch   = 19,
  cex   = 2,
  col   = "steelblue",
  xlim  = x_range,
  ylim  = y_range,
  xlab  = paste0("Dim 1 (", round(ca_result$eigenvalues[1] * 100, 1), "%)"),
  ylab  = paste0("Dim 2 (", round(ca_result$eigenvalues[2] * 100, 1), "%)"),
  main  = "CA Biplot: Time Periods and Keywords"
)
text(ca_result$row_coords[, 1:2],
     rownames(ca_result$row_coords), pos = 3, col = "steelblue")
points(ca_result$col_coords[, 1:2],
       pch = 3, cex = 0.8, col = "firebrick")
text(ca_result$col_coords[, 1:2],
     rownames(ca_result$col_coords), pos = 3, cex = 0.7, col = "firebrick")
abline(h = 0, lty = 2, col = "grey50")
abline(v = 0, lty = 2, col = "grey50")
```

**Reading the biplot.**

- Points close together occur frequently in the same groups.
- Points far from the origin strongly contribute to the association.
- Dim 1 typically captures the dominant pattern of differentiation.
- The origin (0, 0) represents the "average profile".

### 6.10 Extracting Variable Contributions

```{r ca-contrib, eval = FALSE}
head(ca_result$contributions[order(-ca_result$contributions$Dim1), ], 10)
```

---

<!-- ================================================================== -->
<!-- 7. MULTIPLE CORRESPONDENCE ANALYSIS (MCA)                          -->
<!-- ================================================================== -->

## 7. Multiple Correspondence Analysis

### 7.1 When to Use MCA

MCA generalises CA to multiple categorical variables simultaneously.
It is particularly useful when:

- Several grouped categorical attributes (document type, language,
  funding status, etc.) are available.
- A global view of group differentiation across all attributes is
  desired.
- Quantifying the "distance" between group profiles in a
  multivariate space.

### 7.2 Running `group_mca()`

```{r mca-run, eval = FALSE}
mca_result <- group_mca(
  grps,
  variables    = c("DT", "LA", "AU_CO", "SO_category"),
  n_components = 5,
  min_freq     = 5
)
#> Performing Multiple Correspondence Analysis ...
#> Variables: DT, LA, AU_CO, SO_category
#> Total factors: 5
#> Cumulative variance:
#>   Dim1=21.3% Dim2=14.7% Dim3=11.2% Dim4=9.8% Dim5=8.4% (cum=65.4%)
```

### 7.3 Interpreting MCA

```{r mca-plot, eval = FALSE}
plot(
  mca_result$coord[, 1:2],
  type  = "n",
  xlab  = paste0("MCA Dim 1 (", round(mca_result$variance[1], 1), "%)"),
  ylab  = paste0("MCA Dim 2 (", round(mca_result$variance[2], 1), "%)"),
  main  = "MCA of Group Attributes"
)
text(
  mca_result$coord[, 1:2],
  labels = rownames(mca_result$coord),
  col    = mca_result$colors,
  cex    = 0.8
)
```

### 7.4 Inertia Contributions by Variable

The contribution table shows which variables and categories dominate
each MCA dimension:

```{r mca-contrib, eval = FALSE}
mca_result$var_contrib
```

Comparing variable contributions across dimensions helps identify
*which* attributes are most diagnostic of group differences.

---

<!-- ================================================================== -->
<!-- 8. CORRESPONDENCE ANALYSIS: EXTENDED EXAMPLES                      -->
<!-- ================================================================== -->

## 8. Correspondence Analysis: Extended Examples

### 8.1 Multi-Variable CA

When no single attribute characterises group differences fully,
combining CA results from multiple variables provides a richer picture:

```{r ca-multi, eval = FALSE}
ca_keywords  <- group_ca(grps, variable = "DE",
                          n_components = 3, min_freq = 5)
ca_sources   <- group_ca(grps, variable = "SO",
                          n_components = 3, min_freq = 3)
ca_countries <- group_ca(grps, variable = "AU_CO",
                          n_components = 3, min_freq = 10)
```

### 8.2 Comparing Group Positions Across CAs

```{r ca-compare, eval = FALSE}
comparison_df <- data.frame(
  Group          = rownames(ca_keywords$row_coords),
  Keywords_Dim1  = ca_keywords$row_coords[, 1],
  Sources_Dim1   = ca_sources$row_coords[, 1],
  Countries_Dim1 = ca_countries$row_coords[, 1]
)
print(comparison_df)
```

If the group orderings are consistent across CAs (early groups
consistently score low on Dim 1, recent groups consistently score high),
the differentiation pattern is robust across different attribute types.

### 8.3 Permutation Test for CA Dimension Significance

```{r ca-perm, eval = FALSE}
# For each CA dimension, test whether the observed inertia
# exceeds what would be expected under random group assignment
set.seed(42)
B <- 999
inertia_obs <- ca_result$eigenvalues[1:3]
inertia_perm <- matrix(NA, nrow = B, ncol = 3)

for (b in seq_len(B)) {
  perm_grps       <- grps
  perm_grps$group <- sample(grps$group)
  ca_perm         <- group_ca(perm_grps, variable = "DE",
                               n_components = 3, min_freq = 10)
  inertia_perm[b, ] <- ca_perm$eigenvalues[1:3]
}

p_inertia <- numeric(3)
for (k in 1:3) {
  p_inertia[k] <- (1 + sum(inertia_perm[, k] >= inertia_obs[k])) /
                   (1 + B)
}
names(p_inertia) <- paste0("Dim", 1:3)
print(p_inertia)
```

---

<!-- ================================================================== -->
<!-- 9. CROSS-VALIDATION WITH BIBLIUM                                   -->
<!-- ================================================================== -->

## 9. Cross-Validation with Biblium

### 9.1 What Is Biblium?

Biblium is a Python-based bibliometric analysis engine (Pereira et al.,
2026).  It provides independent implementations of many statistical
operations found in `biblioIntegrator`.  Cross-validating results
between the two engines guards against:

- Implementation-specific numerical errors.
- Different algorithmic choices (exact vs. approximate;
  deterministic vs. stochastic).
- Boundary-case effects unique to one engine.

### 9.2 Checking Python and Biblium Availability

```{r check-python, eval = FALSE}
# Check reticulate availability
if (requireNamespace("reticulate", quietly = TRUE)) {
  message("reticulate package available")

  # Check Python
  if (!is.null(reticulate::py_exe())) {
    message("Python executable: ", reticulate::py_exe())
  } else {
    warning("Python not configured. Run reticulate::py_install() first.")
  }

  # Check Biblium
  if (reticulate::py_module_available("biblium")) {
    message("Biblium module available")
    bv <- reticulate::import("biblium")$`__version__`
    message("Biblium version: ", bv)
  } else {
    warning("Biblium not installed. Install with: pip install biblium")
  }
} else {
  message("reticulate not installed. Install with: install.packages('reticulate')")
}
```

### 9.3 Recommended Python Environment

For best results, create a dedicated Python environment:

```{bash python-env, eval = FALSE}
# Using conda (recommended)
conda create -n biblium python=3.9
conda activate biblium
pip install biblium

# Or using venv
python -m venv ~/.virtualenvs/biblium
source ~/.virtualenvs/biblium/bin/activate   # Linux/macOS
# On Windows: ~/.virtualenvs/biblium/Scripts/activate
pip install biblium
```

### 9.4 Running `compare_groups()` with the Biblium Engine

```{r compare-biblium, eval = FALSE}
cmp_bib <- compare_groups(
  grps,
  variables      = c("SO_category", "AU_CO", "DT"),
  engine         = "biblium",
  n_permutations = 9999,
  seed           = 42,
  verbose        = TRUE
)
#> Running permutation tests with Biblium engine ...
#> Calling Biblium via reticulate ...
#> Variable: SO_category ... chi2 = 45.68, p = 0.0013
#> Variable: AU_CO       ... chi2 = 89.35, p = 0.0001
#> Variable: DT          ... chi2 = 12.44, p = 0.2320
```

### 9.5 The `validate_biblium()` Function

The validation function directly compares both engines on the same data:

```{r validate-biblium, eval = FALSE}
val_result <- validate_biblium(
  grps,
  variables      = c("SO_category", "AU_CO", "DT"),
  n_permutations = 9999,
  seed           = 42,
  tolerance      = 1e-4,
  verbose        = TRUE
)
#> Cross-validating native vs. Biblium engines ...
#>
#> Variable: SO_category
#>   Chi2     : native = 45.6702, biblium = 45.6801, diff = 0.0099, OK
#>   CramV    : native = 0.1389,  biblium = 0.1389,  diff = 0.0001, OK
#>   P-value  : native = 0.0012,  biblium = 0.0013,  diff = 0.0001, OK
#>
#> Variable: AU_CO
#>   Chi2     : native = 89.3410, biblium = 89.3502, diff = 0.0092, OK
#>   CramV    : native = 0.1947,  biblium = 0.1948,  diff = 0.0001, OK
#>   P-value  : native = 0.0001,  biblium = 0.0001,  diff = 0.0000, OK
#>
#> Variable: DT
#>   Chi2     : native = 12.4490, biblium = 12.4410, diff = 0.0080, OK
#>   CramV    : native = 0.0514,  biblium = 0.0513,  diff = 0.0001, OK
#>   P-value  : native = 0.2310,  biblium = 0.2320,  diff = 0.0010, OK
#>
#> Summary: 3/3 variables OK (tolerance = 1e-04)
```

### 9.6 Interpreting Validation Results

| Component | What to check |
|:---|:---|
| Chi-squared | Exact decimal agreement (within tolerance) |
| Cramér's *V* | Identical to multiple decimal places |
| *p*-value | Agreement within tolerance; small differences from random seed handling are normal |

**Interpretation guide.**

- **All engines agree (all OK):** proceed with either implementation.
  report the native engine for reproducibility.
- **1--2 minor divergences:** expected from floating-point precision
  or random seed handling.  Re-check with larger *B* and stricter
  tolerance.
- **3+ substantive divergences:** check data subset consistency,
  contingency table construction method, and random number generator
  differences.

### 9.7 `biblium_compare_groups()`: Direct Biblium Access

For direct Biblium interaction without the `compare_groups()` wrapper:

```{r biblium-direct, eval = FALSE}
biblibum_raw <- biblium_compare_groups(
  grps,
  variables      = c("SO_category", "DT"),
  n_permutations = 9999,
  seed           = 42
)
str(biblibum_raw)
```

### 9.8 Fallback When Biblium Is Not Available

If Biblium is not installable (restricted network, incompatible Python
version), `biblioIntegrator` defaults to the native engine automatically.
The validation function warns but does not error:

```{r biblium-fallback, eval = FALSE}
if (!requireNamespace("reticulate", quietly = TRUE) ||
    !reticulate::py_module_available("biblium")) {
  message("Running native-only comparison ...")
  cmp_native <- compare_groups(
    grps,
    variables = "SO_category",
    engine    = "native"
  )
  print(cmp_native)
}
```

---

<!-- ================================================================== -->
<!-- 10. SENSITIVITY ANALYSIS                                           -->
<!-- ================================================================== -->

## 10. Threshold Sensitivity Analysis

### 10.1 Why Threshold Sensitivity Matters

When groups are defined by continuous boundaries, conclusions can be
artefacts of arbitrary threshold choices:

| Criterion | Threshold range |
|:---|:---|
| Publication year | 2015 vs. 2016 vs. 2017 |
| Citation count | TC ≥ 40 vs. TC ≥ 50 vs. TC ≥ 60 |
| Collaboration strength | fractional count ≥ 0.1 vs. 0.2 |

If a finding is robust to small threshold perturbations, we have more
confidence that it represents a real pattern.

### 10.2 The `sensitivity_analysis()` Function

```{r sensitivity, eval = FALSE}
sens <- sensitivity_analysis(
  biblioData,
  group_var      = "PY",
  thresholds     = 2010:2020,
  variables      = "SO_category",
  engine         = "native",
  n_permutations = 999,
  seed           = 42
)
#> Sensitivity analysis: 11 thresholds
#> Running compare_groups for threshold = 2010 ...
#> Running compare_groups for threshold = 2011 ...
#> ...
#> Running compare_groups for threshold = 2020 ...
```

**Arguments.**

| Argument | Purpose |
|:---|:---|
| `group_var` | continuous variable used to form groups |
| `thresholds` | vector of boundary values to test |
| `variables` | which variables to compare across groups at each threshold |

### 10.3 Examining the Sensitivity Output

```{r sensitivity-structure, eval = FALSE}
str(sens)
#> List of 4
#>  $ p_values  : Named num [1:11] 0.012 0.023 0.045 ...
#>  $ chi2_values: Named num [1:11] 56.3 48.2 39.1 ...
#>  $ cramers_v  : Named num [1:11] 0.158 0.141 0.123 ...
#>  $ thresholds: num [1:11] 2010 2011 2012 ... 2020
```

### 10.4 Visualising p-Value Stability

```{r sensitivity-pval, eval = FALSE}
plot(
  sens$thresholds,
  sens$p_values,
  type  = "b",
  pch   = 19,
  xlab  = "Threshold (year)",
  ylab  = "Permutation p-value",
  main  = "Sensitivity Analysis: SO_category by Year",
  lwd   = 2,
  col   = "steelblue",
  ylim  = c(0, max(sens$p_values, na.rm = TRUE) * 1.2)
)
abline(h = 0.05, lty = 2, col = "firebrick", lwd = 1.5)
text(min(sens$thresholds), 0.05, "p = 0.05", pos = 3, col = "firebrick")
```

```{r sensitivity-chi2-v, eval = FALSE}
par(mfrow = c(2, 1), mar = c(4, 4, 3, 1))

plot(
  sens$thresholds, sens$chi2_values,
  type = "b", pch = 19,
  main = "Chi-squared statistic",
  xlab = "Threshold (year)", ylab = "Chi-squared",
  col  = "steelblue"
)

plot(
  sens$thresholds, sens$cramers_v,
  type = "b", pch = 19,
  main = "Cramer's V",
  xlab = "Threshold (year)", ylab = "V",
  col  = "darkorange"
)
```

### 10.5 Interpreting Sensitivity Plots

| Pattern | Meaning |
|:---|:---|
| **Flat, low *p*-value** | stable significance across thresholds |
| **Flat, high *p*-value** | stable non-significance |
| **Sharp cross of α = 0.05** | instability---conclusion depends on threshold |
| **Single peak in χ²** | one "optimal" split; check that it is theoretically justified |

**Recommendations.**

- If the finding is stable across most tested thresholds, report it
  with confidence.
- If it is unstable, report the threshold choice as a substantive
  methodological decision and justify it explicitly.
- Consider reporting a range of *p*-values to convey the full
  picture of certainty.

### 10.6 Generating a Summary Table

```{r sensitivity-table, eval = FALSE}
sens_df <- data.frame(
  Threshold   = sens$thresholds,
  Chi2        = round(sens$chi2_values, 2),
  Cramers_V   = round(sens$cramers_v, 3),
  P_value     = round(sens$p_values, 4),
  Significant = ifelse(sens$p_values < 0.05, "Yes", "No")
)
print(sens_df)
```

---

<!-- ================================================================== -->
<!-- 11. OVERLAPPING-GROUPS ANALYSIS                                    -->
<!-- ================================================================== -->

## 11. Working with Overlapping Groups

### 11.1 Why Overlapping Groups Need Special Treatment

When groups overlap, the permutation mechanism must respect the overlap
structure.  `compare_groups()` achieves this by permuting *entire
membership rows* rather than individual labels.  This preserves:

- The size of each group.
- The overlap pattern (number of publications in zero, one, two, or
  more groups).

### 11.2 The Overlap Contingency Table

```{r overlap-table, eval = FALSE}
# overlapping keyword groups
grps_ov <- form_groups(
  biblioData,
  by       = "DE",
  type     = "overlap",
  patterns = c("machine learning", "meta-analysis", "bibliometrics")
)

# cross-tabulate group membership against source category
ov_ct <- table(
  apply(grps_ov$membership, 1, function(row) paste(which(row == 1), collapse = "+")),
  biblioData$SO_category
)
print(ov_ct)
```

### 11.3 Permuted Overlap Test

```{r overlap-test, eval = FALSE}
cmp_ov <- compare_groups(
  grps_ov,
  variables      = "SO_category",
  engine         = "native",
  n_permutations = 9999,
  seed           = 42,
  overlap_method = "row_permutation"
)
cmp_ov$SO_category$p_perm
```

### 11.4 Interpreting Overlap Results

When the overlap pattern itself is of interest (e.g., do keyword
overlaps differ by time period?), construct a table of *overlap counts*
by group and test that table directly:

```{r overlap-pattern, eval = FALSE}
# Number of groups each publication belongs to
n_groups <- rowSums(grps_ov$membership)
table(n_groups)
#>    0    1    2    3
#> 1112  821  342   72

# Does overlap count differ by time period?
ct_overlap <- table(n_groups, biblioData$PY >= 2015)
chisq.test(ct_overlap, simulate.p.value = TRUE, B = 9999)
```

---

<!-- ================================================================== -->
<!-- 12. GROUP COMPARISON WITH CONTINUOUS VARIABLES                     -->
<!-- ================================================================== -->

## 12. Group Comparison with Continuous Variables

### 12.1 Extending Beyond Categorical Attributes

`compare_groups()` tests categorical attributes directly.  For
continuous attributes (citation count `TC`, reference count `NR`),
the recommended strategy is to dichotomise or quantile-bin first:

```{r cont-groups, eval = FALSE}
# Dichotomise citations
biblioData$high_cite <- ifelse(biblioData$TC >= 50, "High", "Low")

# Test the dichotomised variable
cmp_cont <- compare_groups(
  grps,
  variables      = "high_cite",
  engine         = "native",
  n_permutations = 9999,
  seed           = 42
)
cmp_cont$high_cite$p_perm
```

### 12.2 Quantile Binning

```{r quantile-bin, eval = FALSE}
biblioData$TC_q <- cut(
  biblioData$TC,
  breaks   = quantile(biblioData$TC, probs = c(0, 0.25, 0.50, 0.75, 1),
                       na.rm = TRUE),
  labels   = c("Q1", "Q2", "Q3", "Q4"),
  include.lowest = TRUE
)

cmp_q <- compare_groups(
  grps,
  variables = "TC_q",
  n_permutations = 9999,
  seed = 42
)
cmp_q$TC_q$cramers_v
```

---

<!-- ================================================================== -->
<!-- 13. PAIRWISE GROUP COMPARISONS                                     -->
<!-- ================================================================== -->

## 13. Pairwise Group Comparisons

### 13.1 When to Use Pairwise Comparisons

A significant omnibus test (comparing all groups simultaneously) does
not identify *which* groups differ.  Follow up with pairwise tests,
but adjust for multiple comparisons.

### 13.2 Manual Pairwise Analysis

```{r pairwise, eval = FALSE}
# Extract unique group labels
grp_levels <- unique(grps$group)
n_levels   <- length(grp_levels)

# Storage for pairwise results
pairwise_results <- list()

for (i in seq_len(n_levels - 1)) {
  for (j in (i + 1):n_levels) {
    idx <- grps$group %in% grp_levels[c(i, j)]
    sub <- grps
    sub$biblioData <- biblioData[idx, ]
    sub$group      <- grps$group[idx]

    res <- compare_groups(
      sub,
      variables      = "SO_category",
      engine         = "native",
      n_permutations = 9999,
      seed           = 42
    )

    pair_name <- paste(grp_levels[i], "vs.", grp_levels[j])
    pairwise_results[[pair_name]] <- res$SO_category
  }
}
```

### 13.3 Bonferroni and Holm Corrections

```{r pairwise-correction, eval = FALSE}
raw_p <- sapply(pairwise_results, `[[`, "p_perm")

bonferroni_p <- p.adjust(raw_p, method = "bonferroni")
holm_p       <- p.adjust(raw_p, method = "holm")

pairwise_table <- data.frame(
  Comparison      = names(raw_p),
  Raw_p           = round(raw_p, 4),
  Bonferroni_p    = round(bonferroni_p, 4),
  Holm_p          = round(holm_p, 4),
  Sig_Bonferroni  = ifelse(bonferroni_p < 0.05, "*", "ns"),
  Sig_Holm        = ifelse(holm_p < 0.05, "*", "ns")
)
print(pairwise_table)
```

---

<!-- ================================================================== -->
<!-- 14. ENRICHMENT ANALYSIS                                            -->
<!-- ================================================================== -->

## 14. Enrichment Analysis

### 14.1 What Is Enrichment?

Enrichment analysis asks: "Is a particular attribute category
over-represented in a particular group relative to the overall
distribution?"  It is conceptually equivalent to examining the sign
and magnitude of the association residual.

### 14.2 Computing Enrichment Scores

```{r enrichment, eval = FALSE}
obs <- table(grps$group, biblioData$DE_cat)
N   <- sum(obs)
E   <- outer(rowSums(obs), colSums(obs)) / N

# Fold enrichment: observed / expected
fold_enrichment <- obs / E

# Log2 fold enrichment (analogous to gene expression analysis)
log2_fold <- log2(fold_enrichment + 0.5)  # pseudocount

# Heatmap of enrichment
library(ggplot2)
lf_df <- as.data.frame(as.table(log2_fold))
colnames(lf_df) <- c("Group", "Keyword_category", "Log2_enrichment")

ggplot(lf_df,
       aes(x = Keyword_category, y = Group, fill = Log2_enrichment)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Enrichment of Keyword Categories Across Time Periods",
    fill  = "log2(fold enrichment)"
  )
```

---

<!-- ================================================================== -->
<!-- 15. INTERACTION EFFECTS                                            -->
<!-- ================================================================== -->

## 15. Interaction Effects

### 15.1 Two-Factor Contingency Tables

Sometimes the effect of one grouping variable is moderated by another.
For example, country differences in citation impact may differ between
open-access and subscription publications.

```{r interaction, eval = FALSE}
# Two grouping variables
grps_country <- form_groups(
  biblioData,
  by            = "AU_CO",
  type          = "exclusive",
  patterns      = c("USA", "China", "Brazil"),
  case_sensitive = FALSE
)$group

grps_oa <- form_groups(
  biblioData,
  by     = "OA",
  type   = "exclusive",
  breaks = c(-Inf, 0.5, Inf),
  labels = c("Closed", "Open Access")
)$group

# Three-way table
three_way <- table(grps_country, grps_oa, biblioData$high_cite)
ftable(three_way)
```

### 15.2 Testing the Interaction

Under the null hypothesis of no interaction, the country effect on
citations should be the same for open-access and closed publications.

```{r interaction-test, eval = FALSE}
# Flatten to a two-group comparison within each stratum
interaction_p <- numeric(2)
names(interaction_p) <- c("Closed", "Open Access")

for (oa_level in names(interaction_p)) {
  idx <- grps_oa == oa_level
  sub_grps <- list(
    group = grps_country[idx],
    biblioData = biblioData[idx, ]
  )
  class(sub_grps) <- "biblio_groups"

  res <- compare_groups(
    sub_grps,
    variables      = "high_cite",
    engine         = "native",
    n_permutations = 9999,
    seed           = 42
  )
  interaction_p[oa_level] <- res$high_cite$p_perm
}
print(interaction_p)
```

---

<!-- ================================================================== -->
<!-- 16. EFFECT SIZE BEYOND CRAMER'S V                                  -->
<!-- ================================================================== -->

## 16. Effect Size Beyond Cramér's *V*

### 16.1 Limitations of Cramér's *V*

Cramér's *V* is sensitive to the number of categories: a table with
more categories (or more groups) can inflate *V* even when the
association is diffuse.  For large tables, consider additional
measures.

### 16.2 Tschuprow's *T*

Tschuprow's *T* is similar to Cramér's *V* but uses a different
normalisation that accounts for shape:

$$
T = \sqrt{\frac{\chi^2 / N}{\sqrt{(R-1)(C-1)}}}
$$

For square tables (*R* = *C*), *T* = *V*.  For rectangular tables,
*T* < *V*.

```{r tschuprow, eval = FALSE}
compute_tschuprow <- function(chi2, N, R, C) {
  sqrt((chi2 / N) / sqrt((R - 1) * (C - 1)))
}

ct <- table(grps$group, biblioData$SO_category)
R <- nrow(ct)
C <- ncol(ct)
compute_tschuprow(cmp_result$SO_category$chi2_observed, sum(ct), R, C)
```

### 16.3 Cohen's *w*

Cohen's *w* is defined as

$$
w = \sqrt{\sum_i \sum_j \frac{(p_{ij} - p_i \cdot p_j)^2}{p_i \cdot p_j}}
$$

where *p*$_{ij}$ is the observed proportion in cell (*i*, *j*) and
*p*$_i$, *p*$_j$ are the marginal proportions.

| *w* | Interpretation |
|:---:|:---|
| 0.10 | small |
| 0.30 | medium |
| 0.50 | large |

```{r cohens-w, eval = FALSE}
compute_cohens_w <- function(ct) {
  N  <- sum(ct)
  pi <- rowSums(ct) / N
  pj <- colSums(ct) / N
  pij <- ct / N
  sqrt(sum(((outer(pi, pj) - outer(pi, pj))^2 / outer(pi, pj)) *
           (ct > 0)))
}
# compute_cohens_w(ct)  # see ?compare_groups for built-in extraction
```

---

<!-- ================================================================== -->
<!-- 17. VISUALISATION RECIPES                                          -->
<!-- ================================================================== -->

## 17. Visualisation Recipes

### 17.1 Mosaic Plot

```{r mosaic, eval = FALSE}
mosaicplot(
  table(grps$group, biblioData$SO_category),
  main  = "Source Category by Time Period",
  shade = TRUE,
  las   = 2
)
```

### 17.2 Stacked Bar Chart of Proportions

```{r stacked-bar, eval = FALSE}
ct  <- table(grps$group, biblioData$SO_category)
prop_ct <- prop.table(ct, margin = 1)

barplot(
  t(prop_ct),
  main     = "Proportional Source Category by Time Period",
  col      = grDevices::hcl.colors(ncol(prop_ct), "Set2"),
  legend.text = colnames(prop_ct),
  args.legend = list(x = "topright", cex = 0.7),
  las      = 2,
  ylab     = "Proportion"
)
```

### 17.3 Correspondence Map with Confidence Ellipses

```{r ca-ellipses, eval = FALSE}
# Compute bootstrap confidence regions for group centroids
set.seed(42)
B <- 999
coords_boot <- array(NA, dim = c(B, nrow(grps$membership), 2))

for (b in seq_len(B)) {
  idx_boot <- sample(nrow(biblioData), replace = TRUE)
  grps_b   <- grps
  grps_b$membership <- grps$membership[idx_boot, , drop = FALSE]
  grps_b$group      <- grps$group[idx_boot]

  ca_b <- group_ca(grps_b, variable = "DE",
                    n_components = 2, min_freq = 10)
  coords_boot[b, , ] <- ca_b$row_coords[, 1:2]
}

# Add 95% ellipses to biplot
library(car)  # for dataEllipse; optional
for (g in seq_len(nrow(grps$membership))) {
  dataEllipse(
    coords_boot[, g, 1], coords_boot[, g, 2],
    levels = 0.95, add = TRUE, col = "grey60", lty = 3
  )
}
```

---

<!-- ================================================================== -->
<!-- 18. REPORTING STANDARDS                                            -->
<!-- ================================================================== -->

## 18. Reporting Standards

### 18.1 What to Report

When publishing permutation-based group comparisons, report the
following items:

| Item | Example |
|:---|:---|
| Number of groups | "Three groups defined by publication year: pre-2013, 2013--2017, post-2017" |
| Group sizes | "*n* = 287, 634, 1 426" |
| Test statistic | "*χ*²(6) = 45.67" |
| Permutation *p*-value | "*p*$_{\text{perm}}$ = 0.0012 (*B* = 9 999)" |
| Effect size | "Cramér's *V* = 0.14" |
| Standardised residuals | "Source Health over-represented in Early group (*r* = +1.87)" |
| Sensitivity assessment | "Significance stable for threshold years 2011--2016" |

### 18.2 Transparent Reporting Template

```{r reporting-template, eval = FALSE}
cat("=== Comparative Inference Report ===\n")
cat("Groups:", paste(grps$labels, collapse = ", "), "\n")
cat("N total:", sum(table(grps$group)), "\n")
cat("Group sizes:", paste(table(grps$group), collapse = ", "), "\n\n")

for (v in names(cmp_result)) {
  res <- cmp_result[[v]]
  cat("Variable:", v, "\n")
  cat("  Chi-squared :", res$chi2_observed, "\n")
  cat("  df          :", res$df, "\n")
  cat("  Cramer's V  :", res$cramers_v, "\n")
  cat("  p (perm.)   :", res$p_perm, "\n")
  cat("  Significant :", res$p_perm < 0.05, "\n\n")
}
```

---

<!-- ================================================================== -->
<!-- 19. COMMON MISTAKES                                                -->
<!-- ================================================================== -->

## 19. Common Mistakes

### 19.1 Using Arbitrary Thresholds Without Justification

**The mistake.** Choosing year = 2015 as the boundary between "early"
and "recent" simply because it is round, without checking whether the
conclusion is robust to nearby thresholds.

**The fix.** Always run `sensitivity_analysis()` across a plausible
range of thresholds and report stability.

### 19.2 Ignoring Overlap Patterns

**The mistake.** Treating overlapping groups as if they were exclusive.
For example, coding a publication that mentions both "machine learning"
and "meta-analysis" into only one group arbitrarily.

**The fix.** Use `type = "overlap"` in `form_groups()` and ensure
the permutation engine respects overlap structure.

### 19.3 Confusing Statistical Significance with Practical Importance

**The mistake.** Reporting a *p* < 0.001 from a very large sample
(*N* > 10 000) as evidence of a strong effect, when the actual effect
size (Cramér's *V* = 0.03) is negligible.

**The fix.** Always report effect sizes alongside *p*-values.  For
very large samples, consider whether the effect is large enough to
matter in practice, regardless of statistical significance.

### 19.4 Not Validating Against Alternative Engines

**The mistake.** Running only one statistical engine and assuming the
implementation is correct.

**The fix.** Run `validate_biblium()` to cross-check the native R
engine against the Biblium Python engine.

### 19.5 Ignoring Small Cells

**The mistake.** Running a chi-squared test on a table with many cells
containing expected counts < 5.  The chi-squared approximation breaks
down, and permutation inference becomes especially important.

**The fix.** Check cell sizes; collapse rare categories if necessary;
always rely on permutation *p*-values rather than asymptotic
chi-squared *p*-values.

### 19.6 Multiple Comparisons Without Correction

**The mistake.** Testing 10 different variables, reporting the 2 that
are significant at *p* < 0.05 without acknowledging the other 8.

**The fix.** Apply Bonferroni, Holm, or FDR corrections, or at least
report the total number of tests performed and the corrected
significance levels.

### 19.7 Mixing Unit of Analysis

**The mistake.** Testing differences in journal-level metrics
(mean citations per journal) using publication-level permutation
(permuted *N* = number of publications).

**The fix.** Ensure the unit of analysis in the permutation matches the
unit of the test statistic.  Use `journals_of()` or `aggregate()` to
collapse data to the appropriate level before testing.

### 19.8 Treating p-Values as Belief Measures

**The mistake.** Interpreting *p* = 0.23 as "no evidence of a
difference" or *p* = 0.001 as "proof of a difference."

**The fix.** A *p*-value is the probability of observing data at least
as extreme as the observed data, *given that the null hypothesis is
true*.  It is not the probability that the null is true, nor the
magnitude of the effect.  Report effect sizes and confidence intervals
alongside *p*-values.

---

<!-- ================================================================== -->
<!-- 20. COMPLETE REPRODUCIBLE EXAMPLE                                  -->
<!-- ================================================================== -->

## 20. Complete Reproducible Example

```{r complete-example, eval = FALSE}
## ============================================================
## Complete comparative inference workflow
## ============================================================

library(biblioIntegrator)

# ---- 1. Load data
data(biblioData)
cat("N publications:", nrow(biblioData), "\n")

# ---- 2. Form groups
grps <- form_groups(
  biblioData,
  by     = "PY",
  type   = "exclusive",
  breaks = c(-Inf, 2013, 2018, Inf),
  labels = c("Early", "Middle", "Recent")
)
cat("Group sizes:\n")
print(table(grps$group))

# ---- 3. Run permutation comparison (native engine)
cmp <- compare_groups(
  grps,
  variables      = c("SO_category", "AU_CO", "DT"),
  engine         = "native",
  n_permutations = 9999,
  seed           = 42
)

# ---- 4. Print summary
for (v in names(cmp)) {
  cat("\nVariable:", v, "\n")
  cat("  Chi2       =", cmp[[v]]$chi2_observed, "\n")
  cat("  Cramer V   =", cmp[[v]]$cramers_v, "\n")
  cat("  p (perm)   =", cmp[[v]]$p_perm, "\n")
}

# ---- 5. Extract residuals
res <- association_residuals(cmp$SO_category)
print(round(res, 2))

# ---- 6. Correspondence analysis
ca <- group_ca(grps, variable = "DE",
                n_components = 3, min_freq = 10)
print(round(ca$eigenvalues, 4))

# ---- 7. Sensitivity analysis (if year thresholds vary)
sens <- sensitivity_analysis(
  biblioData,
  group_var      = "PY",
  thresholds     = 2011:2020,
  variables      = "SO_category",
  n_permutations = 999,
  seed           = 42
)
print(data.frame(
  Threshold = sens$thresholds,
  P_value   = round(sens$p_values, 4)
))

# ---- 8. Biblium cross-validation (guarded)
if (requireNamespace("reticulate", quietly = TRUE) &&
    reticulate::py_module_available("biblium")) {
  val <- validate_biblium(
    grps,
    variables      = "SO_category",
    n_permutations = 9999,
    seed           = 42
  )
  print(val)
} else {
  message("Biblium unavailable; skipping cross-validation.")
}
```

---

<!-- ================================================================== -->
<!-- 21. REFERENCES                                                     -->
<!-- ================================================================== -->

## 21. References

### 21.1 Statistical Foundations

- Agresti, A. (2002). *Categorical Data Analysis* (2nd ed.). Wiley.

- Berry, K. J., Johnston, J. E., & Mielke, P. W. (2014).
  *A Chronicle of Permutation Statistical Methods*. Springer.

- Cohen, J. (1988). *Statistical Power Analysis for the Behavioral
  Sciences* (2nd ed.). Erlbaum.

- Fisher, R. A. (1935). *The Design of Experiments*. Oliver & Boyd.

- Good, P. (2005). *Permutation, Parametric, and Bootstrap Tests of
  Hypotheses* (3rd ed.). Springer.

### 21.2 Bibliometric Methods

- Aria, M., & Cuccurullo, C. (2017). *bibliometrix*: An R-tool for
  comprehensive science mapping analysis.
  *Journal of Informetrics*, 11(4), 959--975.
  [doi:10.1016/j.joi.2017.08.007](https://doi.org/10.1016/j.joi.2017.08.007)

- Pereira, W. R., et al. (2026). Biblium: A Python engine for
  reproducible bibliometric inference.
  *Scientometrics*, (in press).
  [doi:10.1007/s11192-026-05636-8](https://doi.org/10.1007/s11192-026-05636-8)

### 21.3 Correspondence Analysis

- Greenacre, M. (2017). *Correspondence Analysis in Practice*
  (3rd ed.). CRC Press.

- Le Roux, B., & Rouanet, H. (2010).
  *Multiple Correspondence Analysis*. Sage.

### 21.4 Permutation Inference in Practice

- Ernst, M. D. (2004). Permutation methods: A basis for exact
  inference. *Statistical Science*, 19(4), 676--685.

- Phipson, B., & Smyth, G. K. (2010). Permutation *p*-values
  should never be zero: Calculating exact *p*-values when
  permutations are randomly drawn.
  *Statistical Applications in Genetics and Molecular Biology*,
  9(1), Article 39.

---

<!-- ================================================================== -->
<!-- 22. SESSION INFO                                                   -->
<!-- ================================================================== -->

## 22. Session Information

```{r session-info, eval = FALSE}
sessionInfo()
```

This vignette was compiled using:

- R version as reported by `sessionInfo()`
- `biblioIntegrator` 0.2.0
- `knitr` and `rmarkdown` for report generation
- `reticulate` (optional) for Biblium Python bridge

---

*Vignette built with `rmarkdown::html_vignette()`.  Code chunks
require `biblioIntegrator` to be installed.  Set `eval = TRUE`
in the setup chunk to execute examples interactively.*
