---
title: "Networks and stability"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Networks and stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biblioIntegrator)
```

# Purpose

Network construction is distinct from network interpretation. The native engine is always available; biblionetwork can accelerate coauthorship construction.

# Workflow

```{r}
x <- as_biblio_project(example_biblio())
g <- bibliographic_network(x, "coauthor")
network_centrality(g)
network_communities(g)
network_stability(x, B = 20, seed = 4)
if (requireNamespace("biblionetwork", quietly = TRUE)) {
  gb <- bibliographic_network(x, "coauthor", engine = "biblionetwork")
  attr(gb, "engine")
}
```

# Interpretation

Results should be interpreted in relation to database coverage, time window, entity normalization and analytical thresholds. Provenance should be retained whenever data are merged, deduplicated or enriched.
