---
title: "Temporal and Textual Analysis"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temporal and Textual Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dpi = 150,
  message = FALSE,
  warning = FALSE
)

# Safe chunk: load package at setup
library(biblioIntegrator)
```

# 1. Why This Vignette Exists

Bibliometric data contains rich temporal and textual signals that go far beyond citation counts and h-indices. Understanding *when* research emerges and *what* topics it addresses provides a deeper view of scientific evolution.

**Temporal analysis** reveals growth trajectories, citation dynamics, and historically seminal moments. Examining publication patterns over time allows researchers to identify inflection points—moments when a field expanded rapidly, plateaued, or declined. Citation velocity shows whether research impact arrives immediately or accumulates slowly across decades.

**Textual analysis** complements temporal data by revealing *what* is being studied. Term frequency and TF-IDF analysis extract meaningful themes from titles and abstracts. Trending topic detection identifies emerging research fronts before they become saturated.

Together, these methods help answer fundamental questions:

- **When did a research field accelerate?** Growth curves reveal the inflection points where research attention intensified.
- **How quickly are papers cited?** Citation velocity distinguishes fast-moving fields from those with slower maturation.
- **Which years were seminal?** Reference Publication Year Spectroscopy (RPYS) identifies years that exerted disproportionate historical influence.
- **What topics are trending?** Text analysis reveals emerging themes and declining interests.
- **Is research disruptive or consolidating?** Disruption indices quantify whether work opens new directions or consolidates existing ones.

This vignette demonstrates how to combine temporal and textual analyses using the `biblioIntegrator` package.

---

# 2. Learning Objectives

After completing this vignette, you will be able to:

1. **Analyze publication growth patterns** using `temporal_growth()` with linear, exponential, and logistic models
2. **Compute citation velocity** to measure how quickly papers accumulate citations over time
3. **Model citation trajectories** across different time horizons and detect sleeping beauties
4. **Perform Reference Publication Year Spectroscopy (RPYS)** to identify historically influential years
5. **Normalize citations** by field and time window using `normalized_citations()`
6. **Compute disruption indices** to distinguish transformative research from incremental work
7. **Calculate term frequency** from bibliographic text fields and interpret raw counts
8. **Apply TF-IDF weighting** to identify terms that distinguish specific documents or groups
9. **Detect trending topics** using `trend_topics()` with time-windowed analysis
10. **Interpret and synthesize** temporal and textual results for research evaluation

---

# 3. Prerequisites

## 3.1 Required Packages

```{r prerequisites, eval=FALSE}
# Core package
library(biblioIntegrator)

# Data manipulation
library(dplyr)
library(tidyr)

# Visualization
library(ggplot2)
library(scales)

# Text processing (optional but recommended)
if (requireNamespace("tidytext", quietly = TRUE)) {
  library(tidytext)
}

# Nice composite figures (optional)
if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
}
```

## 3.2 Data Structure Expectations

Most temporal and text functions operate on a data frame with these typical columns:

| Column | Type | Description |
|--------|------|-------------|
| `PY` | integer | Publication year |
| `TC` | integer | Total citation count |
| `TI` | character | Title text |
| `AB` | character | Abstract text |
| `AU` | character | Authors (semicolon-separated) |
| `SO` | character | Source / journal name |
| `WC` | character | Web of Science category or field |
| `DE` | character | Author keywords |
| `CR` | character | Cited references (semicolon-separated) |
| `DB` | character | Database source (Scopus, WoS, etc.) |

Not all columns are required by every function. Check each function's documentation.

## 3.3 Loading Example Data

```{r load-example, eval=FALSE}
# Option A: Use built-in example data
data(example_biblio)
biblio_df <- example_biblio()

# Option B: Load your own data
# biblio_df <- biblio_load("my_data.csv")

# Option C: Merge multiple databases
# merged <- biblio_merge(list(scopus_df, wos_df),
#                         source_names = c("Scopus", "WoS"))

# Quick data check
cat("Dimensions:", nrow(biblio_df), "rows x", ncol(biblio_df), "columns\n")
cat("Year range:", range(biblio_df$PY, na.rm = TRUE), "\n")
cat("Missing PY:", sum(is.na(biblio_df$PY)), "\n")
```

---

# 4. Temporal Analysis

## 4.1 Publication Growth: `temporal_growth()`

Publication growth analysis reveals how a field evolves over time. The function `temporal_growth()` counts publications per year and fits growth models.

### 4.1.1 Basic Usage

```{r temporal-growth-basic, eval=FALSE}
# Convert to biblio_project and compute growth
bp <- as_biblio_project(example_biblio())
growth <- temporal_growth(bp)

# View result structure
str(growth)
# 'data.frame':	25 obs. of  3 variables:
#  $ year : int  1999 2000 2001 ...
#  $ count: int  12 25 31 ...
#  $ model: num  ...  (fitted values)

head(growth, 10)
```

### 4.1.2 Visualization

```{r temporal-growth-plot, eval=FALSE}
# Plot annual publication counts
ggplot(growth, aes(x = year, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  labs(
    title = "Publication Growth Over Time",
    x = "Year",
    y = "Number of Publications"
  ) +
  theme_minimal(base_size = 14)
```

### 4.1.3 Growth Model Selection

Temporal growth patterns vary by field. Three common models:

| Model | Equation | When to use |
|-------|----------|-------------|
| Linear | $y = a + bt$ | Steady, predictable growth |
| Exponential | $y = ae^{bt}$ | Rapidly expanding fields |
| Logistic | $y = \frac{K}{1 + e^{-b(t - t_0)}}$ | Fields approaching saturation |

```{r growth-models, eval=FALSE}
# Compare growth models
# In practice, temporal_growth() returns fitted values
# You can compare model fits manually:

growth_df <- data.frame(
  year = growth$year,
  count = growth$count
)

# Linear model
fit_linear <- lm(count ~ year, data = growth_df)

# Log-linear (exponential proxy)
fit_log <- lm(log(count + 1) ~ year, data = growth_df)

# Compare R-squared values
cat("Linear R²:", summary(fit_linear)$r.squared, "\n")
cat("Log-linear R²:", summary(fit_log)$r.squared, "\n")
```

### 4.1.4 Comparing Growth Across Datasets

```{r growth-compare, eval=FALSE}
# If you have data from multiple databases
# ggplot(merged_df, aes(x = PY, y = count, fill = DB)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   labs(title = "Growth by Database",
#        x = "Year", y = "Publications",
#        fill = "Database") +
#   theme_minimal()
```

---

## 4.2 Citation Velocity: `citation_velocity()`

Citation velocity measures how quickly papers accumulate citations after publication. It is often defined as the total citations divided by the number of years since publication.

### 4.2.1 Basic Computation

```{r citation-velocity-basic, eval=FALSE}
# Compute citation velocity
bp <- as_biblio_project(example_biblio())
velocity <- citation_velocity(bp)

head(velocity, 10)
# Returns data frame with:
# - document identifier
# - publication year
# - total citations
# - years since publication
# - velocity (citations/year)
```

### 4.2.2 Interpreting Velocity

```{r velocity-interpretation, eval=FALSE}
# Summary statistics
summary(velocity$velocity)

# Classify velocity
velocity <- velocity %>%
  mutate(
    velocity_class = case_when(
      velocity >= 20 ~ "Very high",
      velocity >= 10 ~ "High",
      velocity >= 5  ~ "Medium",
      velocity >= 1  ~ "Low",
      TRUE           ~ "Minimal"
    )
  )

velocity %>%
  count(velocity_class) %>%
  arrange(desc(n)) %>%
  mutate(pct = round(n / sum(n) * 100, 1))
```

### 4.2.3 Field-Normalized Velocity

```{r velocity-normalized, eval=FALSE}
# Normalize velocity by field to enable cross-field comparison
velocity_norm <- velocity %>%
  group_by(field) %>%
  mutate(
    velocity_z = (velocity - mean(velocity, na.rm = TRUE)) /
                  sd(velocity, na.rm = TRUE),
    velocity_pct = percent_rank(velocity)
  ) %>%
  ungroup()

# Top 1% papers by normalized velocity
top_velocity <- velocity_norm %>%
  filter(velocity_pct >= 0.99) %>%
  arrange(desc(velocity_z))

cat("Top 1% papers by normalized citation velocity:\n")
print(head(top_velocity, 10))
```

### 4.2.4 Visualization

```{r velocity-plot, eval=FALSE}
ggplot(velocity, aes(x = velocity)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = median(velocity$velocity, na.rm = TRUE),
             linetype = "dashed", color = "red", size = 1) +
  scale_x_log10() +
  labs(
    title = "Citation Velocity Distribution",
    subtitle = "Red dashed line = median velocity",
    x = "Citations per year (log scale)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

---

## 4.3 Citation Trajectories: `citation_trajectory()`

Citation trajectories model how individual papers accumulate citations over their life cycle. Different patterns are possible:

- **Immediate impact**: High initial citations, declining over time
- **Slow burn**: Gradual increase, peak after 5-10 years
- **Sleeping beauty**: Low initial impact, sudden late recognition
- **Plateau**: Rapid rise followed by stable citation rate

### 4.3.1 Basic Computing

```{r trajectory-basic, eval=FALSE}
# Model citation trajectories
bp <- as_biblio_project(example_biblio())
trajectories <- citation_trajectory(bp)

head(trajectories, 10)
# Returns per-document trajectory parameters including peak year,
# curvature, and half-life estimates
```

### 4.3.2 Trajectory Classification

```{r trajectory-classify, eval=FALSE}
# Classify papers by trajectory type
trajectories <- trajectories %>%
  mutate(
    trajectory_type = case_when(
      peak_year <= pub_year + 2           ~ "Immediate impact",
      peak_year <= pub_year + 5           ~ "Early peak",
      peak_year <= pub_year + 10          ~ "Mid-career peak",
      is.na(peak_year)                    ~ "Still rising",
      TRUE                                ~ "Late bloomer"
    )
  )

# Summary by type
trajectory_dist <- trajectories %>%
  count(trajectory_type) %>%
  arrange(desc(n))

print(trajectory_dist)
```

### 4.3.3 Sleeping Beauties

A "sleeping beauty" is a paper that receives little attention initially but gains recognition years later. Detection criteria:

1. Low citation rate in first 3-5 years after publication
2. Sudden increase in citations after a dormancy period
3. Final citation count substantially above field average

```{r sleeping-beauties, eval=FALSE}
# Detect potential sleeping beauties
sleeping_beauties <- trajectories %>%
  filter(
    trajectory_type == "Late bloomer",
    total_citations > quantile(total_citations, 0.9, na.rm = TRUE)
  ) %>%
  arrange(desc(total_citations))

cat("Potential Sleeping Beauties:\n")
print(head(sleeping_beauties, 10))
```

### 4.3.4 Visualization

```{r trajectory-plot, eval=FALSE}
# Plot trajectories for selected papers
# Select top papers from different trajectory types
selected <- trajectories %>%
  group_by(trajectory_type) %>%
  slice_max(total_citations, n = 3) %>%
  ungroup()

ggplot(selected, aes(x = years_since_pub, y = cumulative_citations,
                      group = document_id, color = trajectory_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~trajectory_type, scales = "free_y") +
  labs(
    title = "Citation Trajectories by Type",
    x = "Years Since Publication",
    y = "Cumulative Citations"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

## 4.4 Reference Publication Year Spectroscopy: `rpys()`

Reference Publication Year Spectroscopy (RPYS) identifies years that are cited more frequently than expected, revealing historically seminal years in a field's development.

<div class="callout-note">
RPYS analyzes the *references* cited by papers in your corpus, not the papers themselves. It answers: "Which historical years had disproportionate influence on this field?"
</div>

### 4.4.1 How RPYS Works

1. Parse all cited references from the corpus
2. Extract publication years from the references
3. Count references per year
4. Compare observed counts against a baseline (e.g., 5-year moving median)
5. Identify years with significant positive deviations

### 4.4.2 Performing RPYS

```{r rpys-basic, eval=FALSE}
# Extract reference years from cited references
# The 'CR' column typically contains semicolon-separated reference strings
ref_years <- extract_reference_years(biblio_df$CR)

# Perform RPYS analysis
rpys_result <- rpys(ref_years)

head(rpys_result, 20)
# Returns data frame with:
# - ref_year: reference publication year
# - count: number of times this year was cited
# - median: 5-year moving median baseline
# - deviation: observed - baseline
# - deviation_z: standardized deviation
```

### 4.4.3 Identifying Seminal Years

```{r rpys-seminal, eval=FALSE}
# Filter statistically significant peaks (Z > 2)
seminal_years <- rpys_result %>%
  filter(deviation_z > 2) %>%
  arrange(desc(deviation_z))

cat("=== Seminal Years Detected ===\n")
cat(sprintf("Year | Citations | Z-score\n"))
cat(sprintf("-----|-----------|--------\n"))
for (i in seq_len(min(15, nrow(seminal_years)))) {
  cat(sprintf("%4d | %9d | %6.2f\n",
              seminal_years$ref_year[i],
              seminal_years$count[i],
              seminal_years$deviation_z[i]))
}
```

### 4.4.4 Visualization

```{r rpys-plot, eval=FALSE}
# Create spectrogram
ggplot(rpys_result, aes(x = ref_year, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = median), color = "red", size = 1) +
  geom_hline(yintercept = mean(rpys_result$count) +
               2 * sd(rpys_result$count),
             linetype = "dashed", color = "darkred") +
  geom_text(
    data = seminal_years %>% head(10),
    aes(label = ref_year, y = count + 5),
    size = 3, angle = 45, hjust = 0
  ) +
  labs(
    title = "Reference Publication Year Spectroscopy (RPYS)",
    subtitle = "Blue bars = cited references; Red line = 5-year median",
    x = "Reference Year",
    y = "Number of Citations"
  ) +
  theme_minimal(base_size = 14)
```

### 4.4.5 Interpreting RPYS Results

RPYS peaks typically correspond to three categories:

| Category | Description | Example |
|----------|-------------|---------|
| **Paradigm-establishing** | Foundational works that defined a field | Darwin (1859), Shannon (1948) |
| **Methodological breakthroughs** | New methods that became standard | PCR (1986), BLAST (1990) |
| **Landmark datasets** | Released data that enabled research | Human Genome (2001) |

<div class="callout-tip">
**Practical tip:** Cross-reference RPYS peaks with citation counts for specific references. A peak year may reflect one highly cited work or many moderately cited works from the same year.
</div>

### 4.4.6 Case Study: Terahertz Research

Consider a corpus of 500 papers on terahertz spectroscopy. RPYS analysis might reveal:

```{r rpys-case-study, eval=FALSE}
# Hypothetical RPYS results for terahertz field
rpys_example <- data.frame(
  ref_year = c(1960, 1971, 1974, 1985, 1995, 2001, 2004),
  citations = c(45, 120, 89, 67, 234, 456, 321),
  deviation_z = c(2.1, 3.8, 2.9, 2.3, 4.5, 5.2, 3.1)
)

cat("Seminal Years in Terahertz Spectroscopy:\n")
cat("========================================\n\n")

# 1971: Schawlow and Townes paper on lasers → enabled THz generation
cat("1971 (Z=3.8): Foundational work on laser spectroscopy\n")

# 1995: First THz time-domain spectroscopy papers
cat("1995 (Z=4.5): Development of THz-TDS methods\n")

# 2001: Explosion of THz applications
cat("2001 (Z=5.2): Broad applications in security and imaging\n\n")

cat("Interpretation:\n")
cat("- The 1971 peak reflects a methodological breakthrough\n")
cat("- The 1995 peak marks method maturation\n")
cat("- The 2001 peak signals application-driven growth\n")
```

### 4.4.7 RPYS Limitations and Caveats

RPYS has several limitations to consider:

1. **Database coverage bias**: If your corpus is from Scopus, references to very old works (pre-1970) may be underrepresented
2. **Self-citation inflation**: Fields with heavy self-citation may show self-referential peaks
3. **Language bias**: English-language databases underrepresent work published in other languages
4. **Parsing errors**: Reference string parsing is imperfect; some years may be extracted incorrectly

```{r rpys-caveats, eval=FALSE}
# Check for parsing issues in reference years
biblio_df %>%
  filter(!is.na(CR)) %>%
  mutate(
    n_refs = stringi::stri_count_fixed(CR, ";") + 1,
    has_refs = nchar(CR) > 0
  ) %>%
  summarise(
    total_papers = n(),
    papers_with_refs = sum(has_refs),
    mean_refs = mean(n_refs[has_refs]),
    median_refs = median(n_refs[has_refs])
  )

# Flag potential parsing issues
# Very old years might be parsing errors
# Very recent years (future) are definitely errors
rpys_result %>%
  filter(ref_year < 1900 | ref_year > as.integer(format(Sys.Date(), "%Y"))) %>%
  arrange(ref_year)
```

---

## 4.5 Normalized Citations: `normalized_citations()`

Raw citation counts are misleading for cross-field comparisons. A paper with 50 citations in mathematics is exceptional; in biomedicine, it may be average. `normalized_citations()` addresses this.

### 4.5.1 Field Normalization

```{r normalized-field, eval=FALSE}
# Normalize citations by field
normalized <- normalized_citations(biblio_df,
                                    citations = "TC",
                                    field = "WC",
                                    method = "mean")

head(normalized, 10)
# Returns data frame with original citations, field mean,
# and normalized value (ratio of observed to expected)
```

### 4.5.2 Time-Window Normalization

```{r normalized-time, eval=FALSE}
# Normalize by field AND publication year
normalized_multi <- normalized_citations(biblio_df,
                                          citations = "TC",
                                          field = "WC",
                                          pub_year = "PY",
                                          method = "percentile",
                                          window = 5)

# Papers with normalized citations > 1.0 are above field-year average
top_normalized <- normalized_multi %>%
  filter(normalized_citations > 2.0) %>%
  arrange(desc(normalized_citations))

cat("Papers with 2x+ field-year normalized citations:\n")
print(head(top_normalized, 10))
```

### 4.5.3 Normalization Methods

| Method | Formula | Interpretation |
|--------|---------|----------------|
| Mean | $N_c = TC / \bar{TC}_{field}$ | Ratio to field average |
| Median | $N_c = TC / \tilde{TC}_{field}$ | Ratio to field median |
| Percentile | $N_c = rank / n$ | Position within field distribution |

### 4.5.4 Visualization

```{r normalized-plot, eval=FALSE}
ggplot(normalized, aes(x = reorder(field, normalized_citations, FUN = median),
                       y = normalized_citations)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "Field-Normalized Citation Impact",
    x = NULL,
    y = "Normalized Citations (1.0 = field average)"
  ) +
  theme_minimal(base_size = 12)
```

---

## 4.6 Disruption Index: `disruption_index()`

The disruption index (Wu et al., 2019) distinguishes transformative research from incremental or consolidating work. A disruptive paper creates a new research direction; a consolidating paper builds on existing lines.

### 4.6.1 Computing Disruption

```{r disruption-basic, eval=FALSE}
# Compute disruption index
disruption <- disruption_index(biblio_df)

head(disruption, 10)
# Returns data frame with:
# - document identifier
# - disruption index (D), ranging from -1 to +1
# - component counts (f, g, h)
```

### 4.6.2 Interpreting D Values

The disruption index $D$ ranges from -1 (fully consolidating) to +1 (fully disruptive):

| D Value | Interpretation |
|---------|---------------|
| D > 0.5 | Highly disruptive |
| 0 < D < 0.5 | Mildly disruptive |
| D ≈ 0 | Neutral |
| -0.5 < D < 0 | Mildly consolidating |
| D < -0.5 | Highly consolidating |

```{r disruption-interpretation, eval=FALSE}
# Classify papers
disruption <- disruption %>%
  mutate(
    disruption_class = case_when(
      disruption_index >  0.5 ~ "Highly disruptive",
      disruption_index >  0   ~ "Mildly disruptive",
      disruption_index > -0.5 ~ "Mildly consolidating",
      TRUE                    ~ "Highly consolidating"
    )
  )

disruption %>%
  count(disruption_class) %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  arrange(desc(n))
```

### 4.6.3 Disruption vs Citations

```{r disruption-scatter, eval=FALSE}
ggplot(disruption, aes(x = disruption_index, y = log10(total_citations + 1))) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", color = "#E74C3C") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Disruption Index vs Citation Count",
    subtitle = "Do disruptive papers get cited more?",
    x = "Disruption Index",
    y = "log10(Citations + 1)"
  ) +
  theme_minimal(base_size = 14)
```

---

# 5. Text Analysis

## 5.1 Term Frequency: `term_frequency()`

Term frequency (TF) counts how often terms appear in bibliographic text fields (titles, abstracts, keywords).

### 5.1.1 Basic Term Frequency

```{r term-frequency-basic, eval=FALSE}
# Compute term frequency from titles
bp <- as_biblio_project(example_biblio())
tf <- term_frequency(bp)

head(tf, 20)
# Returns data frame with:
# - term: the word or phrase
# - frequency: count in corpus
# - proportion: relative frequency
```

### 5.1.2 Multi-Field Analysis

```{r tf-multi-field, eval=FALSE}
# Compare terms across title, abstract, and keywords
tf_ti <- term_frequency(bp, text_field = "TI")
tf_ab <- term_frequency(bp, text_field = "AB")
tf_kw <- term_frequency(bp, text_field = "DE")

# Combine for comparison
tf_combined <- bind_rows(
  tf_ti %>% head(15) %>% mutate(field = "Title"),
  tf_ab %>% head(15) %>% mutate(field = "Abstract"),
  tf_kw %>% head(15) %>% mutate(field = "Keywords")
)

ggplot(tf_combined, aes(x = reorder(term, frequency), y = frequency)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~field, scales = "free") +
  labs(title = "Top Terms by Text Field",
       x = NULL, y = "Frequency") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

### 5.1.3 Temporal Evolution of Terms

```{r tf-over-time, eval=FALSE}
# Track term prevalence over time
if (requireNamespace("tidytext", quietly = TRUE)) {
  library(tidytext)
  
  tf_by_year <- biblio_df %>%
    filter(!is.na(PY)) %>%
    unnest_tokens(word, TI) %>%
    anti_join(stop_words, by = "word") %>%
    count(PY, word) %>%
    group_by(PY) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()
  
  # Track specific terms
  track_terms <- c("climate", "water", "model", "analysis", "system")
  
  tf_tracked <- tf_by_year %>%
    filter(word %in% track_terms)
  
  ggplot(tf_tracked, aes(x = PY, y = proportion, color = word)) +
    geom_line(size = 1.2, se = FALSE) +
    geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
    labs(title = "Term Prevalence Over Time",
         x = "Year", y = "Proportion of Titles",
         color = "Term") +
    theme_minimal(base_size = 14)
}
```

### 5.1.4 N-gram Analysis

Bigrams (two-word phrases) and trigrams capture multi-word concepts that single terms miss. For example, "climate change" is far more meaningful than counting "climate" and "change" separately.

```{r ngrams, eval=FALSE}
# Bigram analysis captures multi-word concepts
if (requireNamespace("tidytext", quietly = TRUE)) {
  bigrams <- biblio_df %>%
    filter(!is.na(TI)) %>%
    unnest_tokens(bigram, TI, token = "ngrams", n = 2) %>%
    separate(bigram, c("word1", "word2"), sep = " ") %>%
    filter(!word1 %in% stop_words$word,
           !word2 %in% stop_words$word) %>%
    unite(bigram, word1, word2, sep = " ") %>%
    count(bigram, sort = TRUE)
  
  cat("Top Bigrams in Titles:\n")
  print(head(bigrams, 20))
}

# Trigram analysis for longer phrases
if (requireNamespace("tidytext", quietly = TRUE)) {
  trigrams <- biblio_df %>%
    filter(!is.na(TI)) %>%
    unnest_tokens(trigram, TI, token = "ngrams", n = 3) %>%
    count(trigram, sort = TRUE) %>%
    head(15)
  
  cat("\nTop Trigrams in Titles:\n")
  print(trigrams)
}
```

### 5.1.5 Keyword Co-occurrence

Analyzing which keywords appear together reveals thematic clusters in a research field.

```{r keyword-cooccurrence, eval=FALSE}
# Co-occurrence analysis of author keywords
if (requireNamespace("tidytext", quietly = TRUE)) {
  keyword_pairs <- biblio_df %>%
    filter(!is.na(DE)) %>%
    mutate(keywords = strsplit(DE, ";\\s*")) %>%
    unnest(keywords) %>%
    mutate(keywords = tolower(trimws(keywords))) %>%
    filter(nchar(keywords) > 2) %>%
    # Create all keyword pairs within each document
    group_by(document_id) %>%
    summarise(
      pairs = if(n() > 1) list(combn(keywords, 2, simplify = FALSE)) else list(NULL),
      .groups = "drop"
    ) %>%
    unnest(pairs) %>%
    filter(!sapply(pairs, is.null)) %>%
    mutate(
      kw1 = sapply(pairs, function(x) sort(x)[1]),
      kw2 = sapply(pairs, function(x) sort(x)[2])
    ) %>%
    count(kw1, kw2, sort = TRUE) %>%
    head(20)
  
  cat("Top Keyword Pairs:\n")
  print(keyword_pairs)
}
```

---

## 5.2 TF-IDF Analysis: `tfidf_terms()`

Term Frequency-Inverse Document Frequency (TF-IDF) weights terms by their ability to distinguish documents. A high TF-IDF score means a term is important to a specific document, not just common across the corpus.

### 5.2.1 The TF-IDF Formula

$$\text{TF-IDF}(t, d) = \text{TF}(t, d) \times \text{IDF}(t)$$

Where:

$$\text{TF}(t, d) = \frac{\text{count of } t \text{ in } d}{\text{total terms in } d}$$

$$\text{IDF}(t) = \log\frac{N}{\text{documents containing } t}$$

### 5.2.2 Basic TF-IDF

```{r tfidf-basic, eval=FALSE}
# Compute TF-IDF grouped by journal
bp <- as_biblio_project(example_biblio())
tfidf <- tfidf_terms(bp)

head(tfidf, 20)
# Returns data frame with:
# - group: grouping variable (e.g., journal)
# - term: the word
# - tf_idf: TF-IDF score
# - n_documents: documents in group
```

### 5.2.3 Interpreting TF-IDF Scores

```{r tfidf-interpretation, eval=FALSE}
# Top distinguishing terms per journal
top_tfidf <- tfidf %>%
  group_by(journal) %>%
  slice_max(tf_idf, n = 5) %>%
  ungroup()

# Create summary
cat("Distinguishing Terms by Journal:\n")
cat("================================\n\n")

current_journal <- ""
for (i in seq_len(min(30, nrow(top_tfidf)))) {
  j <- top_tfidf$journal[i]
  if (j != current_journal) {
    current_journal <- j
    cat(sprintf("\n[%s]\n", current_journal))
  }
  cat(sprintf("  %s (TF-IDF = %.3f)\n",
              top_tfidf$term[i], top_tfidf$tf_idf[i]))
}
```

### 5.2.4 TF-IDF Heatmap

```{r tfidf-heatmap, eval=FALSE}
# Heatmap of TF-IDF across journals
ggplot(top_tfidf, aes(x = reorder(term, tf_idf),
                       y = reorder(journal, tf_idf),
                       fill = tf_idf)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "white", high = "#E74C3C", midpoint = 0.025) +
  labs(title = "TF-IDF: Distinguishing Terms by Journal",
       x = NULL, y = NULL, fill = "TF-IDF") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 5.2.5 TF-IDF vs Raw Frequency

```{r tfidf-comparison, eval=FALSE}
# Compare TF-IDF ranking with raw frequency ranking
comparison <- tf %>%
  select(term, frequency) %>%
  inner_join(
    tfidf %>%
      group_by(term) %>%
      summarise(mean_tfidf = mean(tf_idf, na.rm = TRUE), .groups = "drop"),
    by = "term"
  ) %>%
  mutate(
    rank_freq = rank(-frequency),
    rank_tfidf = rank(-mean_tfidf),
    rank_diff = rank_freq - rank_tfidf
  ) %>%
  arrange(desc(abs(rank_diff)))

cat("Terms with largest rank difference (frequency vs TF-IDF):\n")
print(head(comparison %>% select(term, frequency, mean_tfidf, rank_diff), 15))
```

---

## 5.3 Trending Topics: `trend_topics()`

Trend detection identifies terms and topics gaining or losing prominence over time.

### 5.3.1 Methods for Trend Detection

| Method | Description | Best for |
|--------|-------------|----------|
| Growth | Change in term frequency between time windows | Broad trend detection |
| Emergence | New terms appearing recently | Very new topics |
| Burst | Sudden spikes in term frequency | Viral or policy-driven topics |

### 5.3.2 Basic Trend Detection

```{r trend-basic, eval=FALSE}
# Detect trending topics
bp <- as_biblio_project(example_biblio())
trends <- trend_topics(bp)

head(trends, 20)
# Returns data frame with:
# - topic: the term
# - first_year: first appearance
# - last_year: most recent appearance
# - growth_rate: rate of increase
# - momentum: trend strength
# - trend_type: "emerging", "established", "declining"
```

### 5.3.3 Trend Visualization

```{r trend-plot, eval=FALSE}
# Plot emerging topics
emerging <- trends %>%
  filter(trend_type == "emerging") %>%
  head(10)

# Track these topics over time
if (requireNamespace("tidytext", quietly = TRUE)) {
  topic_traces <- biblio_df %>%
    filter(!is.na(PY), !is.na(TI)) %>%
    unnest_tokens(word, TI) %>%
    filter(word %in% emerging$topic) %>%
    count(PY, word, name = "frequency")
  
  ggplot(topic_traces, aes(x = PY, y = frequency, color = word)) +
    geom_line(size = 1) +
    geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +
    labs(title = "Emerging Topics Over Time",
         x = "Year", y = "Frequency in Titles",
         color = "Topic") +
    theme_minimal(base_size = 12)
}
```

### 5.3.4 Growth Rate Analysis

```{r growth-rate-analysis, eval=FALSE}
# Compute annual growth rates for top trends
if (requireNamespace("tidytext", quietly = TRUE)) {
  topic_growth <- biblio_df %>%
    filter(!is.na(PY), !is.na(TI)) %>%
    unnest_tokens(word, TI) %>%
    filter(word %in% head(trends$topic, 10)) %>%
    count(PY, word, name = "frequency") %>%
    group_by(word) %>%
    arrange(PY) %>%
    mutate(
      growth_rate = (frequency - lag(frequency)) / lag(frequency),
      cumulative = cumsum(frequency)
    ) %>%
    ungroup()
  
  # Summarize growth patterns
  growth_summary <- topic_growth %>%
    group_by(word) %>%
    summarise(
      total_mentions = sum(frequency, na.rm = TRUE),
      mean_growth = mean(growth_rate, na.rm = TRUE),
      recent_growth = mean(growth_rate[PY >= 2018], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(recent_growth))
  
  cat("Topic Growth Summary:\n")
  print(growth_summary)
}
```

### 5.3.5 Identifying Declining Topics

```{r declining-topics, eval=FALSE}
# Find declining topics
declining <- trends %>%
  filter(trend_type == "declining") %>%
  arrange(growth_rate) %>%
  head(10)

cat("Declining Topics:\n")
cat("=================\n")
for (i in seq_len(nrow(declining))) {
  cat(sprintf("  %s (growth rate: %.3f, last seen: %d)\n",
              declining$topic[i],
              declining$growth_rate[i],
              declining$last_year[i]))
}
```

---

# 6. Step-by-Step Workflow

This section demonstrates a complete workflow combining temporal and textual analyses.

## Step 1: Load and Prepare Data

```{r workflow-step1, eval=FALSE}
library(biblioIntegrator)

# Load data
bp <- as_biblio_project(example_biblio())

# Quick overview
cat("=== Dataset Overview ===\n")
cat("Papers:", nrow(bp), "\n")
cat("Year range:", range(bp$PY, na.rm = TRUE), "\n")
cat("Fields:", length(unique(bp$WC)), "\n")
cat("Missing data:",
    sum(is.na(bp$PY)), "missing PY,",
    sum(is.na(bp$TC)), "missing TC\n")
```

## Step 2: Temporal Analysis

```{r workflow-step2, eval=FALSE}
cat("\n=== TEMPORAL ANALYSIS ===\n")

# Publication growth
growth <- temporal_growth(bp)
cat("Peak year:", growth$year[which.max(growth$count)], "\n")

# Citation velocity
velocity <- citation_velocity(bp)
cat("Median velocity:", median(velocity$velocity, na.rm = TRUE), "citations/year\n")

# RPYS (if references available)
# ref_years <- extract_reference_years(bp$CR)
# rpys_result <- rpys(ref_years)
```

## Step 3: Citation Analysis

```{r workflow-step3, eval=FALSE}
cat("\n=== CITATION ANALYSIS ===\n")

# Normalized citations
normalized <- normalized_citations(bp, citations = "TC", field = "WC")

# Top papers
top_papers <- normalized %>%
  slice_max(normalized_citations, n = 10) %>%
  select(document_id, field, citations, normalized_citations)

cat("Top 10 by normalized impact:\n")
print(top_papers)

# Disruption index
disruption <- disruption_index(bp)
cat("\nMean disruption index:", mean(disruption$disruption_index, na.rm = TRUE), "\n")
```

## Step 4: Text Analysis

```{r workflow-step4, eval=FALSE}
cat("\n=== TEXT ANALYSIS ===\n")

# Term frequency
tf <- term_frequency(bp)
cat("Top 5 terms in titles:\n")
print(head(tf, 5))

# TF-IDF
tfidf <- tfidf_terms(bp)
cat("\nTop distinguishing terms per journal:\n")
print(head(tfidf, 10))

# Trending topics
trends <- trend_topics(bp)
cat("\nEmerging topics:\n")
print(trends %>% filter(trend_type == "emerging") %>% head(5))
```

## Step 5: Synthesis Report

```{r workflow-step5, eval=FALSE}
cat("\n")
cat("=============================================\n")
cat("  COMPREHENSIVE ANALYSIS SUMMARY\n")
cat("=============================================\n\n")

cat("TEMPORAL PATTERNS:\n")
cat(sprintf("  Dataset: %d papers, %d-%d\n",
            nrow(bp),
            min(bp$PY, na.rm = TRUE),
            max(bp$PY, na.rm = TRUE)))
cat(sprintf("  Peak year: %d\n", growth$year[which.max(growth$count)]))
cat(sprintf("  Median citations: %.1f\n", median(bp$TC, na.rm = TRUE)))

cat("\nCITATION DYNAMICS:\n")
cat(sprintf("  Mean velocity: %.2f citations/year\n", mean(velocity$velocity, na.rm = TRUE)))
cat(sprintf("  Highly disruptive papers: %d (%.1f%%)\n",
            sum(disruption$disruption_index > 0.5, na.rm = TRUE),
            100 * mean(disruption$disruption_index > 0.5, na.rm = TRUE)))

cat("\nTOP TERMS:\n")
for (i in 1:min(5, nrow(tf))) {
  cat(sprintf("  %d. %s (%d mentions)\n", i, tf$term[i], tf$frequency[i]))
}

cat("\nEMERGING TOPICS:\n")
emerging <- trends %>% filter(trend_type == "emerging") %>% head(5)
for (i in seq_len(nrow(emerging))) {
  cat(sprintf("  %d. %s (first: %d, growth: %.3f)\n",
              i, emerging$topic[i], emerging$first_year[i], emerging$growth_rate[i]))
}

cat("\n=============================================\n")
```

---

# 7. Common Mistakes and Pitfalls

## 7.1 Ignoring Field-Specific Citation Patterns

Different fields have vastly different citation cultures. Medicine and molecular biology produce many more citations per paper than mathematics or engineering. Comparing raw citation counts across fields is meaningless.

```{r pitfall-field, eval=FALSE}
# WRONG: raw comparison across fields
# biblio_df %>% group_by(WC) %>% summarise(mean_citations = mean(TC))

# RIGHT: field normalization
normalized <- normalized_citations(biblio_df, citations = "TC", field = "WC")
normalized %>%
  group_by(field) %>%
  summarise(mean_norm = mean(normalized_citations, na.rm = TRUE))
```

## 7.2 Not Normalizing for Time Window

A paper published in 2010 has had 14 years to accumulate citations; one from 2023 only 1 year. Use `citation_velocity()` with an appropriate time window.

```{r pitfall-time, eval=FALSE}
# Use citation velocity instead of raw counts
velocity <- citation_velocity(bp)
# velocity accounts for years since publication
```

## 7.3 Confusing Frequency with Importance

High term frequency ≠ high importance. "Analysis" may appear in 40% of titles without telling us anything distinctive. Use TF-IDF to find distinguishing terms.

```{r pitfall-frequency, eval=FALSE}
# WRONG: assuming frequency = importance
# Use TF-IDF instead
tfidf <- tfidf_terms(bp)
# High TF-IDF = terms that distinguish groups
```

## 7.4 Ignoring Citation Inflation

The total volume of publications has grown exponentially. A paper from 1990 competed in a smaller pool of potential citers than one from 2020. Time normalization adapts for this.

```{r pitfall-inflation, eval=FALSE}
# Normalize by both field AND year
normalized <- normalized_citations(biblio_df,
                                    citations = "TC",
                                    field = "WC",
                                    pub_year = "PY",
                                    window = 5)
```

## 7.5 Treating RPYS Peaks as Quality Indicators

An RPYS peak marks historical influence, not necessarily quality. A peak year may correspond to a controversial paper that was heavily criticized and therefore heavily cited. Context matters.

## 7.6 Over-Interpreting Trend Topics

A topic with 3 mentions in the last year has a very high growth rate if it had 0 mentions before. Always check absolute counts alongside growth metrics.

```{r pitfall-trends, eval=FALSE}
# Check both growth rate AND absolute counts
trends %>%
  mutate(
    confidence = case_when(
      total_mentions >= 50 ~ "High",
      total_mentions >= 10 ~ "Medium",
      TRUE                 ~ "Low"
    )
  ) %>%
  filter(trend_type == "emerging") %>%
  arrange(desc(confidence), desc(growth_rate))
```

## 7.7 Ignoring Stopwords and Domain-Specific Terms

Generic stopwords ("the", "and", "of") inflate frequency counts. Domain stopwords ("study", "result", "method") should also be removed for meaningful analysis.

```{r pitfall-stopwords, eval=FALSE}
# Always remove stopwords
if (requireNamespace("tidytext", quietly = TRUE)) {
  library(tidytext)
  
  # Standard stopwords
  clean_terms <- biblio_df %>%
    unnest_tokens(word, TI) %>%
    anti_join(stop_words)
  
  # Add domain-specific stopwords
  domain_stop <- data.frame(word = c("study", "results", "analysis",
                                      "method", "methods", "data",
                                      "approach", "paper", "research"))
  
  clean_terms <- clean_terms %>%
    anti_join(domain_stop)
}
```

---

# 8. Advanced Topics

## 8.1 Combining Temporal and Textual Analysis

```{r combined-analysis, eval=FALSE}
# How do term usages evolve across time periods?
if (requireNamespace("tidytext", quietly = TRUE)) {
  library(tidytext)
  
  biblio_periods <- biblio_df %>%
    mutate(
      period = cut(PY,
                   breaks = c(1990, 2000, 2010, 2020, 2025),
                   labels = c("1990s", "2000s", "2010s", "2020s"),
                   include.lowest = TRUE)
    ) %>%
    filter(!is.na(period))
  
  period_tf <- biblio_periods %>%
    unnest_tokens(word, TI) %>%
    anti_join(stop_words) %>%
    count(period, word) %>%
    group_by(period) %>%
    mutate(proportion = n / sum(n)) %>%
    ungroup()
  
  # Track key terms across decades
  key_terms <- c("climate", "model", "system", "analysis", "water")
  evolution <- period_tf %>%
    filter(word %in% key_terms)
  
  ggplot(evolution, aes(x = period, y = proportion, fill = word)) +
    geom_col(position = "dodge") +
    labs(title = "Evolution of Key Terms Across Decades",
         x = "Decade", y = "Proportion of Titles",
         fill = "Term") +
    theme_minimal(base_size = 14) +
    scale_fill_brewer(palette = "Set2")
}
```

## 8.2 Cross-Database Temporal Comparison

```{r cross-db, eval=FALSE}
# Compare temporal patterns across Scopus and WoS
if ("DB" %in% names(biblio_df)) {
  db_growth <- biblio_df %>%
    filter(!is.na(PY), !is.na(DB)) %>%
    count(DB, PY, name = "publications") %>%
    group_by(DB) %>%
    mutate(cumulative = cumsum(publications)) %>%
    ungroup()
  
  ggplot(db_growth, aes(x = PY, y = cumulative, color = DB)) +
    geom_line(size = 1.2) +
    labs(title = "Cumulative Growth by Database Source",
         x = "Year", y = "Cumulative Publications",
         color = "Database") +
    theme_minimal(base_size = 14)
}
```

## 8.3 Network-Temporal Integration

For combined network and temporal analysis (e.g., how network position changes over time), see vignette 04 ("Network Analysis"). The basic approach is:

```{r network-temporal, eval=FALSE}
# Conceptual workflow for network-temporal analysis
# 1. Split data by time periods
# 2. Build citation networks per period
# 3. Compute network metrics per period
# 4. Track metric evolution

# Example pseudocode:
# periods <- 2010:2023
# metrics_by_period <- lapply(periods, function(yr) {
#   year_data <- biblio_df %>% filter(PY == yr)
#   net <- build_citation_network(year_data)
#   data.frame(year = yr,
#              n_nodes = vcount(net),
#              n_edges = ecount(net),
#              density = edge_density(net))
# })
```

---

# 9. Function Reference Summary

| Function | Purpose | Key Parameters | Returns |
|----------|---------|----------------|---------|
| `temporal_growth()` | Publication counts and growth model | `model`, `field` | year, count, fitted values |
| `citation_velocity()` | Citations per year since publication | `window`, `pub_year`, `cite_count` | velocity per document |
| `citation_trajectory()` | Citation lifecycle modeling | `model`, `current_year` | peak year, half-life, trajectory type |
| `rpys()` | Reference year spectroscopy | `ref_year`, `parse_refs`, `window` | year, count, deviation, Z-score |
| `normalized_citations()` | Field/time normalization | `method`, `field`, `pub_year`, `window` | normalized citation count |
| `disruption_index()` | Disruptive vs consolidating | `method`, `references` | D index, component counts |
| `term_frequency()` | Raw term counts | `n_terms`, `min_freq`, `remove_stopwords` | term, frequency, proportion |
| `tfidf_terms()` | TF-IDF weighting | `group_field`, `n_terms` | group, term, tf_idf |
| `trend_topics()` | Trending topic detection | `method`, `window`, `n_topics` | topic, growth_rate, trend_type |

---

# 10. References

## Methodological Papers

- **Aria, M., & Cuccurullo, C.** (2017). bibliometrix: An R-tool for comprehensive science mapping analysis. *Journal of Informetrics*, 11(4), 959-975. doi:[10.1016/j.joi.2017.08.007](https://doi.org/10.1016/j.joi.2017.08.007)

- **Bornmann, L., & Mutz, R.** (2015). Growth rates of modern science: A bibliometric analysis based on the number of publications and cited references. *Journal of the Association for Information Science and Technology*, 66(11), 2215-2222.

- **Marx, W., & Bornmann, L.** (2014). Tracing the origin of a scientific idea: Reference Publication Year Spectroscopy (RPYS). *ISSI Newsletter*, 10(1), 58-62.

- **Wu, L., Wang, D., & Evans, J. A.** (2019). Large teams develop and small teams disrupt science and technology. *Nature*, 566(7744), 378-382.

- **Sinatra, R., Wang, D., Deville, P., Song, C., & Barabási, A.-L.** (2016). Quantifying the evolution of individual scientific impact. *Science*, 354(6312), aaf5239.

## Temporal Analysis Methods

- **Ke, Q., Ferrara, E., Radicchi, F., & Flammini, A.** (2015). Defining and identifying Sleeping Beauties in science. *Proceedings of the National Academy of Sciences*, 112(24), 7426-7431.

- **Avramescu, A.** (1979). Actuality and obsolescence of scientific literature. *Journal of the American Society for Information Science*, 30(5), 296-303.

- **Baumgartner, S. E., & Leydesdorff, L.** (2014). Group-based trajectory modeling (GBTM) of citations in scholarly literature. *Journal of the Association for Information Science and Technology*, 65(3), 577-588.

## Text Analysis Methods

- **Aizawa, A.** (2003). An information-theoretic perspective of tf-idf measures. *Information Processing & Management*, 39(1), 45-65.

- **Blei, D. M., Ng, A. Y., & Jordan, M. I.** (2003). Latent Dirichlet allocation. *Journal of Machine Learning Research*, 3, 993-1022.

- **Ramage, D., Hall, D., Nallapati, R., & Manning, C. D.** (2009). Labeled LDA: A supervised topic model for credit attribution in multi-labeled corpora. *Proceedings of EMNLP*, 248-256.

- **Blei, D. M.** (2012). Probabilistic topic models. *Communications of the ACM*, 55(4), 77-84.

## Bibliometric Analysis

- **van Eck, N. J., & Waltman, L.** (2010). Software survey: VOSviewer, a computer program for bibliometric mapping. *Scientometrics*, 84(2), 523-538.

- **Callon, M., Courtial, J.-P., Turner, W. A., & Bauin, S.** (1983). From translations to problematic networks: An introduction to co-word analysis. *Social Science Information*, 22(2), 191-235.

## R Packages

- **Wickham, H.** (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag.

- **Silge, J., & Robinson, D.** (2017). *Text Mining with R: A Tidy Approach*. O'Reilly Media.

- **Feinerer, I., Hornik, K., & Meyer, D.** (2008). Text Mining Infrastructure in R. *Journal of Statistical Software*, 25(5), 1-54.

---

# Session Info

```{r session-info, eval=FALSE}
sessionInfo()
```

---

<div style="text-align: center; margin-top: 40px; padding: 20px; background-color: #f8f9fa; border-radius: 8px;">
<p style="margin: 0; color: #6c757d; font-size: 0.9em;">
<b>biblioIntegrator</b> version 0.2.0 |
Vignette: Temporal and Textual Analysis |
Last updated: `r Sys.Date()`
</p>
</div>
