---
title: "Foundations-to-advanced tutorial"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundations-to-advanced tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biblioIntegrator)
```

# Purpose

This tutorial links data architecture, audit, descriptive analysis, formal group comparison, networks, sensitivity, scalable storage and reporting in one agronomic workflow.

# Workflow

```{r}
library(biblioIntegrator)
raw <- example_biblio()
x <- as_biblio_project(raw, source = "agronomy tutorial")
# 1. Diagnose the corpus
biblio_health(x)
# 2. Describe production and impact
describe_biblio(x)
biblio_metrics(x)
# 3. Compare periods with inferential support
period <- ifelse(x$works$year <= 2021, "earlier", "recent")
cmp <- compare_groups(x, period, permutations = 199, bootstrap = 99, seed = 42)
cmp
head(association_residuals(cmp), 10)
# 4. Study the collaboration structure
g <- bibliographic_network(x, "coauthor")
head(network_centrality(g))
network_communities(g)
network_stability(x, B = 30, seed = 42)
# 5. Examine topical change
trend_topics(x)
head(tfidf_terms(x))
# 6. Check threshold sensitivity
sensitivity_analysis(x, period, 1:2, permutations = 99, seed = 42)
# 7. Generate an auditable report
f <- tempfile(fileext = ".md")
biblio_report(x, f)
readLines(f, n = 12)
```

# Interpretation

Results should be interpreted in relation to database coverage, time window, entity normalization and analytical thresholds. Provenance should be retained whenever data are merged, deduplicated or enriched.
