---
title: "Scalable storage and backends"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scalable storage and backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biblioIntegrator)
```

# Purpose

Arrow and DuckDB are optional storage engines. They keep the core package lightweight while enabling larger-than-memory workflows and SQL queries.

# Workflow

```{r}
backend_status()
x <- as_biblio_project(example_biblio())
if (requireNamespace("arrow", quietly = TRUE)) {
  p <- tempfile()
  biblio_store(x, p, "arrow")
  nrow(biblio_load(p, "arrow")$works)
}
if (requireNamespace("duckdb", quietly = TRUE) && requireNamespace("DBI", quietly = TRUE)) {
  p <- tempfile(fileext = ".duckdb")
  biblio_store(x, p, "duckdb")
  biblio_query(p, "SELECT year, COUNT(*) AS n FROM works GROUP BY year ORDER BY year")
}
```

# Interpretation

Results should be interpreted in relation to database coverage, time window, entity normalization and analytical thresholds. Provenance should be retained whenever data are merged, deduplicated or enriched.
