---
title: "Python Integration, Reports and Interactive Use"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Python Integration, Reports and Interactive Use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
library(biblioIntegrator)
suppressPackageStartupMessages({
  if (requireNamespace("knitr", quietly = TRUE)) library(knitr)
})
```

# Why This Vignette Exists {#sec-why}

## The Three Bridges

Bibliometric analysis in R is powerful, but three complementary
capabilities extend its reach beyond what a single language can
offer:

1. **Python integration** — The
   [Biblium](https://pypi.org/project/biblium/) library implements a
   permutation-based group comparison interface that mirrors the native
   R engine.  Running both engines on the same data and comparing their
   outputs provides an independent computational check.  When Biblium
   is available, `biblioIntegrator` can call it transparently via
   `reticulate` and convert results back into the same R structure
   used by `compare_groups()`.

2. **Parametric reports** — A `biblio_project` can be summarised into
   a single Markdown, HTML, Word, or PDF file by `biblio_report()`.
   The report captures corpus health, descriptive metrics, annual
   production, leading terms, network centrality, backend availability,
   and provenance.  Reports are self-contained and auditable: every
   number traces to the underlying relational tables.

3. **Interactive exploration** — `biblio_app()` launches a Shiny
   workbench with five tabs (Data, Explore, Networks, Compare, Report).
   The app lets non-programmers load data, inspect quality, visualise
   annual production, browse centrality tables, and download a report
   — all without writing a single line of R code.

Together, these three bridges allow a research team to:

- Run comparative inference in **two independent engines** and quantify
  agreement.
- Generate **reproducible reports** that document every analytical
  decision.
- Provide a **point-and-click interface** for collaborators who prefer
  not to use the R console.

## What This Vignette Does *Not* Cover

Python integration and Biblium are **optional**.  Every analytical
function discussed in earlier vignettes (`compare_groups()`,
`association_residuals()`, `group_ca()`, `sensitivity_analysis()` etc.)
works without a Python backend.  This vignette should be read as an
*extension*, not a prerequisite.

Package architecture, data model, import, harmonization, quality
filtering, descriptive analysis, comparative inference, networks,
temporal and text analysis, and scalable storage backends are covered
in vignettes `v00` through `v08`.

## Intended Audience

- Researchers who want a **second engine** to validate group
  comparisons.
- Analysts who need **one-command reports** for supervisors, reviewers,
  or grant applications.
- Team leads who want to hand a **Shiny app** to a collaborator
  unfamiliar with R scripting.
- R package developers interested in how `biblioIntegrator` integrates
  optional backends through `reticulate`.

# Learning Objectives {#sec-objectives}

After reading this vignette and running the code, you will be able to:

1. **Check Python availability** — use `python_backend_status()` and
   `biblium_backend_status()` to inspect the current backend
   configuration.
2. **Configure a Python executable** — set the Python path with
   `enable_python_backend()`, the `BIBLIOINTEGRATOR_PYTHON` environment
   variable, or the `biblioIntegrator.python` option.
3. **Install Biblium** — create an isolated virtual environment and
   install Biblium 2.16 with `install_biblium_backend()`.
4. **Convert data to Biblium format** — call `to_biblium()` to produce
   the canonical data frame that the Python bridge expects.
5. **Run Biblium group comparison** — call `biblium_compare_groups()`
   and interpret the returned `biblio_group_comparison` object.
6. **Cross-validate engines** — call `validate_biblium()` to compare
   native R and Biblium outputs on the same data, groups, and seed.
7. **Generate reports** — use `biblio_report()` to produce Markdown,
   HTML, Word, or PDF reports with corpus health, temporal trends,
   network centrality, and provenance.
8. **Customise reports** — modify the title and inspect report structure
   to adapt the template to project-specific needs.
9. **Launch the Shiny app** — call `biblio_app()` to hand an
   interactive workbench to collaborators who do not use R.
10. **Create and execute analysis plans** — define reusable plans with
    `form_plan()`, validate them with `validate_plan()`, and execute
    them with `run_plan()`.

# Setting Up the Environment {#sec-setup}

## Working Directory and Packages

All code in this vignette assumes:

- `biblioIntegrator` is installed and loaded.
- The built-in teaching dataset (`example_biblio()`) is used
  throughout.
- Optional packages (`reticulate`, `shiny`, `rmarkdown`) are guarded
  with `requireNamespace()` so that vignette compilation succeeds even
  when those packages are not installed.

```{r}
library(biblioIntegrator)
```

## The Teaching Dataset

```{r}
#| label: teaching-dataset
raw <- example_biblio()
x   <- as_biblio_project(raw, source = "v09 teaching corpus")
x
```

The project contains six relational tables:

```{r}
#| label: project-dimensions
cat("Works:      ", nrow(x$works),       "\n")
cat("Authors:    ", nrow(x$authors),     "\n")
cat("Authorships:", nrow(x$authorships), "\n")
cat("Keywords:   ", nrow(x$keywords),    "\n")
cat("References: ", nrow(x$references),  "\n")
cat("Provenance: ", nrow(x$provenance),  "\n")
```

We will return to this `biblio_project` object in every section that
follows.

# Part I — Python Integration {#sec-python}

## Why a Second Engine?

Permutation-based group comparison is the analytical core of
`biblioIntegrator`'s comparative inference module.  Running the same
test in two independent implementations — one native R, one powered by
Biblium in Python — provides a computational **cross-check**.  Agreement
between engines increases confidence; disagreement signals a bug, a
version mismatch, or a subtle difference in how overlaps are handled.

Biblium exposes a public `BiblioGroup` permutation interface.  When
Biblium 2.16 is configured, `biblioIntegrator` can call that interface
from R via `reticulate`, convert the result into the identical
`biblio_group_comparison` structure, and compare it element by element
with the native result.

The benefits are:

- **Reproducibility** — Running the same seed in two languages and
  obtaining the same chi-square, p-value, and Cramér's V strengthens
  any claim.
- **Portability** — If a co-author works in Python, the Biblium output
  can be compared directly with the R output.
- **Extensibility** — Biblium may contribute additional metrics in the
  future; the bridge lets `biblioIntegrator` access them without
  rewriting R code.

## Checking Python Availability

### `python_backend_status()`

The simplest way to inspect the current Python configuration:

```{r}
#| label: python-status
st <- python_backend_status()
st
```

The return value is a list with four fields:

| Field       | Meaning                                           |
|:------------|:--------------------------------------------------|
| `available` | Logical: is Biblium importable right now?         |
| `python`    | Path to the Python executable that was tested     |
| `version`   | Biblium version string, or `NA` if not found      |
| `reason`    | Human-readable status (`"ok"` or an explanation)  |

```{r}
#| label: python-status-names
names(st)
cat("Available:", st$available, "\n")
cat("Python:   ", st$python,    "\n")
cat("Version:  ", st$version,   "\n")
cat("Reason:   ", st$reason,    "\n")
```

### `biblium_backend_status()`

`python_backend_status()` is a backward-compatible alias for
`biblium_backend_status()`.  Both functions return the same list:

```{r}
#| label: alias-check
identical(python_backend_status(), biblium_backend_status())
```

Both functions accept an optional `python` argument to override the
default search:

```{r}
#| label: status-with-python-arg
#| eval: FALSE
biblium_backend_status(python = "/usr/bin/python3")
biblium_backend_status(python = Sys.which("python"))
```

### The Lookup Order

When `python` is `NULL` the package resolves the Python executable
through a three-tier cascade:

1. **Environment variable** — `BIBLIOINTEGRATOR_PYTHON`.
2. **R option** — `biblioIntegrator.python` (set by
   `enable_python_backend()`).
3. **System PATH** — whatever `reticulate` discovers by default.

```{r}
#| label: lookup-order
cat("Env var: ",
    Sys.getenv("BIBLIOINTEGRATOR_PYTHON", unset = "<not set>"), "\n")
cat("Option : ",
    getOption("biblioIntegrator.python",  default = "<not set>"), "\n")
```

If none of these lead to a Python with Biblium installed, `available`
will be `FALSE`.

### Interpreting the Status

```{r}
#| label: interpret-status
if (isTRUE(st$available)) {
  cat("Biblium is ready.  Version:", st$version, "\n")
} else {
  cat("Biblium is NOT available.\n")
  cat("Reason:", st$reason, "\n")
  cat("Next: install_biblium_backend() or",
      "enable_python_backend(path)\n")
}
```

## Installing Biblium

### `install_biblium_backend()`

This function creates an isolated Python virtual environment and
installs Biblium and its dependencies (`huggingface_hub`, `plotly`).

```{r}
#| label: install-biblium
#| eval: FALSE
# Default: virtual env called "r-bibliointegrator",
#          Biblium version 2.16.0
install_biblium_backend()

# Specify the Biblium version explicitly
install_biblium_backend(version = "2.16.0")

# Use a custom environment name
install_biblium_backend(envname = "my-biblium-env")
```

The function returns the virtual environment Python path invisibly and
prints the `options()` command needed to register it:

```{r}
#| label: install-output
#| eval: FALSE
# After installation, the output looks like:
# > Set options(biblioIntegrator.python = "/home/user/.virtualenvs/
#     r-bibliointegrator/bin/python") before initializing Python.
```

Copy-paste that `options()` call into your script or `.Rprofile`.

### Choosing the `envname`

| Scenario                       | Recommended `envname`       |
|:-------------------------------|:----------------------------|
| Single-user, default location  | `"r-bibliointegrator"`      |
| Project-specific isolation     | `"biblio-project-x"`        |
| CI / GitHub Actions            | `"r-bibliointegrator"`      |

### What Gets Installed

`install_biblium_backend()` installs three Python packages into the
virtual environment:

1. `biblium==2.16.0` — the core comparison library.
2. `huggingface_hub` — dependency of Biblium's model interface.
3. `plotly` — for optional interactive figures.

### Verifying the Installation

```{r}
#| label: verify-install
#| eval: FALSE
# After install, re-check:
st <- python_backend_status()
st$available   # should be TRUE
st$version     # should be "2.16.0"
```

## Configuring the Bridge

### `enable_python_backend()`

If Biblium is already installed in an existing environment (system
Python, `conda`, `virtualenv`, or `pipx`), point `biblioIntegrator` to
it:

```{r}
#| label: enable-backend
#| eval: FALSE
enable_python_backend("/usr/local/bin/python3")
enable_python_backend(Sys.which("python"))
enable_python_backend(
  reticulate::virtualenv_python("r-bibliointegrator")
)
```

The function sets `options(biblioIntegrator.python = ...)` and returns
the current backend status.  All subsequent calls to
`python_backend_status()`, `biblium_compare_groups()`, and
`validate_biblium()` will use that Python.

### Environment Variable

In non-interactive contexts (Makefiles, CI), the environment variable
takes precedence:

```{r}
#| label: env-var
#| eval: FALSE
Sys.setenv(BIBLIOINTEGRATOR_PYTHON = "/usr/local/bin/python3")
```

### R Option Directly

```{r}
#| label: option-direct
#| eval: FALSE
options(biblioIntegrator.python = "/usr/local/bin/python3")
```

### Precedence Summary

```
[1] BIBLIOINTEGRATOR_PYTHON        (environment variable)
[2] biblioIntegrator.python        (R option)
[3] reticulate auto-discovery      (system PATH)
```

Higher precedence wins.  If none resolves to a Python with Biblium,
`available` is `FALSE`.

## Converting Data to Biblium Format

### `to_biblium()`

The Python bridge expects a simple data frame with canonical column
names.  `to_biblium()` converts a `biblio_project` into that format:

```{r}
#| label: to-biblium
bm <- to_biblium(x)
class(bm)
names(bm)
head(bm, 3)
```

The returned columns are:

| Column            | Source                        |
|:------------------|:------------------------------|
| `Title`           | `x$works$title`               |
| `Year`            | `x$works$year`                |
| `Authors`         | Collapsed from `x$authorships`|
| `Author Keywords` | Collapsed from `x$keywords`   |

All values are character strings with semicolon separators.

### When to Use `to_biblium()`

- Before calling `biblium_compare_groups()` (the bridge calls it
  internally).
- When exporting data for a co-author who uses the Python `biblium`
  package directly.
- When comparing the R-side and Python-side raw input to rule out
  data-mismatch bugs.

### Column Details

```{r}
#| label: biblium-columns
cat("Class of Title:         ", class(bm$Title),         "\n")
cat("Class of Year:          ", class(bm$Year),          "\n")
cat("Class of Authors:       ", class(bm$Authors),       "\n")
cat("Class of Author Keywords:",
    class(bm$`Author Keywords`), "\n")
cat("Rows:                   ", nrow(bm),                "\n")
```

### Verifying Round-Trip Fidelity

A well-designed bridge should be lossless.  Check that every work
survives the round trip:

```{r}
#| label: round-trip
stopifnot(nrow(bm) == nrow(x$works))
stopifnot(all(bm$Year == x$works$year))
cat("Round-trip check passed: ",
    nrow(bm), "==", nrow(x$works), "\n")
```

## The Python Bridge Module

### Architecture

`biblioIntegrator` ships a Python helper at
`inst/python/biblium_bridge.py`.  When `biblium_compare_groups()` is
called, the package:

1. Loads the bridge via `reticulate::import_from_path()`.
2. Passes the `to_biblium()` data frame and the group membership
   matrix to the bridge's `compare_groups()` function.
3. Receives a Python dictionary with `observed`, `expected`,
   `residuals`, `chi_square`, `p_value`, `cramers_v`, and
   `permutation_residuals_p_adj`.
4. Wraps the result in the R `biblio_group_comparison` class.

```{r}
#| label: bridge-path
bridge_path <- system.file("python", package = "biblioIntegrator")
cat("Bridge location:", bridge_path, "\n")
dir(bridge_path)
```

### Error Handling

If `reticulate` is not installed, `biblium_backend_status()` returns:

```{r}
#| label: reticulate-explain
cat("If reticulate is absent:\n")
cat("  available: FALSE\n")
cat("  reason: 'reticulate not installed'\n")
```

If `reticulate` is installed but Biblium is not found:

```{r}
#| label: not-found-explain
cat("If Biblium is not in the Python environment:\n")
cat("  available: FALSE\n")
cat("  version:   NA\n")
cat("  reason:    'Biblium could not be imported'\n")
```

# Part II — Biblium Comparative Analysis {#sec-biblium-comparison}

## The Native Engine Recap

Before comparing engines, recall how the native R engine works.  The
`compare_groups()` function from vignette `v04` computes:

- A group-entity contingency table (`observed`).
- Expected frequencies under independence.
- Standardised residuals.
- A permutation p-value for the chi-square statistic.
- Cramér's V (with optional bootstrap confidence interval).

```{r}
#| label: native-recap
g <- ifelse(x$works$year < 2022, "early", "recent")
native_cmp <- compare_groups(
  x, g,
  entity      = "keyword",
  permutations = 199,
  seed         = 42
)
native_cmp
```

## `biblium_compare_groups()`

### Calling Biblium Explicitly

```{r}
#| label: biblium-compare
#| eval: FALSE
bm_cmp <- biblium_compare_groups(
  x, g,
  entity       = "keyword",
  permutations = 199,
  seed         = 42
)
bm_cmp
```

If Biblium is not available, this call stops with an informative error:

```{r}
#| label: biblium-availability-guard
#| eval: FALSE
tryCatch(
  biblium_compare_groups(x, g, permutations = 199, seed = 42),
  error = function(e) message("Expected: ", conditionMessage(e))
)
```

### Return Structure

The returned `biblio_group_comparison` has the same fields as the native
output:

```{r}
#| label: biblium-fields
#| eval: FALSE
names(bm_cmp)
# [1] "engine"        "entity"        "groups"
# [4] "observed"      "expected"      "residuals"
# [7] "chi_square"    "p_value"       "cramers_v"
# [10] "cramers_v_ci"  "overlap"       "permutations"
# [13] "cell_p_adjusted"
```

### The `engine` Field

The `engine` field distinguishes the two implementations:

```{r}
#| label: engine-field
#| eval: FALSE
native_cmp$engine   # "native"
bm_cmp$engine       # "biblium"
```

Both objects pass through `print.biblio_group_comparison()`,
`association_residuals()`, and `group_ca()`.

### Overlapping Groups with Biblium

```{r}
#| label: biblium-overlap
#| eval: FALSE
overlapping <- cbind(
  early  = x$works$year <= 2021,
  recent = x$works$year >= 2021
)
bm_ov <- biblium_compare_groups(
  x, overlapping,
  permutations = 199,
  seed         = 42
)
bm_ov$overlap   # TRUE
```

When groups overlap, Biblium permutes complete membership rows, just
as the native engine does.

### Author-Level Comparison

```{r}
#| label: biblium-author
#| eval: FALSE
bm_author <- biblium_compare_groups(
  x, g,
  entity       = "author",
  permutations = 199,
  seed         = 42
)
head(association_residuals(bm_author), 10)
```

## `validate_biblium()`

### Purpose

`validate_biblium()` runs the same comparison in both engines and
returns a data frame of key metrics side by side, plus both fitted
objects.  This is the recommended way to cross-check.

```{r}
#| label: validate-biblium
#| eval: FALSE
vb <- validate_biblium(
  x, g,
  entity       = "keyword",
  permutations = 199,
  seed         = 42
)
vb$comparison
```

### Interpreting the Comparison Table

```
    metric        native    biblium   difference
1 chi_square   12.340000  12.340000   0.0000000
2 p_value       0.041000   0.041000   0.0000000
3 cramers_v     0.312000   0.312000   0.0000000
```

- **difference = 0** — perfect agreement after rounding.
- **difference ≈ 1e-10** — floating-point noise; harmless.
- **difference > 0.001** — investigate; may indicate a version
  mismatch or a bug.

### Fitted Objects

```{r}
#| label: validate-objects
#| eval: FALSE
vb$native    # native biblio_group_comparison
vb$biblium   # Biblium biblio_group_comparison
```

### Author-Level Validation

```{r}
#| label: validate-author
#| eval: FALSE
vb_author <- validate_biblium(
  x, g,
  entity       = "author",
  permutations = 199,
  seed         = 42
)
vb_author$comparison
```

### Overlapping Groups Validation

```{r}
#| label: validate-overlap
#| eval: FALSE
vb_ov <- validate_biblium(
  x, cbind(
    early  = x$works$year <= 2021,
    recent = x$works$year >= 2021
  ),
  permutations = 199,
  seed         = 42
)
vb_ov$comparison
```

## Automating Engine Comparison

A practical workflow runs both engines whenever a group comparison is
performed and logs a warning when they disagree:

```{r}
#| label: auto-compare
#| eval: FALSE
safe_validate <- function(x, g, tol = 0.001, ...) {
  if (!isTRUE(biblium_backend_status()$available)) {
    message("Biblium unavailable; running native only.")
    return(compare_groups(x, g, ...))
  }
  vb <- validate_biblium(x, g, ...)
  diffs <- abs(vb$comparison$difference)
  if (any(diffs > tol, na.rm = TRUE)) {
    warning("Engine disagreement detected:\n",
            paste(capture.output(print(vb$comparison)),
                  collapse = "\n"))
  }
  vb$native
}
```

## Decision Guide

| Question                             | Recommendation              |
|:-------------------------------------|:----------------------------|
| Just need one engine?                | `compare_groups()` (native) |
| Need a second-opinion check?         | `validate_biblium()`        |
| Co-author uses Python?               | `biblium_compare_groups()`  |
| Must log agreement for audit trail?  | `validate_biblium()`        |

# Part III — Report Generation {#sec-reports}

## `biblio_report()` Overview

`biblio_report()` produces a self-contained document summarising a
`biblio_project`.  The report includes:

- **Corpus summary** — document count, total citations, year range.
- **Data quality** — results of `biblio_health()`.
- **Annual production** — output of `temporal_growth()`.
- **Leading terms** — output of `term_frequency()`.
- **Coauthorship centrality** — output of `network_centrality()`.
- **Optional backends** — output of `backend_status()`.
- **Provenance** — output of `audit_biblio()`.
- **Interpretation notes** — standard caveats.

## Markdown Reports

### Basic Usage

```{r}
#| label: report-md
f_md <- tempfile(fileext = ".md")
biblio_report(x, f_md)
file.exists(f_md)
cat("Path:", normalizePath(f_md, winslash = "/"), "\n")
```

### Inspecting the Output

```{r}
#| label: report-read
lines <- readLines(f_md, warn = FALSE)
cat("Total lines:", length(lines), "\n")
cat("First 30 lines:\n")
writeLines(lines[seq_len(min(30, length(lines)))])
```

### Structure of the Markdown File

```
# <title>

Generated: <timestamp>

## Corpus summary
Documents: **<n>**
Total citations: **<total>**
Years: **<first>-<last>**

## Data quality
| ... |
| ... |

## Annual production
| year | documents | citations | cumulative |
| ...  | ...       | ...       | ...        |

## Leading terms
| keyword | n |
| ...     | ... |

## Coauthorship centrality
| name | degree | betweenness | ... |
| ...  | ...    | ...         | ... |

## Optional backends
| ... |

## Provenance
| ... |

## Interpretation notes
...
```

### Custom Title

```{r}
#| label: report-custom-title
f_title <- tempfile(fileext = ".md")
biblio_report(
  x, f_title,
  title = "Agronomic Bibliometric Map — Nutrition Studies"
)
readLines(f_title, n = 3, warn = FALSE)
```

## HTML Reports

HTML reports require `rmarkdown`.  The function creates a temporary
`.Rmd` wrapper and renders it:

```{r}
#| label: report-html
#| eval: FALSE
f_html <- tempfile(fileext = ".html")
biblio_report(x, f_html, format = "html")
file.exists(f_html)
```

If `rmarkdown` is not installed, `biblio_report()` will stop with an
error when `format` is not `"markdown"`:

```{r}
#| label: report-rmarkdown-guard
#| eval: FALSE
# The internal check is:
if (!requireNamespace("rmarkdown", quietly = TRUE))
  stop("Install 'rmarkdown' for rendered reports.", call. = FALSE)
```

## Word Reports (DOCX)

```{r}
#| label: report-docx
#| eval: FALSE
f_docx <- tempfile(fileext = ".docx")
biblio_report(x, f_docx, format = "docx")
file.exists(f_docx)
```

Word reports use `rmarkdown::render()` with the `word_document` output
format.  The resulting `.docx` file can be opened in Microsoft Word or
LibreOffice Writer.

## PDF Reports

```{r}
#| label: report-pdf
#| eval: FALSE
f_pdf <- tempfile(fileext = ".pdf")
biblio_report(x, f_pdf, format = "pdf")
file.exists(f_pdf)
```

PDF reports require a LaTeX distribution.  On systems where LaTeX is
not available, use the HTML or DOCX format.

## Report Format Summary

| Format     | Extension | Requires          | Use Case                    |
|:-----------|:----------|:------------------|:----------------------------|
| `markdown` | `.md`     | (none)            | Quick check, version control|
| `html`     | `.html`   | `rmarkdown`       | Sharing, presenting         |
| `docx`     | `.docx`   | `rmarkdown`       | Word-based workflows        |
| `pdf`      | `.pdf`    | `rmarkdown` + LaTeX| Formal submission, archival|

## Report from Any Data Frame

`biblio_report()` accepts a plain data frame if it matches the schema
of `example_biblio()`:

```{r}
#| label: report-from-df
f_df <- tempfile(fileext = ".md")
biblio_report(example_biblio(), f_df,
              title = "Report from raw data frame")
file.exists(f_df)
```

## Report Provenance

The Provenance section of every report is generated by
`audit_biblio(x)`, which lists every transformation that has been
logged:

```{r}
#| label: report-provenance
audit_biblio(x)
```

This is what makes reports **auditable**: a reviewer can trace every
number back to a named transformation step.

## Report in a Scripted Workflow

```{r}
#| label: report-workflow
# Typical scripted pipeline:
x2 <- as_biblio_project(example_biblio(), source = "grant review")
f_report <- file.path(tempdir(), "grant-review-report.md")
biblio_report(x2, f_report, title = "Grant literature mapping")
cat("Report saved to:", f_report, "\n")
```

## Multi-Format Export

```{r}
#| label: multi-format
#| eval: FALSE
# Generate all four formats from the same project:
formats <- c("markdown", "html", "docx", "pdf")
for (fmt in formats) {
  f <- tempfile(fileext = paste0(".", substr(fmt, 1, 3)))
  if (fmt == "markdown") f <- tempfile(fileext = ".md")
  tryCatch({
    biblio_report(x, f, format = fmt, title = paste("Report -", fmt))
    cat(fmt, ": ", f, "\n")
  }, error = function(e)
    message(fmt, " failed: ", conditionMessage(e)))
}
```

# Part IV — Shiny Application {#sec-shiny}

## `biblio_app()` Overview

`biblio_app()` launches a full Shiny application with five tabs:

1. **Data** — load a CSV/TSV/JSON file or the teaching demo.
2. **Explore** — annual production plot and top keywords table.
3. **Networks** — choose network type and minimum weight, inspect
   centrality.
4. **Compare** — backend status and a note directing to
   `compare_groups()` for formal inference.
5. **Report** — download an HTML report; inspect provenance.

```{r}
#| label: app-launch
#| eval: FALSE
# Launch with teaching data pre-loaded:
biblio_app(x)

# Launch empty (user loads data via file input):
biblio_app()

# Launch with any biblio_project:
biblio_app(as_biblio_project(example_biblio()))
```

## Loading the App

### Prerequisites

```{r}
#| label: app-prerequisites
#| eval: FALSE
if (requireNamespace("shiny", quietly = TRUE)) {
  cat("Shiny is available.\n")
} else {
  cat("Install 'shiny' to use biblio_app().\n")
}
```

### Basic Launch

```{r}
#| label: app-basic
#| eval: FALSE
# In an interactive R session:
shiny::runApp(biblio_app())
```

### Passing Data Directly

```{r}
#| label: app-with-data
#| eval: FALSE
shiny::runApp(biblio_app(x))
```

## Tab 1: Data

The Data tab provides:

- A **file input** to upload a CSV, TSV, or JSON file.
- An **"Load agronomy demo"** button that loads `example_biblio()`.
- A **text output** showing project dimensions (the `print.biblio_project`
  method).
- A **table output** showing `biblio_health()` results.

```{r}
#| label: app-data-tab
# Conceptual server logic for the Data tab:
# rv <- shiny::reactiveVal(NULL)
# shiny::observeEvent(input$demo, {
#   rv(as_biblio_project(example_biblio()))
# })
# shiny::observeEvent(input$file, {
#   shiny::req(input$file)
#   rv(biblio_import(input$file$datapath))
# })
# output$summary <- shiny::renderPrint({
#   shiny::req(rv())
#   print(rv())
#   print(describe_biblio(rv()))
# })
# output$health <- shiny::renderTable({
#   shiny::req(rv())
#   biblio_health(rv())
# })
```

### Acceptable Input Formats

The `biblio_import()` function called by the Shiny server supports:

- CSV with bibliographic columns.
- TSV with bibliographic columns.
- JSON arrays of bibliographic records.

### The `biblio_import()` Path Inside the App

```{r}
#| label: import-path
#| eval: FALSE
# The server calls:
# rv(biblio_import(input$file$datapath))
# which returns a biblio_project ready for downstream analysis.
```

## Tab 2: Explore

The Explore tab provides:

- An **annual production plot** built with base R `plot()`.
- A **top terms table** showing the 15 most frequent keywords.

```{r}
#| label: app-explore-tab
# Conceptual server logic:
# output$annual <- shiny::renderPlot({
#   shiny::req(rv())
#   a <- temporal_growth(rv())
#   plot(a$year, a$documents,
#        type = "b",
#        xlab = "Year",
#        ylab = "Documents")
# })
# output$terms <- shiny::renderTable({
#   shiny::req(rv())
#   head(term_frequency(rv()), 15)
# })
```

### Annual Production Plot

```{r}
#| label: production-table
a <- temporal_growth(x)
head(a)
```

### Leading Terms

```{r}
#| label: leading-terms
head(term_frequency(x), 15)
```

## Tab 3: Networks

The Networks tab provides:

- A **select input** to choose between `"coauthor"` and `"keyword"`
  networks.
- A **numeric input** for minimum edge weight.
- A **centrality table** built from `network_centrality()`.

```{r}
#| label: app-networks-tab
# Conceptual server logic:
# output$centrality <- shiny::renderTable({
#   shiny::req(rv())
#   g <- bibliographic_network(rv(),
#                              input$ntype,
#                              min_weight = input$minw)
#   head(network_centrality(g), 20)
# })
```

### Coauthorship Centrality

```{r}
#| label: coauthor-centrality
g_ca <- bibliographic_network(x, "coauthor")
cent <- network_centrality(g_ca)
head(cent, 15)
```

### Keyword Network Centrality

```{r}
#| label: keyword-centrality
g_kw <- bibliographic_network(x, "keyword")
head(network_centrality(g_kw), 15)
```

### Adjusting Minimum Weight

Higher minimum weights prune weak edges:

```{r}
#| label: weight-filter
g_filtered <- bibliographic_network(x, "coauthor",
                                     min_weight = 2)
cat("Vertices (min=1):", igraph::vcount(g_ca),       "\n")
cat("Edges    (min=1):", igraph::ecount(g_ca),       "\n")
cat("Vertices (min=2):", igraph::vcount(g_filtered),  "\n")
cat("Edges    (min=2):", igraph::ecount(g_filtered),  "\n")
```

## Tab 4: Compare

The Compare tab displays:

- A `print(backend_status())` call showing the availability of
  optional backends.
- A text note directing users to `compare_groups()` for formal
  permutation inference.

The Shiny app deliberately does **not** run permutation inference
automatically: that is a CPU-intensive operation best controlled by a
script.

```{r}
#| label: app-compare-tab
# Conceptual server logic:
# output$backends <- shiny::renderPrint({
#   backend_status()
# })
```

### Backend Status

```{r}
#| label: backend-status
backend_status()
```

### Why Comparison Is Not Automated in the App

Permutation inference can require hundreds to thousands of random
shuffles.  In a reactive Shiny context this may block the interface.
For formal comparison, use `compare_groups()` or
`validate_biblium()` in a script and import the result.

## Tab 5: Report

The Report tab provides:

- A **download button** that generates an HTML report via
  `biblio_report()`.
- A **provenance table** built from `audit_biblio()`.

```{r}
#| label: app-report-tab
# Conceptual server logic:
# output$report <- shiny::downloadHandler(
#   filename = function() "biblio-report.html",
#   content  = function(file) {
#     shiny::req(rv())
#     biblio_report(rv(), file, "html")
#   }
# )
# output$provenance <- shiny::renderTable({
#   shiny::req(rv())
#   audit_biblio(rv())
# })
```

### What the Downloaded Report Contains

The HTML report is the same document produced by
`biblio_report(x, format = "html")` — the full structure described in
Section [Part III](#sec-reports).

### Customising the Shiny App

`biblio_app()` returns a `shinyApp` object.  You can modify the UI
or server before launching:

```{r}
#| label: app-customise
#| eval: FALSE
app <- biblio_app(x)
# app$ui     — the UI object
# app$server — the server function
# Modify as needed, then:
# shiny::runApp(app)
```

### Embedding the App in a Package

If you are building a companion package and want a one-button launch:

```{r}
#| label: app-embed
#| eval: FALSE
launch_bibliometrics <- function(data = NULL) {
  if (!requireNamespace("shiny", quietly = TRUE))
    stop("Install 'shiny' first.", call. = FALSE)
  shiny::runApp(biblio_app(data))
}
```

## Shiny App Architecture Summary

```
UI (navbarPage)
├── Data tab
│   ├── fileInput
│   ├── actionButton("demo")
│   ├── verbatimTextOutput("summary")
│   └── tableOutput("health")
├── Explore tab
│   ├── plotOutput("annual")
│   └── tableOutput("terms")
├── Networks tab
│   ├── selectInput("ntype")
│   ├── numericInput("minw")
│   └── tableOutput("centrality")
├── Compare tab
│   ├── p() note
│   └── verbatimTextOutput("backends")
└── Report tab
    ├── downloadButton("report")
    └── tableOutput("provenance")

Server
├── reactiveVal(rv): the biblio_project
├── observeEvent: demo → example_biblio()
├── observeEvent: file → biblio_import()
├── renderPrint, renderPlot, renderTable, renderTable
├── renderPrint (backends)
└── downloadHandler → biblio_report(format = "html")
```

# Part V — Analysis Plans {#sec-plans}

## Why Analysis Plans?

When a bibliometric study is repeated over multiple corpora — for
example, one dataset per grant application, one per country, or one
per time window — it is useful to define the analytical steps once and
execute them identically on every dataset.  Analysis plans serve this
purpose.

An analysis plan is a **declarative specification** of:

- The analysis blocks to run.
- The network type.
- The group definition.
- The report format.
- The reproducibility seed.

Plans are created by `form_plan()`, validated by `validate_plan()`,
and executed by `run_plan()`.

## `form_plan()`

### Basic Usage

```{r}
#| label: form-plan-basic
plan <- form_plan()
plan
class(plan)
```

The default plan requests all five analysis blocks:

```{r}
#| label: plan-analyses
plan$analyses
```

### Customising the Plan

```{r}
#| label: form-plan-custom
plan_net <- form_plan(
  analyses = c("health", "descriptive", "network"),
  network  = "keyword",
  report   = "html",
  seed     = 42
)
plan_net
```

### All Arguments

| Argument    | Default                        | Description                    |
|:------------|:-------------------------------|:-------------------------------|
| `source`    | `NULL`                         | File path or label             |
| `analyses`  | `c("health","descriptive",     | Analysis blocks                |
|             |  "temporal","network","text")  |                                |
| `network`   | `"coauthor"`                   | Network type                   |
| `group`     | `NULL`                         | Group definition               |
| `report`    | `"markdown"`                   | Report format                  |
| `seed`      | `123`                          | Reproducibility seed           |

### Plan with Group Comparison

```{r}
#| label: plan-groups
g <- ifelse(x$works$year < 2022, "early", "recent")
plan_grp <- form_plan(
  analyses = c("health", "descriptive", "groups"),
  group    = g,
  seed     = 42
)
plan_grp
```

### Plan with Network Type

```{r}
#| label: plan-network-type
plan_kw <- form_plan(network = "keyword")
plan_kw$network
```

## `validate_plan()`

### Basic Validation

```{r}
#| label: validate-plan
validate_plan(plan)
```

Validation checks:

1. The object must be a `biblio_plan`.
2. Every element of `analyses` must be one of the allowed values.
3. The `network` must be `"coauthor"`, `"keyword"`, or `"citation"`.

### Allowed Analysis Blocks

```{r}
#| label: allowed-analyses
allowed <- c("health", "descriptive", "temporal",
             "network", "text", "groups")
cat("Allowed blocks:", paste(allowed, collapse = ", "), "\n")
```

### Invalid Plans

```{r}
#| label: validate-invalid
#| eval: FALSE
tryCatch(
  validate_plan(form_plan(analyses = c("health", "unknown_block"))),
  error = function(e) message("Expected: ", conditionMessage(e))
)
```

### Plan as a Validation Gate

In a function that accepts user-specified analysis blocks, validation
catches typos early:

```{r}
#| label: plan-gate
safe_create_plan <- function(analyses, ...) {
  p <- form_plan(analyses = analyses, ...)
  validate_plan(p)
  p
}

p <- safe_create_plan(c("health", "descriptive"))
class(p)
```

## `run_plan()`

### Basic Usage

```{r}
#| label: run-plan-basic
res <- run_plan(plan, x)
names(res)
```

The result is a named list.  Each element corresponds to one analysis
block:

```{r}
#| label: run-plan-structure
sapply(res, class)
```

### What Each Block Produces

| Block         | Output Class / Type           |
|:--------------|:------------------------------|
| `project`     | `biblio_project`               |
| `health`      | `data.frame`                   |
| `descriptive` | list (`describe_biblio`)       |
| `temporal`    | `data.frame` (`temporal_growth`|
| `network`     | `igraph`                       |
| `centrality`  | `data.frame`                   |
| `terms`       | `data.frame`                   |
| `groups`      | `biblio_group_comparison`      |

### Health Block

```{r}
#| label: run-health
res$health
```

### Descriptive Block

```{r}
#| label: run-descriptive
res$descriptive
```

### Temporal Block

```{r}
#| label: run-temporal
head(res$temporal)
```

### Network Block

```{r}
#| label: run-network
cat("Vertices:", igraph::vcount(res$network), "\n")
cat("Edges:   ", igraph::ecount(res$network), "\n")
head(res$centrality, 10)
```

### Terms Block

```{r}
#| label: run-terms
head(res$terms, 15)
```

### Group Comparison Block

```{r}
#| label: run-groups
plan_g <- form_plan(
  analyses = c("groups"),
  group    = ifelse(x$works$year < 2022, "early", "recent"),
  seed     = 42
)
res_g <- run_plan(plan_g, x)
res_g$groups
```

### Running from a Data Frame

`run_plan()` accepts a raw data frame; it calls `as_biblio_project()`
internally:

```{r}
#| label: run-plan-df
res_raw <- run_plan(form_plan(analyses = "health"),
                    example_biblio())
head(res_raw$health)
```

### Reproducibility

Plans are self-contained: all random seeds are captured, so the same
plan run on the same data produces identical results.

```{r}
#| label: reproducibility
#| eval: FALSE
res1 <- run_plan(plan, x)
res2 <- run_plan(plan, x)
identical(names(res1), names(res2))
# TRUE — structure is identical
```

## Plan Composition

### Combining Plans

You can create separate plans for each phase of a study and run them
sequentially:

```{r}
#| label: plan-composition
p_qc   <- form_plan(analyses = "health")
p_desc <- form_plan(analyses = c("descriptive", "temporal"))
p_net  <- form_plan(analyses = c("network", "text"), network = "keyword")

res_qc   <- run_plan(p_qc,   x)
res_desc <- run_plan(p_desc, x)
res_net  <- run_plan(p_net,  x)

cat("QC blocks:      ", names(res_qc),   "\n")
cat("Desc blocks:    ", names(res_desc), "\n")
cat("Network blocks: ", names(res_net),  "\n")
```

### Batch Execution Over Multiple Corpora

```{r}
#| label: batch-plan
#| eval: FALSE
# Suppose you have multiple files:
corpus_files <- c("agronomy.csv",
                   "nutrition.csv",
                   "soil_science.csv")

batch_results <- lapply(corpus_files, function(f) {
  raw_f <- read.csv(f)
  run_plan(form_plan(analyses = c("health", "descriptive")))
})
```

## Plans and Reports Together

Plans and reports are designed to work together seamlessly:

```{r}
#| label: plan-report-pipeline
#| eval: FALSE
plan <- form_plan(
  analyses = c("health", "descriptive", "network", "text"),
  network  = "coauthor",
  report   = "html",
  seed     = 42
)
# Note: plan$report is used by downstream reporting logic,
#       but run_plan() itself returns analysis objects.
res <- run_plan(plan, x)

# Generate the report separately:
f_out <- file.path(tempdir(), "plan-report.md")
biblio_report(x, f_out,
              format = plan$report,
              title  = "Plan-driven report")
cat("Report:", f_out, "\n")
```

# Common Mistakes {#sec-mistakes}

## Mistake 1: Not Checking Python Availability First

**Wrong pattern:**

```{r}
#| label: mistake-1-wrong
#| eval: FALSE
biblium_compare_groups(x, g,
                       permutations = 199,
                       seed = 42)
# Error: Biblium 2.16 backend is unavailable or not importable.
```

**Correct pattern:**

```{r}
#| label: mistake-1-right
#| eval: FALSE
st <- python_backend_status()
if (isTRUE(st$available)) {
  biblium_compare_groups(x, g,
                         permutations = 199,
                         seed = 42)
} else {
  message("Biblium not available; using native engine.")
  compare_groups(x, g,
                 permutations = 199,
                 seed = 42,
                 engine = "native")
}
```

## Mistake 2: Ignoring Biblium Version Requirements

The package requires **Biblium 2.16**.  Installing an older or newer
version may yield different results or fail.

```{r}
#| label: mistake-2
#| eval: FALSE
st <- biblium_backend_status()
if (!is.na(st$version) && st$version != "2.16.0") {
  warning("Expected Biblium 2.16.0; found ", st$version)
}
```

## Mistake 3: Not Validating Cross-Engine Results

Running `biblium_compare_groups()` alone does **not** verify that the
two engines agree.  Use `validate_biblium()`:

```{r}
#| label: mistake-3
#| eval: FALSE
# Wrong: only one engine
# bm <- biblium_compare_groups(x, g, permutations = 199)

# Right: cross-validate
vb <- validate_biblium(x, g,
                       entity       = "keyword",
                       permutations = 199,
                       seed         = 1)
vb$comparison
```

## Mistake 4: Over-Relying on Shiny for Reproducible Analysis

The Shiny app is for exploration and collaboration, **not** for
reproducible pipelines.  A reproducible workflow requires scripts.

**Wrong pattern:**

```
1. Open biblio_app()
2. Click "Load agronomy demo"
3. Browse tabs
4. Download report
5. Paste report section into manuscript
```

**Correct pattern:**

```{r}
#| label: mistake-4-right
# Write a script that is re-runnable:
library(biblioIntegrator)
x <- as_biblio_project(example_biblio(), source = "thesis review")
g <- ifelse(x$works$year < 2022, "early", "recent")

# Analysis
cmp <- compare_groups(x, g, permutations = 999, seed = 42)
ca  <- group_ca(cmp)

# Report
f <- file.path(tempdir(), "thesis-review.md")
biblio_report(x, f, title = "Thesis literature review")
```

## Mistake 5: Forgetting That `validate_biblium()` Requires Both Engines

If only one engine is available, `validate_biblium()` will error.
Guard the call:

```{r}
#| label: mistake-5
#| eval: FALSE
if (isTRUE(biblium_backend_status()$available)) {
  vb <- validate_biblium(x, g,
                         permutations = 199,
                         seed = 1)
} else {
  message("Biblium unavailable; cannot cross-validate.")
  native <- compare_groups(x, g,
                           permutations = 199,
                           seed = 1)
}
```

## Mistake 6: Skipping Plan Validation

`run_plan()` calls `validate_plan()` internally, but if you build
plans programmatically it is good practice to validate early:

```{r}
#| label: mistake-6
#| eval: FALSE
# Wrong:
# plan <- form_plan(analyses = c("health", "typo"))

# Right:
plan <- form_plan(analyses = c("health", "descriptive"))
validate_plan(plan)
run_plan(plan, x)
```

## Mistake 7: Using `biblio_report()` Without Version Control

Reports overwrite files silently.  In a scripted pipeline, always
uniquely name output files:

```{r}
#| label: mistake-7
#| eval: FALSE
# Wrong:
# biblio_report(x, "report.md")
# biblio_report(x2, "report.md")  # overwrites!

# Right:
date_tag <- format(Sys.Date(), "%Y-%m-%d")
f <- paste0("report-", date_tag, ".md")
biblio_report(x, f, title = paste("Report", date_tag))
```

## Mistake 8: Misinterpreting the Shiny Compare Tab

The Compare tab shows **backend availability only** — it does not
run permutation inference.  Do not report those numbers as analytical
results.

```{r}
#| label: mistake-8
#| eval: FALSE
# What the tab shows:
backend_status()

# What it does NOT show:
# compare_groups(x, g, permutations = 999)
# Use scripts for formal inference.
```

# Working Example: End-to-End Pipeline {#sec-example}

This section assembles all major functions into a single reproducible
script.

```{r}
#| label: end-to-end
# =========================================================
#  End-to-End biblioIntegrator Pipeline
# =========================================================

# 0. Setup
library(biblioIntegrator)

# 1. Import and project creation
raw_data <- example_biblio()
proj     <- as_biblio_project(raw_data,
                               source = "end-to-end example")

# 2. Quality diagnostics
cat("=== Corpus Health ===\n")
print(biblio_health(proj))

# 3. Descriptive analysis
cat("\n=== Descriptive ===\n")
print(describe_biblio(proj))

# 4. Temporal growth
cat("\n=== Temporal Growth ===\n")
gr <- temporal_growth(proj)
print(gr)

# 5. Leading terms
cat("\n=== Leading Terms ===\n")
print(head(term_frequency(proj), 10))

# 6. Group comparison (native)
cat("\n=== Group Comparison (native) ===\n")
g <- ifelse(proj$works$year < 2022, "early", "recent")
cmp <- compare_groups(proj, g,
                      permutations = 199,
                      seed         = 42)
print(cmp)
cat("\nTop residuals:\n")
print(head(association_residuals(cmp, min_abs = 0.5), 10))

# 7. Network analysis
cat("\n=== Coauthorship Network ===\n")
net <- bibliographic_network(proj, "coauthor")
cat("Vertices:", igraph::vcount(net),
    " Edges:", igraph::ecount(net), "\n")
print(head(network_centrality(net), 8))

# 8. Backend status
cat("\n=== Backend Status ===\n")
print(backend_status())

# 9. Python backend check
cat("\n=== Python/Biblium Status ===\n")
print(python_backend_status())

# 10. Report generation
f_report <- file.path(tempdir(), "end-to-end-report.md")
biblio_report(proj, f_report,
              title = "End-to-End Bibliometric Analysis")
cat("\nReport saved to:", f_report, "\n")

# 11. Plan-based execution
plan <- form_plan(
  analyses = c("health", "descriptive", "network", "text"),
  network  = "coauthor",
  report   = "markdown",
  seed     = 42
)
plan_res <- run_plan(plan, proj)
cat("\nPlan results:", paste(names(plan_res), collapse = ", "),
    "\n")
```

# Function Reference Summary {#sec-reference}

## Python Backend Functions

| Function                     | Description                           |
|:-----------------------------|:--------------------------------------|
| `python_backend_status()`    | Check Python/Biblium availability     |
| `biblium_backend_status()`   | Canonical name (same as above)        |
| `enable_python_backend()`    | Set Python executable path            |
| `install_biblium_backend()`  | Create virtual env, install Biblium   |
| `to_biblium()`               | Convert project to Biblium data frame |
| `biblium_compare_groups()`   | Biblium group comparison              |
| `validate_biblium()`         | Cross-validate native vs. Biblium     |

## Report Function

| Function              | Description                               |
|:----------------------|:------------------------------------------|
| `biblio_report()`     | Generate MD/HTML/DOCX/PDF report          |

## Shiny Function

| Function              | Description                               |
|:----------------------|:------------------------------------------|
| `biblio_app()`        | Launch interactive Shiny workbench        |

## Plan Functions

| Function              | Description                               |
|:----------------------|:------------------------------------------|
| `form_plan()`         | Create an analysis plan                   |
| `validate_plan()`     | Validate plan structure                   |
| `run_plan()`          | Execute a plan on data                    |

# References {#sec-references}

## Primary Package References

- Pereira, W.E., Pereira Martinez, M.H. (2026).
  *biblioIntegrator: Harmonized, Comparative and Network-Based
  Bibliometric Analysis*. R package version 0.3.0.

- Umek, L. (2026). Biblium: a Python library for
  comparative bibliometric analysis. *Scientometrics*, 131(5),
  3359–3377.
  doi:[10.1007/s11192-026-05636-8](https://doi.org/10.1007/s11192-026-05636-8)

## Key Methodological References

- Aria, M. & Cuccurullo, C. (2017). bibliometrix: An R-tool for
  comprehensive science mapping analysis. *Journal of Informetrics*,
  11(4), 959–975.
  doi:[10.1016/j.joi.2017.08.007](https://doi.org/10.1016/j.joi.2017.08.007)

## Web Resources

- Biblium on PyPI —
  <https://pypi.org/project/biblium/>
- biblioIntegrator GitHub —
  <https://github.com/wep69/biblioIntegrator>
- reticulate documentation —
  <https://rstudio.github.io/reticulate/>
- Shiny tutorial —
  <https://shiny.posit.co/r/articles/>
- rmarkdown: The Definitive Guide —
  <https://pkg.yihui.org/rmarkdown-book/>

## Session Info

```{r}
#| label: session-info
sessionInfo()
```

---

*This vignette is part of the biblioIntegrator documentation
(vignettes `v00` to `v09`). For the foundational tutorial, see
`v07-foundations-to-advanced-tutorial.Rmd`. For scalable backends
(Arrow, DuckDB), see `v08-scalable-backends.Rmd`.*
