---
title: "Biblium Engines: Diversity, Representativeness, Main Path and Means"
author: "biblioIntegrator team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biblium Engines: Diversity, Representativeness, Main Path and Means}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biblioIntegrator)
has_biblium <- isTRUE(tryCatch(biblium_backend_status()$available,
                               error = function(e) FALSE))
```

The Python backend is used as a second statistical engine behind four R
functions. All chunks below run only when Biblium is importable; otherwise
they print the skip reason. See `backend_status()` for availability.

```{r backend}
biblium_backend_status()
```

## Diversity

Shannon, Simpson and Gini indices over authors, keywords or sources, for the
whole corpus, per year, or per group.

```{r diversity, eval=has_biblium}
x <- as_biblio_project(example_biblio())
d <- biblium_diversity(x, entity = "author")
d$indices
d$summary
biblium_diversity(x, entity = "keyword", level = "temporal",
                  min_items_per_year = 1L)$indices
g <- ifelse(x$works$year < 2022, "early", "late")
biblium_diversity(x, level = "group", groups = g)$indices
```

## Representativeness against OpenAlex

Percentage-point differences between the corpus year distribution and the
global OpenAlex benchmark. The reference is cached on disk; only a cache
miss reaches the network.

```{r representativeness, eval=has_biblium}
r <- biblium_representativeness(x)
r$table
```

An offline reference can be injected for testing:

```{r representativeness-mock, eval=has_biblium}
ref <- data.frame(Year = sort(unique(x$works$year)), Count = 100)
biblium_representativeness(x, reference = ref)$table
```

## Main path

SPC/SPLC/SPNP traversal weights over the citation network (edges flow
cited -> citing). A corpus without references stops with guidance.

```{r mainpath, eval=has_biblium}
ids <- x$works$work_id[1:5]
x$references <- data.frame(citing_id = ids[-1], cited_id = ids[-length(ids)])
p <- biblium_main_path(x)
p$global_main_path
p$edge_weights
```

## Means comparison

Automatic test selection (t/Welch/Mann-Whitney; ANOVA/Welch-ANOVA/
Kruskal-Wallis with post-hoc), assumption checks and effect sizes.

```{r means, eval=has_biblium}
m <- biblium_compare_means(x, groups = g)
m$recommended_test
m$tests
```

## Unified plan

The four engines are plan blocks with a strictness switch:

```{r plan, eval=has_biblium}
plan <- form_plan(analyses = c("health", "diversity", "means"),
                  group = g, strict = FALSE)
run <- run_plan(plan, x)
names(run)
```

## Wave 2: crosstabs, correlations, patterns, concepts, disruption

```{r crosstab, eval=has_biblium}
ct <- biblium_crosstab(x, "source", g)
ct$chi_squared
ct$effect_size
mc <- biblium_correlate(x)
mc$corr_matrix
```

```{r patterns, eval=has_biblium}
cp <- biblium_citation_patterns(x, source = "estimated")
cp$summary
```

Native concepts and reference diversity (Biblium audits Rao-Stirling):

```{r concepts}
x <- as_biblio_project(example_biblio())
head(concept_ngrams(x), 5)
dim(concept_cooccurrence(x))
reference_diversity(x)
```

```{r disruption-check, eval=has_biblium}
ids <- x$works$work_id[1:5]
x$references <- data.frame(citing_id = ids[-1], cited_id = ids[-length(ids)])
vd <- validate_disruption(x, focal_ids = ids[1:4])
vd$comparison
```

## Wave 3: SDG, classifier, geo, topics

```{r sdg, eval=has_biblium}
s <- biblium_sdg(x, text = "title")
s$flags
```

```{r classifier, eval=has_biblium}
cl <- biblium_classify_groups(x, g)
cl$performance
```

```{r topics, eval=has_biblium}
tp <- biblium_topics(x, n_topics = 2L)
tp$summary
```

```{r session-info}
sessionInfo()
```
